---
title: "DeSurv Gene Selection Alignment: Impact on M4/M5 External C-Index"
author: "Andrew Walther"
date: "2026-05-27"
format:
  pdf:
    latex-engine: xelatex
    keep-tex: false
    fig-pos: "htbp"
    include-in-header:
      text: |
        \usepackage{booktabs}
        \usepackage{geometry}
        \usepackage{titling}
        \geometry{margin=0.42in}
        \renewcommand{\familydefault}{\sfdefault}
        \setlength{\droptitle}{-1.8em}
        \pretitle{\begin{center}\large\bfseries}
        \posttitle{\par\end{center}\vspace{-0.7em}}
  html:
    toc: true
    code-fold: true
    theme: cosmo
fontsize: 9pt
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(knitr)
library(ggplot2)
library(dplyr)

OUT <- "../../results/benchmark_sim/outputs/desurv_comparison"
res <- if (file.exists(file.path(OUT, "desurv_comparison_results.csv"))) {
         read.csv(file.path(OUT, "desurv_comparison_results.csv"),
                  stringsAsFactors = FALSE)
       } else {
         NULL
       }

if (!is.null(res)) {
  means <- res |>
    group_by(model, label, K, k_eff) |>
    summarise(mean_c = round(mean(c_index), 3), .groups = "drop")
  mc <- setNames(means$mean_c, means$model)
  ke <- setNames(means$k_eff,  means$model)
  Kv <- setNames(means$K,      means$model)
} else {
  mc <- setNames(rep(NA_real_, 4), c("D1","D2","D3","D4"))
  ke <- setNames(rep(NA_integer_, 4), c("D1","D2","D3","D4"))
  Kv <- setNames(rep(NA_integer_, 4), c("D1","D2","D3","D4"))
}
```

## 1. Motivation

DeSurv (Young et al. 2025) selects genes by independently ranking each gene's
mean expression and variance, then retaining the top-3,000 with the lowest
combined rank-sum per cohort — genes that are both highly expressed *and* highly
variable. Our prior analysis used variance-only selection on the full merged
matrix after normalization, which (a) conflates per-cohort with cross-cohort
variance and (b) can retain lowly-expressed noisy genes near the detection floor.
This report evaluates whether aligning with DeSurv's gene selection changes
the external concordance of the recommended manuscript configurations.

## 2. Configurations Compared

```{r config-table}
tab <- data.frame(
  ID     = c("D1","D2","D3","D4"),
  Label  = c("LB orig (M4)","YFB orig (M5)",
             "LB DeSurv-aligned","YFB DeSurv-aligned"),
  Model  = c("LB","YFB","LB","YFB"),
  top_n  = c(2000, 2000, 3000, 3000),
  Method = c("variance","variance","combined\\_rank","combined\\_rank"),
  Selection = c("post-norm","post-norm","per-cohort","per-cohort"),
  Cohort = c("Yes","No","Yes","No"),
  Mean_C = c(mc["D1"], mc["D2"], mc["D3"], mc["D4"]),
  K      = c(Kv["D1"], Kv["D2"], Kv["D3"], Kv["D4"]),
  K_eff  = c(ke["D1"], ke["D2"], ke["D3"], ke["D4"])
)
knitr::kable(tab, booktabs = TRUE, row.names = FALSE,
             col.names = c("ID","Label","Model","top\\_n","Method",
                           "Selection","Cohort","Mean C","K","$K_{\\text{eff}}$"),
             escape = FALSE,
             caption = "Four configurations compared. D1/D2 reproduce M4/M5 exactly. D3/D4 apply DeSurv gene selection (combined mean+variance rank, top-3000 per cohort before normalization).")
```

## 3. Per-Cohort C-Index

```{r cohort-fig, fig.height=3.2, fig.cap="Per-cohort external C-index for D1–D4. Dashed line = chance (C=0.5). D1/D2 are the original manuscript configs; D3/D4 are DeSurv-aligned."}
if (!is.null(res)) {
  res_plot <- res |>
    mutate(type = ifelse(model %in% c("D1","D2"), "Original", "DeSurv-aligned"),
           label_short = sub(" \\(.*", "", label))
  ggplot(res_plot, aes(x = cohort, y = c_index, fill = label)) +
    geom_col(position = position_dodge(0.75), width = 0.65, color = "white") +
    geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey40") +
    scale_fill_manual(
      values = c("LB orig (M4)"        = "#4c78a8",
                 "YFB orig (M5)"       = "#f58518",
                 "LB DeSurv-aligned"   = "#72b7b2",
                 "YFB DeSurv-aligned"  = "#e45756"),
      name = NULL) +
    coord_cartesian(ylim = c(0.45, 0.75)) +
    labs(x = NULL, y = "C-index") +
    theme_bw(base_size = 8) +
    theme(legend.position = "top",
          axis.text.x = element_text(angle = 20, hjust = 1, size = 7))
} else {
  plot.new(); text(0.5, 0.5, "Results CSV not found", cex = 1.2)
}
```

## 4. Findings and Recommendation

DeSurv-aligned preprocessing (D4, YFB + combined mean+variance rank, top-3000 per
cohort, K=7) achieves a mean external C-index of 0.636 across 5 held-out PDAC
cohorts, compared to 0.624 for the original YFB configuration (M5/D2). The
delta of +0.012 exceeds the pre-specified ±0.005 practical threshold. D4
improves over D2 in 4 of 5 cohorts (PACA\_AU\_array is the one exception,
0.650 vs 0.670). The recommended primary configuration for the manuscript is
**D4 (YFB DeSurv-aligned, K=7)**, with D3 (LB DeSurv-aligned, K=7, mean
C=0.622) as sensitivity. DeSurv-aligned gene selection also improves LB:
delta = +0.006 (D3 vs D1).

## 5. Gene Set Comparison

```{r gene-table}
tab2 <- data.frame(
  Config = c("D1/D2 (original)","D3/D4 (DeSurv-aligned)"),
  top_n  = c(2000, 3000),
  Method = c("variance, post-norm", "combined_rank, per-cohort"),
  Genes_after_selection = c(2000, 2064)
)
knitr::kable(tab2, booktabs = TRUE, row.names = FALSE,
             caption = "Gene set size after selection and intersection. D3/D4 selects top-3000 per cohort on log-transformed data before normalization, then intersects across cohorts.")
```
