---
title: "Merged-Cohort PDAC Benchmark: LB vs. YFB, Joint vs. Per-Platform Preprocessing, With and Without Cohort Indicator"
author: "Andrew Walther"
date: "`r Sys.Date()`"
format:
  pdf:
    toc: true
    number-sections: true
    fig-pos: "H"
  html:
    toc: true
    code-fold: true
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(yaml)
library(ggplot2)
library(tidyr)
library(dplyr)

cfg     <- yaml::read_yaml(here::here("config/globals.yml"))
results <- read.csv(here::here(
  "results/benchmark_sim/outputs/merged_benchmark/merged_benchmark_results.csv"))
top_genes <- tryCatch(
  read.csv(here::here(
    "results/benchmark_sim/outputs/merged_benchmark/merged_benchmark_top_genes.csv")),
  error = function(e) NULL)

model_labels <- c(
  M1 = "LB | joint | no cohort adj.",
  M2 = "LB | joint | cohort_id",
  M3 = "LB | per-platform | no cohort adj.",
  M4 = "LB | per-platform | cohort_id",
  M5 = "YFB | per-platform | no cohort adj.",
  M6 = "YFB | per-platform | cohort_id"
)
results$model_label <- model_labels[results$model]
```

## Study Design

This report compares six model configurations for joint multi-cohort PDAC analysis,
trained on merged TCGA_PAAD (RNA-seq, n=144) and CPTAC (proteomics, n=129) data
and evaluated on five independent external cohorts.

**Goal:** Identify the configuration that produces a parsimonious model (small, CV-selected K)
with non-zero survival associations (β ≠ 0) and biologically coherent gene expression programs
that generalize across platforms and patient populations.

| ID | Model | Preprocessing | Cohort adjustment | K |
|----|-------|---------------|-------------------|---|
| M1 | LB (η = Lβ) | Joint quantile+rank | None | `r cfg$benchmark$k_merged_lb_joint` |
| M2 | LB | Joint quantile+rank | Cohort indicator | `r cfg$benchmark$k_merged_lb_joint` |
| M3 | LB | Per-platform z-std | None | `r cfg$benchmark$k_merged_lb_perplatform` |
| M4 | LB | Per-platform z-std | Cohort indicator | `r cfg$benchmark$k_merged_lb_perplatform` |
| M5 | YFB (η = YFβ) | Per-platform z-std | None | `r cfg$benchmark$k_merged_yfb_perplatform` |
| M6 | YFB | Per-platform z-std | Cohort indicator | `r cfg$benchmark$k_merged_yfb_perplatform` |

YFB × joint preprocessing is excluded: β→0 collapse is structural and all mitigation
strategies (V0–V11) have been exhausted (see DECISIONS.md 2026-05-22/25).

**K selection:** Cross-validated C-index (1-SE rule) on merged training data, separately
for each preprocessing × model combination. This replaces the previous K=20+ARD approach.

## External C-index Results

```{r cindex-table}
wide <- results |>
  select(model_label, cohort, c_index) |>
  pivot_wider(names_from = cohort, values_from = c_index)

means <- results |>
  group_by(model_label) |>
  summarise(Mean = round(mean(c_index), 3), .groups = "drop")

wide <- left_join(wide, means, by = "model_label")
knitr::kable(wide, digits = 3, caption = "External C-index by model and cohort")
```

```{r cindex-heatmap, fig.height=4, fig.width=8}
ggplot(results, aes(x = cohort, y = model_label, fill = c_index)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = sprintf("%.3f", c_index)), size = 3) +
  scale_fill_gradient2(low = "#d73027", mid = "#fee090", high = "#1a9850",
                       midpoint = 0.55, name = "C-index") +
  labs(title = "External C-index heatmap", x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

## Model Summary (β activity)

```{r beta-summary}
summary_df <- results |>
  distinct(model, model_label, k_eff, beta_max, n_iters) |>
  mutate(
    K = c(cfg$benchmark$k_merged_lb_joint, cfg$benchmark$k_merged_lb_joint,
          cfg$benchmark$k_merged_lb_perplatform, cfg$benchmark$k_merged_lb_perplatform,
          cfg$benchmark$k_merged_yfb_perplatform, cfg$benchmark$k_merged_yfb_perplatform)[
      match(model, c("M1","M2","M3","M4","M5","M6"))]
  ) |>
  select(model_label, K, k_eff, beta_max, n_iters)
knitr::kable(summary_df, digits = 4,
             caption = "Model summary: CV-selected K, active factors (K_eff), max |β|, iterations to convergence")
```

## Factor Interpretability: Top Genes

```{r top-genes, results='asis'}
if (!is.null(top_genes)) {
  for (mid in c("M1","M2","M3","M4","M5","M6")) {
    sub <- top_genes[top_genes$model == mid, ]
    if (nrow(sub) == 0) next
    cat(sprintf("\n### %s --- %s\n\n", mid, model_labels[mid]))
    for (k in sort(unique(sub$factor))) {
      genes_k <- sub[sub$factor == k, "gene"]
      cat(sprintf("**Factor %d:** %s\n\n", k, paste(genes_k, collapse = ", ")))
    }
  }
} else {
  cat("Top gene table not available (run benchmark without --quick).\n")
}
```

## Conclusions

*(To be filled in after results are available. Key questions to address:)*

- *Which configuration achieves the highest mean external C-index?*
- *Does per-platform preprocessing consistently improve over joint preprocessing for LB?*
- *Does the cohort indicator help on top of per-platform preprocessing, or is it redundant?*
- *Which configuration produces the most biologically coherent top genes per factor?*
- *What is the recommended configuration for the manuscript?*
