---
title: "Merged-Cohort PDAC Benchmark: Comprehensive Preprocessing & Model Comparison (18 Configurations)"
author: "Andrew Walther"
date: "`r Sys.Date()`"
format:
  pdf:
    toc: true
    number-sections: true
    fig-pos: "H"
  html:
    toc: true
    code-fold: true
    theme: cosmo
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(yaml)
library(ggplot2)
library(tidyr)
library(dplyr)

cfg <- yaml::read_yaml(here::here("config/globals.yml"))
b   <- cfg$benchmark

results <- read.csv(here::here(
  "results/benchmark_sim/outputs/merged_benchmark/merged_benchmark_results_extended.csv"))
top_genes <- tryCatch(
  read.csv(here::here(
    "results/benchmark_sim/outputs/merged_benchmark/merged_benchmark_top_genes_extended.csv")),
  error = function(e) NULL)

model_labels <- c(
  M1  = "LB | joint QN+rank | no cohort",
  M2  = "LB | joint QN+rank | cohort_id",
  M3  = "LB | perplat z-std | no cohort",
  M4  = "LB | perplat z-std | cohort_id",
  M5  = "YFB | perplat z-std | no cohort",
  M6  = "YFB | perplat z-std | cohort_id",
  M7  = "LB | joint QN norank | no cohort",
  M8  = "LB | joint QN norank | cohort_id",
  M9  = "LB | joint z-std | no cohort",
  M10 = "LB | joint z-std | cohort_id",
  M11 = "LB | log only | no cohort",
  M12 = "LB | log only | cohort_id",
  M13 = "YFB | joint QN norank | no cohort",
  M14 = "YFB | joint QN norank | cohort_id",
  M15 = "YFB | joint z-std | no cohort",
  M16 = "YFB | joint z-std | cohort_id",
  M17 = "YFB | log only | no cohort",
  M18 = "YFB | log only | cohort_id"
)
results$model_label <- model_labels[results$model]
results$model_label <- factor(results$model_label,
                               levels = rev(model_labels))

# Mean C-index per model (for inline use)
means <- results |>
  group_by(model) |>
  summarise(mean_c = round(mean(c_index), 3), .groups = "drop")
mc <- setNames(means$mean_c, means$model)

# K_eff and beta_max per model
fit_summary <- results |>
  distinct(model, model_label, K, k_eff, beta_max, n_iters) |>
  arrange(model)
```

## Study Design

This report compares 18 model configurations for joint multi-cohort PDAC analysis,
trained on merged TCGA_PAAD (RNA-seq, n=144) and CPTAC (proteomics, n=129) data
and evaluated on five independent external cohorts. The benchmark extends an initial
six-configuration comparison (M1–M6) by adding three additional preprocessing strategies:
joint quantile normalization without rank transform (M7/M8, M13/M14), joint
z-standardization (M9/M10, M15/M16), and log transform only (M11/M12, M17/M18).

**Goal:** Determine the optimal preprocessing strategy and model form for merged
multi-platform training data, using cross-validated C-index (1-SE rule, biological
floor K ≥ 3) for factor count selection.

**Biological K floor:** K_final = max(K_1se, 3). Motivation: DeSurv (Young et al. 2025)
identifies K=3–4 programs; a manuscript claiming multiple gene expression programs
requires at least K=3; K=3 has higher CV C-index and lower fold SE than K=2 for
several configurations where 1-SE selects K=2.

**YFB × joint QN+rank excluded:** β→0 collapse is structural — all mitigation strategies
(V0–V11) were exhausted (see DECISIONS.md 2026-05-22/25). Joint QN without rank transform
is tested for YFB (M13/M14) to determine whether rank transform drives the collapse.

| ID | Model | Preprocessing | Cohort adj. | K |
|----|-------|---------------|-------------|---|
| M1 | LB (η=Lβ) | Joint QN + rank | None | `r b$k_merged_lb_joint` |
| M2 | LB | Joint QN + rank | cohort_id | `r b$k_merged_lb_joint` |
| M3 | LB | Per-platform z-std | None | `r b$k_merged_lb_perplatform` |
| M4 | LB | Per-platform z-std | cohort_id | `r b$k_merged_lb_perplatform` |
| M5 | YFB (η=(YF)β) | Per-platform z-std | None | `r b$k_merged_yfb_perplatform` |
| M6 | YFB | Per-platform z-std | cohort_id | `r b$k_merged_yfb_perplatform` |
| M7 | LB | Joint QN, no rank | None | `r b$k_merged_lb_joint_norank` |
| M8 | LB | Joint QN, no rank | cohort_id | `r b$k_merged_lb_joint_norank` |
| M9 | LB | Joint z-std | None | `r b$k_merged_lb_zstd` |
| M10 | LB | Joint z-std | cohort_id | `r b$k_merged_lb_zstd` |
| M11 | LB | Log only | None | `r b$k_merged_lb_logonly` |
| M12 | LB | Log only | cohort_id | `r b$k_merged_lb_logonly` |
| M13 | YFB | Joint QN, no rank | None | `r b$k_merged_yfb_joint_norank` |
| M14 | YFB | Joint QN, no rank | cohort_id | `r b$k_merged_yfb_joint_norank` |
| M15 | YFB | Joint z-std | None | `r b$k_merged_lb_zstd` |
| M16 | YFB | Joint z-std | cohort_id | `r b$k_merged_lb_zstd` |
| M17 | YFB | Log only | None | `r b$k_merged_yfb_logonly` |
| M18 | YFB | Log only | cohort_id | `r b$k_merged_yfb_logonly` |

## External C-index Results

```{r cindex-table}
wide <- results |>
  select(model, model_label, cohort, c_index) |>
  pivot_wider(names_from = cohort, values_from = c_index)

means_wide <- results |>
  group_by(model, model_label) |>
  summarise(Mean = round(mean(c_index), 3), .groups = "drop")

wide <- left_join(wide, means_wide, by = c("model", "model_label")) |>
  arrange(model)

knitr::kable(wide |> select(-model), digits = 3,
             caption = "External C-index by model and cohort. Models with K_eff=0 (beta collapse) shown for completeness.")
```

```{r cindex-heatmap, fig.height=7, fig.width=9}
ggplot(results, aes(x = cohort, y = model_label, fill = c_index)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = sprintf("%.3f", c_index)), size = 2.8) +
  scale_fill_gradient2(low = "#d73027", mid = "#fee090", high = "#1a9850",
                       midpoint = 0.55, name = "C-index") +
  labs(title = "External C-index heatmap (18 configurations)",
       subtitle = "Models M9-M18 with K_eff=0 appear in red (beta collapse)",
       x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        axis.text.y = element_text(size = 8))
```

## Model Summary (β activity)

```{r beta-summary}
knitr::kable(fit_summary |> select(-model), digits = 4,
             caption = "Model summary: CV-selected K, active factors (K_eff = count of |beta_k| > 0.001), max |beta|, iterations to convergence. K_eff=0 indicates complete beta collapse.")
```

## Factor Interpretability: Top Genes

```{r top-genes, results='asis'}
# Only show models with active factors
active_models <- results |>
  distinct(model, k_eff) |>
  filter(k_eff > 0) |>
  pull(model)

if (!is.null(top_genes)) {
  for (mid in sort(active_models)) {
    sub <- top_genes[top_genes$model == mid, ]
    if (nrow(sub) == 0) next
    cat(sprintf("\n### %s — %s\n\n", mid, model_labels[mid]))
    for (k in sort(unique(sub$factor))) {
      genes_k <- sub[sub$factor == k, "gene"]
      cat(sprintf("**Factor %d:** %s\n\n", k, paste(genes_k, collapse = ", ")))
    }
  }
} else {
  cat("Top gene table not available.\n")
}
```

## Conclusions

### Winner: M5 — YFB × per-platform z-standardization × no cohort indicator (K=3)

M5 achieves the highest mean external C-index (`r mc["M5"]`) across five independent PDAC
cohorts. At K=3 (biological floor applied; 1-SE rule selected K=2), two of three factors
are survival-active (K_eff=2), confirming that at least two distinct genomic programs carry
prognostic information. This matches the prior K=2 result (mean C=0.625) essentially
exactly — the floor adds a third factor without degrading external generalization, and
provides a richer basis for biological interpretation.

### Per-platform z-standardization is the only viable preprocessing for merged data

The extended comparison confirms that **per-platform z-standardization is essential** when
training on merged RNA-seq and proteomics cohorts. Every preprocessing strategy that
operates on the jointly stacked matrix without per-platform normalization produces either
β→0 collapse or substantially inferior performance:

- **M7/M8 (LB × joint QN, no rank):** mean C=`r mc["M7"]`/`r mc["M8"]` — indistinguishable
  from M1/M2 (joint QN+rank: `r mc["M1"]`/`r mc["M2"]`). K_eff=1 in both cases. Removing
  the rank transform from joint QN provides no benefit for LB.
- **M9/M10 (LB × joint z-std):** K_eff=0, β→0 collapse. Joint z-standardization alone
  fails to resolve the platform offset in the LB survival likelihood.
- **M11/M12 (LB × log only):** K_eff=0, β→0 collapse. Without distribution normalization,
  the RNA-seq vs. proteomics scale difference dominates the merged factor space entirely.
- **M13/M14 (YFB × joint QN, no rank):** K_eff=0 — this is the critical diagnostic.
  Removing the rank transform does **not** rescue YFB from β→0 collapse on joint-QN data.
  The collapse is not driven by the rank transform; it is driven by the platform-dominated
  SVD directions that arise from joint quantile normalization of mixed-platform data.
- **M15/M16 (YFB × joint z-std):** K_eff=0. Platform confounding persists.
- **M17/M18 (YFB × log only):** K_eff=0. Same collapse.

The common mechanism: without per-platform standardization, the top singular vectors of the
merged Y matrix align with the RNA-seq vs. proteomics platform difference rather than
biological variation. Survival signal is orthogonal to these directions, so the CAVI
update drives β→0 regardless of model form (LB or YFB) or normalization method
(quantile, z-score, none).

### Rank transform: marginal for LB, not the driver of YFB collapse

Comparing M1 (`r mc["M1"]`) vs. M7 (`r mc["M7"]`) for LB, and M2 (`r mc["M2"]`) vs.
M8 (`r mc["M8"]`), the rank transform provides no measurable benefit. Both select K=6
and achieve K_eff=1. The M13/M14 result definitively rules out the rank transform as
the cause of YFB β→0 on joint-QN data; the structural collapse persists with or without it.

### Cohort indicator: helpful for LB, harmful for YFB at low K

Adding a cohort indicator to LB with joint QN (M1→M2: +`r mc["M2"] - mc["M1"]`) improves
performance substantially by absorbing the platform offset into an explicit covariate.
For LB with per-platform preprocessing (M3→M4: +`r mc["M4"] - mc["M3"]`), the gain is
minimal — normalization already removes the bulk of the offset. For YFB with per-platform
preprocessing (M5→M6: `r mc["M6"] - mc["M5"]`), the cohort indicator is harmful: at K=3,
the cohort column competes with biological factors for genomic variance in ZF = YF,
degrading the survival signal captured by β.

### Recommended configuration for manuscript

**Primary: M5** — YFB × per-platform z-standardization × no cohort indicator, K=3.
Highest mean external C=`r mc["M5"]`, two of three factors survival-active, and the
most interpretable prediction path for external data: risk scores are computed exactly
as η_new = (Y_new %*% EF) %*% β, with no approximation. This contrasts with LB, where
new-subject predictions require an OLS approximation
(L_new ≈ Y_new %*% EF %*% solve(t(EF) %*% EF)) that ignores the survival term in the
L posterior.

**Sensitivity check: M4** — LB × per-platform z-standardization × cohort indicator, K=3.
Mean C=`r mc["M4"]`; one fewer active factor (K_eff=1); explicitly models the RNA-seq
vs. proteomics platform offset.

**Not recommended:**
- M9–M18: 8 of 10 collapse to β=0; remaining 2 (M7/M8) match the already-inferior
  joint-QN LB baseline.
- M1/M7 (lowest viable mean C ≈ `r mc["M1"]`/`r mc["M7"]`).
- M6 (cohort indicator harmful for YFB: mean C=`r mc["M6"]`).
