---
title: "Multi-Cohort Simulation Study: Shared vs. Study-Specific Factors"
subtitle: "Proposal, implementation, and initial results for the SSBMF model"
author: "Andrew Walther"
date: "2026-06-14"
format:
  pdf:
    latex-engine: xelatex
    keep-tex: false
    fig-pos: "htbp"
    include-in-header:
      text: |
        \usepackage{booktabs}
        \usepackage{geometry}
        \usepackage{titling}
        \geometry{margin=0.55in}
        \renewcommand{\familydefault}{\sfdefault}
        \setlength{\droptitle}{-1.8em}
        \pretitle{\begin{center}\large\bfseries}
        \posttitle{\par\end{center}\vspace{-0.3em}}
  html:
    toc: true
    code-fold: true
    theme: cosmo
    html-math-method: mathjax
fontsize: 10pt
execute:
  echo: false
  warning: false
  message: false
---

## 1. Purpose and Background

Prior synthetic validation of the Supervised Bayesian Matrix Factorization
(SSBMF) model was carried out only in a **single-cohort** setting. The real
application pools several PDAC cohorts that differ both by **platform**
(RNA-seq, microarray, proteomics) and by **biology**. In that setting the latent
factors fall into two kinds:

- **Shared factors** — gene-expression programs present across *all* studies.
  These are the cross-cohort biological signal we want to detect and, where they
  are prognostic, to use for survival prediction.
- **Study-specific factors** — programs private to a *single* study. These absorb
  platform/batch effects and cohort-private biology that should *not* be mistaken
  for shared, generalizable signal.

This proposal specifies a simulation that generates data with a **known**
shared / study-specific factor structure, so we can test whether the current
model **natively** recovers that distinction — the same goal the previously
discussed cohort dummy-indicator approach was meant to serve. If the base model
already separates shared from study-specific factors, the dummy indicator columns
may be unnecessary.

The design incorporates the plan from the last meeting (5/29), with two key elements:

1. **EBMF-grounded ground truth.** Realistic gene programs are taken from an
   Empirical Bayes Matrix Factorization (`flashier`) fit to the merged real
   training cohorts. EBMF is also our **unsupervised benchmark** (Section 5).
2. **Shared factors carry survival signal; study-specific factors do not**
   ($\beta \neq 0$ on shared, $\beta = 0$ on specific). Survival is generated
   from the shared component only.

---

## 2. Notation

| Symbol | Meaning | Shape |
|---|---|---|
| $C$ | number of cohorts (studies) | scalar |
| $n_c$ | patients in cohort $c$; $n = \sum_c n_c$ | scalar |
| $p$ | number of genes | scalar |
| $K$ | total number of latent factors | scalar |
| $\mathcal{S}$ | index set of **shared** factors | $|\mathcal{S}|$ |
| $\mathcal{P}_c$ | index set of factors **specific** to cohort $c$ | $|\mathcal{P}_c|$ |
| $L$ | patient loadings | $n \times K$ |
| $F$ | gene programs (factors) | $p \times K$ |
| $\beta$ | log-hazard coefficients | $K$ |
| $Y$ | observed genomics matrix | $n \times p$ |
| $a_{\text{shared}}, a_{\text{specific}}$ | amplitude (signal strength) of shared / specific factors | scalar |

Rows of $Y$ (and $L$) are **stacked by cohort**: rows $1..n_1$ are cohort 1, the
next $n_2$ rows are cohort 2, and so on. Factors partition disjointly:
$K = |\mathcal{S}| + \sum_{c} |\mathcal{P}_c|$. The remaining noise and survival
parameters ($\tau_j$, $\alpha$, $\lambda_0$, $U_i$) are defined inline where they
first appear in Section 3.

---

## 3. Data-Generating Process

### 3.1 Loadings $L$ ($n \times K$) — block structure encodes specificity

Every loading is drawn non-negative,
$$ L_{ik} \sim \text{Exponential}(\text{rate}=1), \qquad \mathbb{E}[L]=1,\ \mathbb{E}[L^2]=2, $$
matching the `point_exponential` prior used by the fitters. Specificity is then
imposed by **zeroing out-of-cohort loadings**:

- **Shared factor** $k \in \mathcal{S}$: keep $L_{ik}$ for *all* patients $i$.
- **Cohort-$c$-specific factor** $k \in \mathcal{P}_c$: set
  $$ L_{ik} = 0 \quad \text{for every } i \notin \text{cohort } c . $$

A study-specific factor therefore contributes to $Y$ **only in its own cohort's
rows** — this *is* the operational definition of "study-specific." The resulting
$L$ is block-sparse: shared columns are dense, specific columns are zero outside
one cohort block.

### 3.2 Gene Programs / Factors $F$ ($p \times K$)

Each column $F_{\cdot k}$ is a **gene program**: a vector of gene weights,
normalized to unit $L_2$ norm. Two sources are supported:

- **Preferred (EBMF-grounded):** real gene programs from a `flashier` fit to the
  merged training cohorts (realistic, sparse). When $K$ exceeds the EBMF rank (the
  number of programs the `flashier` fit returns), columns are recycled by index.
- **Synthetic fallback** (no real data / no `flashier`): for each factor, pick 5%
  of genes at random and give them equal weight.

Each column is then scaled by an amplitude — $a_{\text{shared}}$ for shared
factors, $a_{\text{specific}}$ for specific ones — which sets how strongly each
factor type contributes signal relative to the noise.

> **Non-negativity must match the prior.** The SSBMF models place a non-negative
> (`point_exponential`) prior on $F$, so the ground-truth programs are made
> non-negative (absolute value) and the EBMF benchmark is run non-negative too.
> Feeding signed programs to a non-negative-prior model is a prior/data mismatch
> that artificially depresses recovery — so both are kept non-negative for a fair
> test.

### 3.3 Genomics model (Gaussian)

$$ Y = L F^{\top} + E, \qquad
   E_{ij} \sim \mathcal{N}(0,\ 1/\tau_j),\quad
   \tau_j \sim \text{Gamma}(\text{shape}=2,\ \text{rate}=2). $$

Each gene $j$ has its own noise precision $\tau_j$, giving noise SD
$1/\sqrt{\tau_j}$ ($\mathbb{E}[\tau_j]=1$), matching the established single-cohort
DGP. $Y$ is column-centered before fitting (standard preprocessing). Batch effects
are absorbed by the study-specific factors by construction, so no explicit
platform offset is added (an optional per-cohort gene-mean offset is available for
stress-testing but is off by default).

### 3.4 Survival model (Weibull proportional hazards) — shared factors only

$$ \eta_i = \sum_{k \in \mathcal{S}} L_{ik}\,\beta_k
   \qquad (\beta_k = 0 \text{ for every specific factor}), $$
$$ T_i = \left( \frac{-\log U_i}{\lambda_0\cdot \exp(\eta_i)} \right)^{1/\alpha},
   \quad U_i \sim \text{Unif}(0,1), $$

where $\alpha = 1.5$ is the Weibull shape and $\lambda_0 = 0.01$ is the baseline scale
(controls the overall event rate). The baseline cumulative hazard is
$H_0(t) = \lambda_0\, t^{\alpha}$.
The default shared coefficients cycle the established pattern
$(1.5, -1.2, 0.8, -0.5)$ to length $|\mathcal{S}|$ (truncated if $|\mathcal{S}| < 4$;
e.g. $|\mathcal{S}|=2$ gives $\beta=(1.5,-1.2)$). **Censoring** is calibrated to
$\approx 30\%$ by bisection: $\text{censor\_base} \sim \text{Exp}(1)$,
$\text{time} = \min(\text{event}, \text{censor})$,
$\text{status} = \mathbf{1}\{\text{event} \le \text{censor}\}$.

> **Anchoring caveat (meeting notes).** Because the survival outcome is linked
> only to shared factors, the supervision "anchors" those factors into the shared
> component. This is exactly the intended mechanism; the results report discusses
> whether it produces cleaner shared/specific separation.

---

## 4. The Three Scenarios

The scenarios differ only in how the $K$ factors are partitioned into shared vs.
specific. For the primary setting $C=2$, $n_c=150$, $p=1000$, $K=6$:

| Scenario | $|\mathcal{S}|$ shared ($\beta\neq 0$) | specific per cohort ($\beta=0$) | Interpretation |
|---|---|---|---|
| **All shared** | 6 | 0, 0 | Studies effectively one cohort |
| **Hybrid** | 2 | 2, 2 | Both shared and private structure |
| **Nothing shared** | 0 | 3, 3 | No cross-cohort structure |

**Nothing-shared is a deliberate false-positive control.** With $\mathcal{S} =
\varnothing$, $\eta \equiv 0$ and survival carries no factor signal. The correct
result is C-index $\approx 0.5$ and all $|\hat\beta|$ below threshold — the model
should *not* fabricate a shared prognostic factor. An optional knob
(`specific_prognostic`, default off) can give each cohort a within-cohort
prognostic factor if we later want survival signal in this regime.

### Worked example — Hybrid, $C=2$, $n_c=150$, $p=1000$, $K=6$

- **Dimensions:** $n=300$; $Y$ is $300\times1000$, $L$ is $300\times6$, $F$ is
  $1000\times6$, $\beta$ is length 6. Rows 1–150 = cohort 1; 151–300 = cohort 2.
- **Factor labels:** `[shared, shared, specific_1, specific_1, specific_2, specific_2]`.
- **Loadings $L$:** columns 1–2 dense (all 300 patients). Columns 3–4 zero for
  rows 151–300 (private to cohort 1). Columns 5–6 zero for rows 1–150 (private to
  cohort 2). $L$ shows a visible block-sparse pattern.
- **Coefficients:** $\beta = [1.5,\ -1.2,\ 0,\ 0,\ 0,\ 0]$ — only the two shared
  factors drive survival.
- **What a correct model should find:** factors 1–2 with loadings across *both*
  cohorts *and* nonzero $\hat\beta$; factors 3–6 cohort-private with
  $\hat\beta \approx 0$. The unsupervised EBMF benchmark should recover the same
  six gene programs but cannot say which two are prognostic.

**Secondary sensitivity:** a 3-cohort hybrid (2 shared + 1 private per cohort), and
a sweep of the shared-to-specific amplitude ratio
($a_{\text{shared}}/a_{\text{specific}}$) within the hybrid regime to stress the
case where study-specific signal dominates.

---

## 5. Model Arms

Each simulated dataset ($Y$, `time`, `status`, `cohort_id`) is fit by **five
arms** — four supervised (LB/YFB $\times$ $\pm$cohort indicator) plus the
unsupervised EBMF benchmark:

| Arm | Model | Cohort indicator |
|---|---|---|
| **YFB_base** | YFB ($\eta = (YF)\beta$) — recommended | none |
| **YFB_cohort** | YFB | yes |
| **LB_base** | LB ($\eta = L\beta$) | none |
| **LB_cohort** | LB | yes |
| **EBMF** | unsupervised `flashier` (survival-blind) | — |

All supervised arms use a non-negative (`point_exponential`) prior on $L$ and $F$
and a normal prior on $\beta$. The cohort indicator, when present, absorbs
per-cohort gene-mean differences without entering the survival predictor.

**The native-identification test.** The headline analysis is the **base** arms
(no cohort indicator): can the fitted factors separate into shared vs.
study-specific on their own? The $\pm$indicator comparison then quantifies whether
the indicator adds anything.

**Why EBMF.** EBMF is fit to the same $Y$ but is survival-blind, so it has no
$\beta$. It is the unsupervised gold standard for the factorization itself,
answering two questions:

1. Does the supervised model recover the shared/study-specific structure *as well
   as* EBMF (i.e., supervision does not cost structural fidelity)?
2. Does supervision *additionally* yield correct $\beta$ — flagging which factors
   are prognostic — which EBMF fundamentally cannot?

This is the value-add argument: same structural recovery, plus correct survival
attribution.

> The EBMF here (fit to each *simulated* $Y$, scored as a competing method) is
> distinct from the EBMF in Section 3.2 (fit to *real* data to build ground-truth
> programs). They do not share a fit.

---

## 6. Evaluation Metrics

For each (scenario × model arm × seed) we record:

- **Shared-factor recovery** — mean best $|\text{cor}|$ between estimated and true
  gene programs ($F$ columns, gene space, comparable across LB/YFB/EBMF), over the
  *true shared* factors.
- **Specific-factor recovery** — same, over the *true study-specific* factors.
- **Specificity-classification accuracy** — label each *estimated* factor
  shared/specific from its per-cohort loading energy (a factor is called specific
  if more than 85% of its squared-loading energy concentrates in a single cohort),
  then compare to the truth. This is the direct test of "does the model natively
  identify the distinction?"
- **Survival-effect recovery** — $|\hat\beta|$ on factors matched to true shared
  (expect above a small threshold) vs. true specific (expect $\approx 0$).
  Reported as true-positive and **false-positive prognostic rates** (the latter is
  the anchoring / nothing-shared check).
- **C-index** — orientation-free held-out concordance, $\max(C, 1-C)$.

**Comparisons reported:** (i) YFB vs. LB; (ii) base vs. $+$cohort-dummy ($\Delta$
in each metric — "are dummies necessary?"); (iii) supervised vs. unsupervised
EBMF.

---

## 7. Reference Data-Generating Code

The following is the complete, self-contained DGP (the project implementation
wraps this same logic with configuration and seeds). A reader can run it directly
given the small `.calibrate` helper.

```{r dgp-reference, eval=FALSE, echo=TRUE}
# Bisection calibration of censoring scale to a target censoring fraction.
.calibrate <- function(ev, base, target = 0.30, n_iter = 40) {
  lo <- 1e-3; hi <- max(ev) * 10
  for (i in seq_len(n_iter)) {
    mid <- sqrt(lo * hi)
    if (mean(base * mid < ev) > target) lo <- mid else hi <- mid
  }
  hi
}

generate_multicohort_data <- function(
    C            = 2L,                 # number of cohorts
    n_per        = rep(150L, C),       # patients per cohort (rows stacked by cohort)
    p            = 1000L,              # genes
    K_shared     = 2L,                 # # shared factors      (beta != 0)
    K_specific   = rep(2L, C),         # # specific factors per cohort (beta = 0)
    F_templates  = NULL,               # p x K_ebmf EBMF programs; NULL -> synthetic sparse
    a_shared     = 1.0,                # amplitude of shared  F columns
    a_specific   = 1.0,                # amplitude of specific F columns
    active_rate  = 0.05,               # synthetic-fallback sparsity
    beta_shared  = NULL,               # length K_shared; NULL -> recycle c(1.5,-1.2,0.8,-0.5)
    offset_sd    = 0.0,                # explicit platform offset SD (0 = off)
    shape        = 1.5, scale0 = 0.01, # Weibull-PH baseline (shape = alpha, scale0 = lambda_0)
    target_cens  = 0.30,               # target censoring fraction
    seed         = 1L) {

  set.seed(seed)
  n  <- sum(n_per)
  K  <- K_shared + sum(K_specific)
  cohort_id <- factor(rep(seq_len(C), times = n_per))

  # label + owner per column (owner 0 = shared)
  labels <- c(rep("shared", K_shared),
              unlist(lapply(seq_len(C), function(cc) rep(paste0("specific_", cc), K_specific[cc]))))
  owner  <- c(rep(0L, K_shared),
              unlist(lapply(seq_len(C), function(cc) rep(cc, K_specific[cc]))))

  # L (n x K): Exponential(1), block-zeroed outside owner cohort
  L <- matrix(rexp(n * K, rate = 1), n, K)
  for (k in seq_len(K)) if (owner[k] != 0L) L[cohort_id != owner[k], k] <- 0

  # F (p x K): EBMF templates (unit-norm) or synthetic sparse (unit-norm)
  Fm <- matrix(0, p, K)
  if (!is.null(F_templates)) {
    idx <- ((seq_len(K) - 1L) %% ncol(F_templates)) + 1L
    Fm  <- F_templates[, idx, drop = FALSE]
    Fm  <- sweep(Fm, 2, sqrt(colSums(Fm^2)) + 1e-10, "/")
  } else {
    for (k in seq_len(K)) {
      active <- sample.int(p, max(1L, round(active_rate * p)))
      Fm[active, k] <- 1 / sqrt(length(active))
    }
  }
  amp <- ifelse(labels == "shared", a_shared, a_specific)   # signal-strength knob
  Fm  <- sweep(Fm, 2, amp, "*")

  # per-gene Gaussian noise: tau_j ~ Gamma(2,2)
  tau <- rgamma(p, shape = 2, rate = 2)
  E   <- sweep(matrix(rnorm(n * p), n, p), 2, sqrt(tau), "/")

  # optional explicit platform offset (off by default)
  offset <- matrix(0, n, p)
  if (offset_sd > 0) for (cc in seq_len(C)) {
    ov <- rnorm(p, 0, offset_sd)
    offset[cohort_id == cc, ] <- matrix(ov, sum(cohort_id == cc), p, byrow = TRUE)
  }

  # assemble + column-center
  Y <- L %*% t(Fm) + offset + E
  Y <- sweep(Y, 2, colMeans(Y), "-")

  # survival from SHARED factors only
  if (is.null(beta_shared))
    beta_shared <- if (K_shared > 0) rep_len(c(1.5, -1.2, 0.8, -0.5), K_shared) else numeric(0)
  beta <- numeric(K); beta[labels == "shared"] <- beta_shared
  eta  <- as.vector(L %*% beta)                  # == 0 in the "nothing shared" scenario

  event <- (-log(runif(n)) / (scale0 * exp(eta)))^(1 / shape)
  cb    <- rexp(n, rate = 1)
  cs    <- .calibrate(event, cb, target = target_cens)
  ctime <- cb * cs
  time   <- pmin(event, ctime)
  status <- as.integer(event <= ctime)

  list(Y = Y, time = time, status = status, cohort_id = cohort_id,
       L_true = L, F_true = Fm, beta_true = beta,
       factor_labels = labels, factor_owner = owner)
}
```

---

## 8. Initial Results

The design above has been implemented and run end-to-end (3 scenarios $\times$ 5
arms $\times$ 5 seeds). This section is a high-level overview; the companion
results report (`multicohort_sim_results_06_14_26`) carries the full tables,
figures, and methodology notes. **These are preliminary and may be revised after
this proposal is discussed.**

| Scenario | Arm | Rec. shared | Rec. specific | Spec. acc. | C-index | $\beta$ FP |
|---|---|:--:|:--:|:--:|:--:|:--:|
| all_shared | YFB_base | 0.92 | — | 1.00 | 0.87 | — |
| all_shared | YFB_cohort | 0.94 | — | 0.90 | 0.86 | — |
| all_shared | LB_base | 0.53 | — | 1.00 | 0.74 | — |
| all_shared | EBMF | 0.82 | — | 0.90 | — | — |
| hybrid | YFB_base | 0.90 | 0.84 | 0.97 | 0.81 | 0.50 |
| hybrid | YFB_cohort | 0.92 | 0.92 | 1.00 | 0.81 | 0.20 |
| hybrid | LB_base | 0.72 | 0.67 | 0.83 | 0.76 | 0.20 |
| hybrid | EBMF | 0.68 | 0.78 | 0.80 | — | — |
| nothing_shared | YFB_base | — | 0.89 | 1.00 | 0.55 | 0.03 |
| nothing_shared | YFB_cohort | — | 0.91 | 1.00 | 0.54 | 0.03 |
| nothing_shared | LB_base | — | 0.78 | 0.87 | 0.54 | 0.07 |
| nothing_shared | EBMF | — | 0.70 | 0.80 | — | — |

*Mean over 5 seeds; LB_cohort omitted for brevity. **Column definitions:**
**Rec. shared / Rec. specific** — gene-program recovery, the mean best
$|\text{correlation}|$ between estimated and true gene programs over the true
shared / study-specific factors (1 = perfect). **Spec. acc.** —
specificity-classification accuracy, the fraction of estimated factors correctly
labeled shared vs. study-specific from their per-cohort loading energy.
**C-index** — held-out (external) concordance on the 25% test split (0.5 = chance).
**$\beta$ FP** — false-positive prognostic rate: fraction of true study-specific
factors ($\beta = 0$) wrongly given a survival effect (lower is better). EBMF is
survival-blind (no C-index, no $\beta$).*

```{r proposal-fig-setup}
#| include: false
suppressPackageStartupMessages({ library(ggplot2); library(dplyr); library(tidyr) })
OUT_MC  <- "../../results/multi_cohort_sim/outputs"
ex_path <- file.path(OUT_MC, "multicohort_sim_example.rds")
have_ex <- file.exists(ex_path)
if (have_ex) ex <- readRDS(ex_path)
```

```{r fig-loadings-proposal, fig.width=9, fig.height=5.5, eval=have_ex, fig.cap="Patient loading heatmaps for the hybrid scenario (seed 1, training split). Rows = patients grouped by cohort (left sidebar: blue = cohort 1, salmon = cohort 2); columns = factors F1-F6. Top annotation: true factor type (Active = shared/prognostic; Shrunk = study-specific/non-prognostic) and estimated |beta-hat| from YFB_base. Cell colour: white (zero) to red (high loading). LEFT: true L showing the ground-truth block-sparse structure. RIGHT: estimated L from YFB_base — the model recovers the same block structure with no indicator columns."}
suppressPackageStartupMessages({ library(pheatmap); library(gridExtra) })
d    <- ex$hybrid$data
spl  <- ex$hybrid$split
cid  <- droplevels(d$cohort_id[spl$train_idx])
fit  <- ex$hybrid$YFB_base
col_idx <- fit$match$match          # col_idx[j] = estimated col for true factor j
labs    <- d$factor_labels; flab <- paste0("F", seq_along(labs))
ord     <- order(cid); cid_ord <- cid[ord]
gap_row <- which(diff(as.integer(cid_ord)) != 0)
Ltrue <- d$L_true[spl$train_idx, , drop = FALSE][ord, , drop = FALSE]
Lhat  <- fit$EL[ord, col_idx, drop = FALSE]   # select estimated cols in true-factor order
rownames(Ltrue) <- rownames(Lhat) <- paste0("P", seq_len(nrow(Ltrue)))
colnames(Ltrue) <- colnames(Lhat) <- flab
row_anno  <- data.frame(Cohort = cid_ord, row.names = rownames(Ltrue))
cohort_pal <- setNames(c("#2166AC", "#D6604D"), levels(cid_ord))
col_anno  <- data.frame(
  Type     = factor(ifelse(labs == "shared", "Active", "Shrunk"), levels = c("Active","Shrunk")),
  Abs_Beta = abs(fit$EBeta[col_idx]), row.names = flab)   # |β̂| in true factor order
anno_colors <- list(Cohort = cohort_pal,
                    Type   = c(Active = "#D62728", Shrunk = "#BDBDBD"),
                    Abs_Beta = c("#FFFFFF", "#08306B"))
heat_pal <- colorRampPalette(c("#FFFFFF", "#FCBBA1", "#FB6A4A", "#A50F15"))(100)
ph_args  <- list(cluster_rows=FALSE, cluster_cols=FALSE, show_rownames=FALSE,
                 show_colnames=TRUE, color=heat_pal, annotation_row=row_anno,
                 annotation_col=col_anno, annotation_colors=anno_colors,
                 gaps_row=gap_row, border_color=NA, fontsize_col=9, fontsize=9,
                 legend=TRUE, annotation_legend=TRUE, silent=TRUE)
p_true <- do.call(pheatmap, c(list(Ltrue, main="True L"), ph_args))
p_hat  <- do.call(pheatmap, c(list(Lhat,  main="Estimated L (YFB_base, no indicator)"), ph_args))
gridExtra::grid.arrange(p_true$gtable, p_hat$gtable, ncol = 2)
```

**The headline findings:**

1. **The model natively identifies the shared / study-specific distinction.** In
   the hybrid scenario, YFB classifies factors as shared vs. specific with
   accuracy 0.97 (base) to 1.00 (+cohort indicator), purely from per-cohort loading
   energy — no indicator columns required. This directly answers the central
   question.
2. **It does not fabricate prognostic signal when none exists.** In nothing_shared
   ($\eta \equiv 0$ by construction), every arm sits at C-index $\approx 0.54$ and
   the $\beta$ false-positive rate is 0.03–0.07 — the model does not invent a
   shared prognostic factor out of purely study-specific structure.
3. **Survival prediction is strong where shared prognostic signal exists**
   (C-index 0.81–0.87, YFB highest), and EBMF — being survival-blind — produces no
   risk score at all. This is the value-add of supervision.
4. **Supervision does not cost structural fidelity.** YFB recovers shared programs
   at $|\text{cor}| \approx 0.90$–0.94 and specific programs at 0.84–0.92, matching
   or exceeding the non-negative EBMF benchmark (0.68–0.82).
5. **The cohort indicator helps under genuine heterogeneity, and is redundant
   otherwise.** In hybrid, adding it halves the $\beta$ false-positive rate
   (0.50 $\to$ 0.20) and lifts specificity accuracy to 1.00; in all_shared it adds
   nothing. This is the quantitative answer to "are the dummy indicators
   necessary?"
6. **YFB is more reliable than LB**, which collapses to near-zero recovery on some
   seeds — consistent with YFB being the recommended model.

---

## 9. Open Questions for Discussion

1. **Specificity definition.** We encode study-specific factors via block-zero
   loadings (a factor is active in exactly one cohort). Is this the structure you
   intend, or should specific factors also have *non-overlapping gene programs*
   per cohort?
2. **Nothing-shared survival.** Default leaves survival null in the nothing-shared
   scenario (false-positive control). Do you prefer instead a within-cohort
   prognostic specific factor so each cohort has its own survival signal?
3. **Cohort count.** Primary is $C=2$ (matches the real training data and the
   fixed-effects rationale); a $C=3$ hybrid is included as sensitivity. Is 2
   sufficient for the methods claim, or should 3 be primary?
4. **Signal ratio.** Default $a_{\text{shared}} = a_{\text{specific}}$. Should the
   hybrid scenario stress a regime where study-specific variance *dominates*
   shared variance (the harder, more realistic case)?

5. **Cohort indicator: design decision and remaining false positives.** In the
   hybrid scenario, adding the cohort dummy indicator halves the $\beta$
   false-positive rate (0.50 $\to$ 0.20), but does not eliminate it — 20% of
   study-specific factors still receive a spurious prognostic coefficient even
   with the indicator present.

   The mechanism is confounding: study-specific factors are active in exactly
   one cohort, and that cohort's patients have different survival trajectories
   (driven by the shared prognostic factors with different mean loadings across
   cohorts). Without a cohort label, the model absorbs this cohort-level
   survival difference by assigning $\hat\beta$ to the specific factors. The
   indicator breaks this confounding, but residual within-cohort correlation
   (finite sample, $n_c = 150$) accounts for the remaining 0.20.

   Two open sub-questions:

   (a) *Should the cohort indicator be part of the default recommended
   configuration?* The current D4 real-data benchmark achieves C = 0.636
   **without** the indicator, suggesting confounding is tolerable in practice.
   But this simulation targets the equal-signal hybrid ($a_\text{shared} =
   a_\text{specific}$); whether the PDAC cohorts sit in a harder regime (Q4
   above) where the FP problem worsens is unknown.

   (b) *What additional approaches could reduce the residual false positives?*
   Three candidates:
   - **Signal-ratio sweep** (planned): stress the specific-dominant regime
     ($a_\text{specific} \gg a_\text{shared}$) to quantify how badly the FP
     rate grows and whether the indicator remains sufficient.
   - **Tighter $\beta$ prior**: a tighter normal prior on $\beta$ would shrink
     small spurious coefficients toward zero without the full collapse that the
     `point_normal` prior causes. This has not been tested in this setting.
   - **Cohort-stratified Cox baseline**: replace the shared baseline hazard with
     cohort-stratified strata, natively absorbing cohort-level survival
     differences without indicator columns. This is a more substantial model
     change.

   **The signal-ratio sweep (Q4/Q5b) has now been completed** (see the results
   report, §7). Key findings: (i) the FP rate is bounded at 0.20–0.50 across
   all variance ratios — it does not grow monotonically as specific factors
   dominate; (ii) the indicator's advantage is regime-dependent and narrows at
   high variance ratios; (iii) structural recovery and C-index are completely
   stable across all ratios. The recommendation: include the cohort indicator
   as a default (greatest benefit at equal-to-moderate heterogeneity), and
   apply a $|\hat\beta|$ magnitude threshold rather than a binary nonzero
   criterion to distinguish true shared prognostic factors from spurious
   specific-factor assignments.

*The working implementation is in `results/multi_cohort_sim/` (runner:
`run_multicohort_sim.R`; outputs: `outputs/multicohort_sim_results.csv`), and the
full results report with figures is at
`docs/reports/multicohort_sim_results_06_14_26.{pdf,html}`. The design,
implementation, and results can all be updated based on feedback to this proposal.*
