---
title: "Multi-Cohort Simulation Study: Shared vs. Study-Specific Factors"
subtitle: "SSBMF multi-cohort simulation — proposal and key results"
author: "Andrew Walther"
date: "2026-06-14"
format:
  pdf:
    latex-engine: xelatex
    keep-tex: false
    fig-pos: "htbp"
    include-in-header:
      text: |
        \usepackage{booktabs}
        \usepackage{geometry}
        \usepackage{titling}
        \geometry{top=0.55in, left=0.55in, right=0.55in, bottom=0.8in}
        \renewcommand{\familydefault}{\sfdefault}
        \setlength{\droptitle}{-1.8em}
        \pretitle{\begin{center}\large\bfseries}
        \posttitle{\par\end{center}\vspace{-0.3em}}
  html:
    toc: true
    code-fold: true
    theme: cosmo
    html-math-method: mathjax
fontsize: 10pt
execute:
  echo: false
  warning: false
  message: false
---

## 1. Purpose and Background

Prior synthetic validation of the Supervised Bayesian Matrix Factorization
(SSBMF) model was carried out only in a **single-cohort** setting. The real
application pools several PDAC cohorts that differ both by **platform**
(RNA-seq, microarray, proteomics) and by **biology**. In that setting the latent
factors fall into two kinds:

- **Shared factors** — gene-expression programs present across *all* studies.
  These are the cross-cohort biological signal we want to detect and, where they
  are prognostic, to use for survival prediction.
- **Study-specific factors** — programs private to a *single* study. These absorb
  platform/batch effects and cohort-private biology that should *not* be mistaken
  for shared, generalizable signal.

This proposal specifies a simulation that generates data with a **known**
shared / study-specific factor structure, so we can test whether the current
model **natively** recovers that distinction — the same goal the previously
discussed cohort dummy-indicator approach was meant to serve. If the base model
already separates shared from study-specific factors, the dummy indicator columns
may be unnecessary.

The design incorporates the plan from the last meeting (5/29), with two key elements:

1. **EBMF-grounded ground truth.** Realistic gene programs are taken from an
   Empirical Bayes Matrix Factorization (`flashier`) fit to the merged real
   training cohorts. EBMF is also our **unsupervised benchmark** (Section 5).
2. **Shared factors carry survival signal; study-specific factors do not**
   ($\beta \neq 0$ on shared, $\beta = 0$ on specific). Survival is generated
   from the shared component only.

---

## 2. Notation

| Symbol | Meaning | Shape |
|---|---|---|
| $C$ | number of cohorts (studies) | scalar |
| $n_c$ | patients in cohort $c$; $n = \sum_c n_c$ | scalar |
| $p$ | number of genes | scalar |
| $K$ | total number of latent factors | scalar |
| $\mathcal{S}$ | index set of **shared** factors | $|\mathcal{S}|$ |
| $\mathcal{P}_c$ | index set of factors **specific** to cohort $c$ | $|\mathcal{P}_c|$ |
| $L$ | patient loadings | $n \times K$ |
| $F$ | gene programs (factors) | $p \times K$ |
| $\beta$ | log-hazard coefficients | $K$ |
| $Y$ | observed genomics matrix | $n \times p$ |
| $a_{\text{shared}}, a_{\text{specific}}$ | amplitude (signal strength) of shared / specific factors | scalar |

Rows of $Y$ (and $L$) are **stacked by cohort**: rows $1..n_1$ are cohort 1, the
next $n_2$ rows are cohort 2, and so on. Factors partition disjointly:
$K = |\mathcal{S}| + \sum_{c} |\mathcal{P}_c|$. The remaining noise and survival
parameters ($\tau_j$, $\alpha$, $\lambda_0$, $U_i$) are defined inline where they
first appear in Section 3.

---

## 3. Data-Generating Process

We *specify* the latent quantities $L$, $F$, and $\beta$ — building in exactly
which factors are shared and which are cohort-specific, and which carry survival
signal — then *generate* the observed data $(Y,\ \text{time},\ \text{status})$
from them. The model is given only the observed data and tasked with recovering
the latent structure; because we know the ground truth, every recovery error is
directly measurable.

**Specified (ground truth):** $F$ (gene programs), $L$ (patient loadings with the
shared/specific block structure), $\beta$ (which factors are prognostic).
**Generated (model input):** $Y$ (expression matrix), and the survival outcomes
$\text{time}$ and $\text{status}$. **Model's task:** recover $\hat F$, $\hat L$,
and $\hat\beta$ from $(Y, \text{time}, \text{status})$ alone.

The components are described below in the order they are constructed.

### 3.1 Gene Programs $F$ ($p \times K$) — what each factor "looks like" in gene space

$F$ is specified first because it defines what each latent factor *means*
biologically: column $k$ is a $p$-vector of gene weights describing the $k$-th
gene expression program.

**Construction.** We use real PDAC gene programs as ground truth rather than
arbitrary synthetic ones.
Concretely, we fit an Empirical Bayes Matrix Factorization (EBMF; `flashier`,
`var_type = 2`, greedy + backfit) to the merged real PDAC training cohorts and
extract the resulting gene-loading matrix — $p \times K_\text{EBMF}$ columns,
each a sparse, data-grounded gene program. These columns are then used as the
$K$ columns of $F$ (recycled by index if $K > K_\text{EBMF}$).

Two adjustments are made before use:

1. **Unit normalization.** Each column is divided by its $L_2$ norm so that
   $\|F_{\cdot k}\| = 1$. This decouples program *shape* from program
   *amplitude*, letting a separate scalar $a_k$ control how strongly factor $k$
   appears in $Y$.
2. **Non-negativity.** The SSBMF model places a non-negative
   (`point_exponential`) prior on $F$ — the same NMF-style assumption used at
   initialization. The EBMF programs are signed; taking their absolute value
   gives non-negative programs consistent with the model's representational
   assumption. The EBMF benchmark arm is run non-negative for the same reason.

Each column is then multiplied by an amplitude scalar: $a_\text{shared}$ for
columns assigned to shared factors and $a_\text{specific}$ for cohort-specific
ones. Because $\|F_{\cdot k}\| = 1$, the variance contributed by factor $k$ to
$Y$ scales as $a_k^2$, making the amplitude the direct signal-strength parameter.

A synthetic fallback is used when real data are unavailable: each column is a
random sparse vector (5% of genes active, equal weights), unit-normalized and
sign-adjusted the same way.

### 3.2 Patient Loadings $L$ ($n \times K$) — who expresses each factor, and how much

$L$ encodes how strongly each patient expresses each gene program. The entry
$L_{ik}$ is the *loading* of patient $i$ on factor $k$: large values mean
patient $i$'s expression profile is strongly shaped by program $k$.

**Distribution.** Every entry is drawn independently from an Exponential
distribution with rate 1:
$$ L_{ik} \sim \text{Exponential}(\text{rate}=1), \qquad
   \mathbb{E}[L_{ik}] = 1, \quad \mathbb{E}[L_{ik}^2] = 2. $$
The Exponential is non-negative and right-skewed: most patients have a moderate
loading near 1, a minority have high loadings, and none are negative. This
matches the `point_exponential` prior the model places on $L$ during inference —
the ground truth and the prior are on the same scale and in the same family.

**Shared vs. specific structure.** After drawing all entries, specificity is
imposed by zeroing out patients who do not belong to the owning cohort:

- **Shared factor** $k \in \mathcal{S}$: $L_{ik}$ is kept for *every* patient,
  regardless of cohort. All patients can express a shared program.
- **Cohort-$c$-specific factor** $k \in \mathcal{P}_c$: $L_{ik}$ is set to
  **zero** for every patient $i$ not in cohort $c$. Only patients from cohort
  $c$ can express this program; it is invisible in all other cohorts.

The resulting $L$ is block-sparse: shared columns are dense (all patients),
specific columns are zero everywhere except one cohort block. This is the
operational definition of "study-specific" built into the simulation.

### 3.3 Survival coefficients $\beta$ ($K$-vector) — which factors are prognostic

$\beta$ is **fixed by design** — it is not drawn from a distribution. It encodes
which factors the model should identify as prognostic:

- **Shared factors** $k \in \mathcal{S}$: assigned nonzero coefficients, cycling
  the pattern $(1.5,\,-1.2,\,0.8,\,-0.5)$ to length $|\mathcal{S}|$.
  For the primary hybrid scenario ($|\mathcal{S}|=2$): $\beta = (1.5,\,-1.2)$,
  so the first shared factor is positively prognostic (high loading → worse
  outcome), the second is protective.
- **Cohort-specific factors** $k \in \mathcal{P}_c$: $\beta_k = 0$. These
  factors drive expression patterns within a cohort but carry no survival
  information. The correct model behaviour is to assign $\hat\beta_k \approx 0$
  to these factors.

The model receives no information about which $\beta_k$ are zero — it must infer
this from the data. Recovering nonzero $\beta$ on shared factors (true positives)
while keeping $\hat\beta \approx 0$ on specific factors (avoiding false
positives) is the central evaluation criterion.

### 3.4 Observed genomics matrix $Y$ ($n \times p$)

Given $L$, $F$, we generate what the model will actually observe:

$$ Y = L F^{\top} + E, \qquad
   E_{ij} \sim \mathcal{N}(0,\,1/\tau_j), \quad
   \tau_j \sim \text{Gamma}(\text{shape}=2,\,\text{rate}=2). $$

Each gene $j$ has its own noise precision $\tau_j$ drawn from a Gamma prior
($\mathbb{E}[\tau_j] = 1$, so typical noise SD $\approx 1$). This gives
heteroscedastic noise across genes, matching real expression data.
$Y$ is column-centered before being passed to the model (standard preprocessing).

At this point the simulation has produced the expression matrix $Y$ — equivalent
to what we would observe from an RNA-seq or array experiment — with the latent
factor structure $(L, F)$ baked in.

### 3.5 Observed survival outcomes — time and status

The survival outcomes are generated from the **shared factors only**, via a
Weibull proportional-hazards model. Specific factors are deliberately excluded
from the survival model ($\beta_k = 0$), so any prognostic signal the model finds
on specific factors is a false positive.

**Linear predictor** (shared factors only):
$$ \eta_i = \sum_{k \in \mathcal{S}} L_{ik}\,\beta_k. $$

**Event time** via Weibull-PH inverse CDF:
$$ T_i = \left( \frac{-\log U_i}{\lambda_0 \cdot \exp(\eta_i)} \right)^{1/\alpha},
   \quad U_i \sim \text{Uniform}(0,1), $$
where $\alpha = 1.5$ (Weibull shape) and $\lambda_0 = 0.01$ (baseline scale).
This is the standard inversion method: $U_i$ is a uniform random draw, and the
formula inverts the Weibull-PH survival function
$S(t\mid\eta_i) = \exp(-\lambda_0 t^\alpha e^{\eta_i})$ to produce a random
event time that respects the proportional-hazards structure.

**Censoring** is applied by drawing a censoring time $C_i = s \cdot B_i$ where
$B_i \sim \text{Exponential}(1)$ and $s$ is a scale chosen by bisection to
achieve approximately 30\% censoring:
$$ \text{time}_i = \min(T_i, C_i), \qquad
   \text{status}_i = \mathbf{1}\{T_i \leq C_i\}. $$

**What the model receives.** The model is given only $(Y, \text{time},
\text{status})$ — the same inputs it would receive from a real study. It does not
see $L$, $F$, or $\beta$. Its task is to jointly factorize $Y$ and fit the
survival model to recover $\hat L$, $\hat F$, and $\hat\beta$.

> **Anchoring.** Because $\beta \neq 0$ only on shared factors, survival
> supervision pulls those factors toward the shared component — the model has an
> outcome-level incentive to represent them accurately. This is the intended
> mechanism and the key advantage of supervised over unsupervised factorization.

---

## 4. The Three Scenarios

The scenarios differ only in how the $K$ factors are partitioned into shared vs.
specific. For the primary setting $C=2$, $n_c=150$, $p=1000$, $K=6$:

| Scenario | $|\mathcal{S}|$ shared ($\beta\neq 0$) | specific per cohort ($\beta=0$) | Interpretation |
|---|---|---|---|
| **All shared** | 6 | 0, 0 | Studies effectively one cohort |
| **Hybrid** | 2 | 2, 2 | Both shared and private structure |
| **Nothing shared** | 0 | 3, 3 | No cross-cohort structure |

**Nothing-shared is a deliberate false-positive control.** With $\mathcal{S} =
\varnothing$, $\eta \equiv 0$ and survival carries no factor signal. The correct
result is C-index $\approx 0.5$ and all $|\hat\beta|$ below threshold — the model
should *not* fabricate a shared prognostic factor. An optional knob
(`specific_prognostic`, default off) can give each cohort a within-cohort
prognostic factor if we later want survival signal in this regime.

### Worked example — Hybrid, $C=2$, $n_c=150$, $p=1000$, $K=6$

- **Dimensions:** $n=300$; $Y$ is $300\times1000$, $L$ is $300\times6$, $F$ is
  $1000\times6$, $\beta$ is length 6. Rows 1–150 = cohort 1; 151–300 = cohort 2.
- **Factor labels:** `[shared, shared, specific_1, specific_1, specific_2, specific_2]`.
- **Loadings $L$:** columns 1–2 dense (all 300 patients). Columns 3–4 zero for
  rows 151–300 (private to cohort 1). Columns 5–6 zero for rows 1–150 (private to
  cohort 2). $L$ shows a visible block-sparse pattern.
- **Coefficients:** $\beta = [1.5,\ -1.2,\ 0,\ 0,\ 0,\ 0]$ — only the two shared
  factors drive survival.
- **What a correct model should find:** factors 1–2 with loadings across *both*
  cohorts *and* nonzero $\hat\beta$; factors 3–6 cohort-private with
  $\hat\beta \approx 0$. The unsupervised EBMF benchmark should recover the same
  six gene programs but cannot say which two are prognostic.

**Secondary sensitivity:** a 3-cohort hybrid (2 shared + 1 private per cohort), and
a sweep of the shared-to-specific amplitude ratio
($a_{\text{shared}}/a_{\text{specific}}$) within the hybrid regime to stress the
case where study-specific signal dominates.

---

## 5. Model Arms

Each simulated dataset ($Y$, `time`, `status`, `cohort_id`) is fit by **five
arms** — four supervised (LB/YFB $\times$ $\pm$cohort indicator) plus the
unsupervised EBMF benchmark:

| Arm | Model | Cohort indicator |
|---|---|---|
| **YFB_base** | YFB ($\eta = (YF)\beta$) — recommended | none |
| **YFB_cohort** | YFB | yes |
| **LB_base** | LB ($\eta = L\beta$) | none |
| **LB_cohort** | LB | yes |
| **EBMF** | unsupervised `flashier` (survival-blind) | — |

All supervised arms use a non-negative (`point_exponential`) prior on $L$ and $F$
and a normal prior on $\beta$. The cohort indicator, when present, absorbs
per-cohort gene-mean differences without entering the survival predictor.

**The native-identification test.** The primary analysis is the **base** arms
(no cohort indicator): can the fitted factors separate into shared vs.
study-specific on their own? The $\pm$indicator comparison then quantifies whether
the indicator adds anything.

**Why EBMF.** EBMF is fit to the same $Y$ but is survival-blind, so it has no
$\beta$. It is the unsupervised gold standard for the factorization itself,
answering two questions:

1. Does the supervised model recover the shared/study-specific structure *as well
   as* EBMF (i.e., supervision does not cost structural fidelity)?
2. Does supervision *additionally* yield correct $\beta$ — flagging which factors
   are prognostic — which EBMF fundamentally cannot?

This is the argument for supervision's benefit: same structural recovery, plus correct survival
attribution.

> The EBMF here (fit to each *simulated* $Y$, scored as a competing method) is
> distinct from the EBMF in Section 3.2 (fit to *real* data to build ground-truth
> programs). They do not share a fit.

---

## 6. Evaluation Metrics

For each (scenario × model arm × seed) we record:

- **Shared-factor recovery** — mean best $|\text{cor}|$ between estimated and true
  gene programs ($F$ columns, gene space, comparable across LB/YFB/EBMF), over the
  *true shared* factors.
- **Specific-factor recovery** — same, over the *true study-specific* factors.
- **Specificity-classification accuracy** — label each *estimated* factor
  shared/specific from its per-cohort loading energy (a factor is called specific
  if more than 85% of its squared-loading energy concentrates in a single cohort),
  then compare to the truth. This is the direct test of "does the model natively
  identify the distinction?"
- **Survival-effect recovery** — $|\hat\beta|$ on factors matched to true shared
  (expect above a small threshold) vs. true specific (expect $\approx 0$).
  Reported as true-positive and **false-positive prognostic rates** (the latter is
  the anchoring / nothing-shared check).
- **C-index** — orientation-free held-out concordance, $\max(C, 1-C)$.

**Comparisons reported:** (i) YFB vs. LB; (ii) base vs. $+$cohort-dummy ($\Delta$
in each metric — "are dummies necessary?"); (iii) supervised vs. unsupervised
EBMF.

---

## 7. Reference Data-Generating Code

The following is the complete, self-contained DGP (the project implementation
wraps this same logic with configuration and seeds). A reader can run it directly
given the small `.calibrate` helper.

```{r dgp-reference, eval=FALSE, echo=TRUE}
# Bisection calibration of censoring scale to a target censoring fraction.
.calibrate <- function(ev, base, target = 0.30, n_iter = 40) {
  lo <- 1e-3; hi <- max(ev) * 10
  for (i in seq_len(n_iter)) {
    mid <- sqrt(lo * hi)
    if (mean(base * mid < ev) > target) lo <- mid else hi <- mid
  }
  hi
}

generate_multicohort_data <- function(
    C            = 2L,                 # number of cohorts
    n_per        = rep(150L, C),       # patients per cohort (rows stacked by cohort)
    p            = 1000L,              # genes
    K_shared     = 2L,                 # # shared factors      (beta != 0)
    K_specific   = rep(2L, C),         # # specific factors per cohort (beta = 0)
    F_templates  = NULL,               # p x K_ebmf EBMF programs; NULL -> synthetic sparse
    a_shared     = 1.0,                # amplitude of shared  F columns
    a_specific   = 1.0,                # amplitude of specific F columns
    active_rate  = 0.05,               # synthetic-fallback sparsity
    beta_shared  = NULL,               # length K_shared; NULL -> recycle c(1.5,-1.2,0.8,-0.5)
    offset_sd    = 0.0,                # explicit platform offset SD (0 = off)
    shape        = 1.5, scale0 = 0.01, # Weibull-PH baseline (shape = alpha, scale0 = lambda_0)
    target_cens  = 0.30,               # target censoring fraction
    seed         = 1L) {

  set.seed(seed)
  n  <- sum(n_per)
  K  <- K_shared + sum(K_specific)
  cohort_id <- factor(rep(seq_len(C), times = n_per))

  # label + owner per column (owner 0 = shared)
  labels <- c(rep("shared", K_shared),
              unlist(lapply(seq_len(C), function(cc) rep(paste0("specific_", cc), K_specific[cc]))))
  owner  <- c(rep(0L, K_shared),
              unlist(lapply(seq_len(C), function(cc) rep(cc, K_specific[cc]))))

  # L (n x K): Exponential(1), block-zeroed outside owner cohort
  L <- matrix(rexp(n * K, rate = 1), n, K)
  for (k in seq_len(K)) if (owner[k] != 0L) L[cohort_id != owner[k], k] <- 0

  # F (p x K): EBMF templates (unit-norm) or synthetic sparse (unit-norm)
  Fm <- matrix(0, p, K)
  if (!is.null(F_templates)) {
    idx <- ((seq_len(K) - 1L) %% ncol(F_templates)) + 1L
    Fm  <- F_templates[, idx, drop = FALSE]
    Fm  <- sweep(Fm, 2, sqrt(colSums(Fm^2)) + 1e-10, "/")
  } else {
    for (k in seq_len(K)) {
      active <- sample.int(p, max(1L, round(active_rate * p)))
      Fm[active, k] <- 1 / sqrt(length(active))
    }
  }
  amp <- ifelse(labels == "shared", a_shared, a_specific)   # signal-strength knob
  Fm  <- sweep(Fm, 2, amp, "*")

  # per-gene Gaussian noise: tau_j ~ Gamma(2,2)
  tau <- rgamma(p, shape = 2, rate = 2)
  E   <- sweep(matrix(rnorm(n * p), n, p), 2, sqrt(tau), "/")

  # optional explicit platform offset (off by default)
  offset <- matrix(0, n, p)
  if (offset_sd > 0) for (cc in seq_len(C)) {
    ov <- rnorm(p, 0, offset_sd)
    offset[cohort_id == cc, ] <- matrix(ov, sum(cohort_id == cc), p, byrow = TRUE)
  }

  # assemble + column-center
  Y <- L %*% t(Fm) + offset + E
  Y <- sweep(Y, 2, colMeans(Y), "-")

  # survival from SHARED factors only
  if (is.null(beta_shared))
    beta_shared <- if (K_shared > 0) rep_len(c(1.5, -1.2, 0.8, -0.5), K_shared) else numeric(0)
  beta <- numeric(K); beta[labels == "shared"] <- beta_shared
  eta  <- as.vector(L %*% beta)                  # == 0 in the "nothing shared" scenario

  event <- (-log(runif(n)) / (scale0 * exp(eta)))^(1 / shape)
  cb    <- rexp(n, rate = 1)
  cs    <- .calibrate(event, cb, target = target_cens)
  ctime <- cb * cs
  time   <- pmin(event, ctime)
  status <- as.integer(event <= ctime)

  list(Y = Y, time = time, status = status, cohort_id = cohort_id,
       L_true = L, F_true = Fm, beta_true = beta,
       factor_labels = labels, factor_owner = owner)
}
```

---

## 8. Initial Results

The design above has been implemented and run end-to-end (3 scenarios $\times$ 5
arms $\times$ 5 seeds). Results are preliminary and may be revised after advisor
feedback. Full tables, verification criteria, and the signal-ratio sweep are in
the companion results report.

| Scenario | Arm | Rec. shared | Rec. specific | Spec. acc. | C-index | $\beta$ FP |
|---|---|:--:|:--:|:--:|:--:|:--:|
| all_shared | YFB_base | 0.92 | — | 1.00 | 0.87 | — |
| all_shared | YFB_cohort | 0.94 | — | 0.90 | 0.86 | — |
| all_shared | LB_base | 0.53 | — | 1.00 | 0.74 | — |
| all_shared | EBMF→Cox | 0.82 | — | 0.90 | 0.86 | — |
| hybrid | YFB_base | 0.90 | 0.84 | 0.97 | 0.81 | 0.50 |
| hybrid | YFB_cohort | 0.92 | 0.92 | 1.00 | 0.81 | 0.20 |
| hybrid | LB_base | 0.72 | 0.67 | 0.83 | 0.76 | 0.20 |
| hybrid | EBMF→Cox | 0.68 | 0.78 | 0.80 | **0.72** | — |
| nothing_shared | YFB_base | — | 0.89 | 1.00 | 0.55 | 0.03 |
| nothing_shared | YFB_cohort | — | 0.91 | 1.00 | 0.54 | 0.03 |
| nothing_shared | LB_base | — | 0.78 | 0.87 | 0.54 | 0.07 |
| nothing_shared | EBMF→Cox | — | 0.70 | 0.80 | 0.55 | — |

*Mean over 5 seeds; LB_cohort omitted for brevity. **Column definitions:**
**Rec. shared / Rec. specific** — gene-program recovery, the mean best
$|\text{correlation}|$ between estimated and true gene programs over the true
shared / study-specific factors (1 = perfect). **Spec. acc.** —
specificity-classification accuracy, the fraction of estimated factors correctly
labeled shared vs. study-specific from their per-cohort loading energy.
**C-index** — held-out (external) concordance on the 25% test split (0.5 = chance).
**$\beta$ FP** — false-positive prognostic rate: fraction of true study-specific
factors ($\beta = 0$) wrongly given a survival effect (lower is better).
**EBMF→Cox** is the two-stage baseline: an unsupervised (survival-blind) EBMF
factorization followed by a Cox model fit on the EBMF factor scores. It has no
$\beta$ false-positive rate (its factors are not chosen for prognostic relevance);
its C-index quantifies how well the "factorize first, predict second" pipeline
generalizes, for direct comparison against the joint supervised model. In the
hybrid scenario the joint model leads by $\approx 0.09$ (0.81 vs.\ 0.72); in
all_shared the two-stage baseline nearly matches (every factor is prognostic), and
in nothing_shared both sit at chance.*

```{r proposal-fig-setup}
#| include: false
suppressPackageStartupMessages({ library(ggplot2); library(dplyr); library(tidyr) })
OUT_MC  <- "../../results/multi_cohort_sim/outputs"
ex_path <- file.path(OUT_MC, "multicohort_sim_example.rds")
have_ex <- file.exists(ex_path)
if (have_ex) ex <- readRDS(ex_path)
res_path <- file.path(OUT_MC, "multicohort_sim_results.csv")
have_res <- file.exists(res_path)
if (have_res) res <- read.csv(res_path, stringsAsFactors = FALSE)
```

```{r fig-loadings-proposal, fig.width=9, fig.height=7.8, eval=have_ex && have_res, fig.cap="TOP — Patient loading heatmaps for the hybrid scenario (seed 1, training split). Rows = patients grouped by cohort (left sidebar: blue = cohort 1, salmon = cohort 2); columns = factors F1-F6. Top annotation: true factor type (Active = shared/prognostic; Shrunk = study-specific/non-prognostic) and estimated |beta-hat| from YFB. LEFT: true L (ground-truth block-sparse structure); RIGHT: estimated L from YFB recovering the same structure with no indicator columns. BOTTOM — Held-out C-index by scenario and arm (mean +/- SE over 5 seeds; dashed line = chance 0.5). The joint supervised model (YFB) and the two-stage EBMF->Cox baseline differ most in the hybrid scenario: when only some factors are prognostic, jointly modelling genomics and survival recovers the prognostic structure that the survival-blind EBMF factorization cannot preferentially sharpen. In all_shared (every factor prognostic) the two-stage baseline nearly matches; in nothing_shared both sit at chance, as designed."}
suppressPackageStartupMessages({ library(pheatmap); library(gridExtra); library(grid) })

# ---- Top: True vs Estimated L heatmaps ----
d    <- ex$hybrid$data
spl  <- ex$hybrid$split
cid  <- droplevels(d$cohort_id[spl$train_idx])
fit  <- ex$hybrid$YFB_base
col_idx <- fit$match$match          # col_idx[j] = estimated col for true factor j
labs    <- d$factor_labels; flab <- paste0("F", seq_along(labs))
ord     <- order(cid); cid_ord <- cid[ord]
gap_row <- which(diff(as.integer(cid_ord)) != 0)
Ltrue <- d$L_true[spl$train_idx, , drop = FALSE][ord, , drop = FALSE]
Lhat  <- fit$EL[ord, col_idx, drop = FALSE]   # select estimated cols in true-factor order
rownames(Ltrue) <- rownames(Lhat) <- paste0("P", seq_len(nrow(Ltrue)))
colnames(Ltrue) <- colnames(Lhat) <- flab
row_anno  <- data.frame(Cohort = cid_ord, row.names = rownames(Ltrue))
cohort_pal <- setNames(c("#2166AC", "#D6604D"), levels(cid_ord))
col_anno  <- data.frame(
  Type     = factor(ifelse(labs == "shared", "Active", "Shrunk"), levels = c("Active","Shrunk")),
  Abs_Beta = abs(fit$EBeta[col_idx]), row.names = flab)   # |β̂| in true factor order
anno_colors <- list(Cohort = cohort_pal,
                    Type   = c(Active = "#D62728", Shrunk = "#BDBDBD"),
                    Abs_Beta = c("#FFFFFF", "#08306B"))
heat_pal <- colorRampPalette(c("#FFFFFF", "#FCBBA1", "#FB6A4A", "#A50F15"))(100)
ph_args  <- list(cluster_rows=FALSE, cluster_cols=FALSE, show_rownames=FALSE,
                 show_colnames=TRUE, color=heat_pal, annotation_row=row_anno,
                 annotation_col=col_anno, annotation_colors=anno_colors,
                 gaps_row=gap_row, border_color=NA, fontsize_col=9, fontsize=9,
                 legend=TRUE, annotation_legend=TRUE, silent=TRUE)
p_true <- do.call(pheatmap, c(list(Ltrue, main="True L"), ph_args))
p_hat  <- do.call(pheatmap, c(list(Lhat,  main="Estimated L (YFB, no indicator)"), ph_args))

# ---- Bottom: held-out C-index — joint model vs. two-stage EBMF->Cox ----
arm_disp <- c(YFB_base="YFB", YFB_cohort="YFB +cohort",
              LB_base="LB", LB_cohort="LB +cohort", EBMF="EBMF→Cox")
cidx <- res %>%
  filter(arm %in% names(arm_disp)) %>%
  group_by(scenario, arm) %>%
  summarise(C = mean(c_index, na.rm=TRUE),
            se = sd(c_index, na.rm=TRUE)/sqrt(sum(!is.na(c_index))), .groups="drop") %>%
  mutate(arm = factor(arm_disp[arm], levels = unname(arm_disp)),
         scenario = factor(scenario, levels=c("all_shared","hybrid","nothing_shared")))
bar_fill <- c("YFB"="#1f77b4", "YFB +cohort"="#4c9bd4",
              "LB"="#7f7f7f", "LB +cohort"="#bcbcbc", "EBMF→Cox"="#d62728")
p_cidx <- ggplot(cidx, aes(arm, C, fill = arm)) +
  geom_hline(yintercept = 0.5, linetype = 2, colour = "grey50") +
  geom_col(alpha = 0.9) +
  geom_errorbar(aes(ymin = C - se, ymax = C + se), width = 0.25) +
  facet_wrap(~scenario) +
  scale_fill_manual(values = bar_fill, guide = "none") +
  coord_cartesian(ylim = c(0.4, 0.9)) +
  labs(x = NULL, y = "held-out C-index",
       title = "Survival validation: joint SSBMF vs. two-stage EBMF→Cox") +
  theme_bw(base_size = 9) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1),
        plot.title = element_text(size = 9, face = "bold"))

# ---- Assemble: heatmaps (top), C-index panel (bottom) ----
gridExtra::grid.arrange(
  p_true$gtable, p_hat$gtable, ggplotGrob(p_cidx),
  layout_matrix = rbind(c(1, 2), c(3, 3)),
  heights = c(1.55, 1))
```

```{r fig-km-scenarios, fig.width=8.2, fig.height=2.8, eval=have_ex, fig.cap="Held-out Kaplan-Meier curves by scenario (YFB_base, seed 1; test split). Patients split at the median predicted risk score. In all_shared every factor is prognostic and separation is strong. In hybrid (the realistic case — shared prognostic factors coexist with study-specific non-prognostic ones) the model cleanly separates risk groups despite never observing the factor labels. In nothing_shared survival is null by construction and the curves correctly overlap (C near 0.5)."}
suppressPackageStartupMessages(library(survival))
scen_labels <- c(all_shared = "All shared", hybrid = "Hybrid",
                 nothing_shared = "Nothing shared")
scen_cols   <- list(low = "steelblue", high = "firebrick")
par(mfrow = c(1, 3), mar = c(4.8, 3.5, 2, 0.5), mgp = c(2, 0.6, 0), cex = 0.82)
for (sc in c("all_shared", "hybrid", "nothing_shared")) {
  if (!sc %in% names(ex) || !"YFB_base" %in% names(ex[[sc]])) {
    plot.new(); title(main = scen_labels[sc]); next
  }
  p  <- ex[[sc]]$YFB_base$pred
  cc <- concordance(Surv(p$time, p$status) ~ p$risk)$concordance
  risk <- if (cc < 0.5) -p$risk else p$risk
  grp  <- factor(ifelse(risk > median(risk), "high", "low"), levels = c("low", "high"))
  sf   <- survfit(Surv(p$time, p$status) ~ grp)
  lr   <- survdiff(Surv(p$time, p$status) ~ grp)
  pval <- pchisq(lr$chisq, df = 1, lower.tail = FALSE)
  plot(sf, col = c(scen_cols$low, scen_cols$high), lwd = 1.8,
       main = scen_labels[sc], xlab = "Time", ylab = "Survival",
       bty = "l", ylim = c(0, 1))
  mtext(sprintf("C = %.2f,  p = %.2g", max(cc, 1 - cc), pval),
        side = 1, line = 3.8, cex = 0.72, col = "grey10")
  if (sc == "all_shared")
    legend("topright", c("low risk", "high risk"),
           col = c(scen_cols$low, scen_cols$high),
           lwd = 1.8, bty = "n", cex = 0.78)
}
```

**The main findings:**

1. **The model natively identifies the shared / study-specific distinction.** In
   the hybrid scenario, YFB classifies factors as shared vs. specific with
   accuracy 0.97 (base) to 1.00 (+cohort indicator), purely from per-cohort loading
   energy — no indicator columns required. This directly answers the central
   question.
2. **It does not fabricate prognostic signal when none exists.** In nothing_shared
   ($\eta \equiv 0$ by construction), every arm sits at C-index $\approx 0.54$ and
   the $\beta$ false-positive rate is 0.03–0.07 — the model does not invent a
   shared prognostic factor out of purely study-specific structure.
3. **The joint model beats the two-stage EBMF→Cox baseline where it matters most.**
   In the hybrid scenario — only some factors prognostic, the realistic case — the
   joint YFB reaches C-index 0.81 vs. 0.72 for "factorize first, fit Cox second"
   ($\approx 0.09$ gain). Because EBMF chooses its factors without seeing the
   outcome, it cannot preferentially sharpen the prognostic ones; jointly modelling
   genomics and survival can. In all_shared (every factor prognostic) the two-stage
   baseline nearly matches (0.86 vs. 0.87), and in nothing_shared both sit at chance
   — exactly where the gap should and should not appear. This is the core benefit
   of supervision.
4. **Supervision does not cost structural fidelity.** YFB recovers shared programs
   at $|\text{cor}| \approx 0.90$–0.94 and specific programs at 0.84–0.92, matching
   or exceeding the non-negative EBMF benchmark (0.68–0.82).
5. **The cohort indicator helps under genuine heterogeneity, and is redundant
   otherwise.** In hybrid, adding it halves the $\beta$ false-positive rate
   (0.50 $\to$ 0.20) and lifts specificity accuracy to 1.00; in all_shared it adds
   nothing. This is the quantitative answer to "are the dummy indicators
   necessary?"
6. **YFB is more reliable than LB**, which collapses to near-zero recovery on some
   seeds — consistent with YFB being the recommended model.

---

## 9. Open Questions for Discussion

1. **Specificity definition.** We encode study-specific factors via block-zero
   loadings (a factor is active in exactly one cohort). Is this the structure you
   intend, or should specific factors also have *non-overlapping gene programs*
   per cohort?
2. **Nothing-shared survival.** Default leaves survival null in the nothing-shared
   scenario (false-positive control). Do you prefer instead a within-cohort
   prognostic specific factor so each cohort has its own survival signal?
3. **Cohort count.** Primary is $C=2$ (matches the real training data and the
   fixed-effects rationale); a $C=3$ hybrid is included as sensitivity. Is 2
   sufficient for the methods claim, or should 3 be primary?
4. **Signal ratio.** Default $a_{\text{shared}} = a_{\text{specific}}$. Should the
   hybrid scenario stress a regime where study-specific variance *dominates*
   shared variance (the harder, more realistic case)?

5. **Cohort indicator: design decision and remaining false positives.** In the
   hybrid scenario, adding the cohort dummy indicator halves the $\beta$
   false-positive rate (0.50 $\to$ 0.20), but does not eliminate it — 20% of
   study-specific factors still receive a spurious prognostic coefficient even
   with the indicator present.

   The mechanism is confounding: study-specific factors are active in exactly
   one cohort, and that cohort's patients have different survival trajectories
   (driven by the shared prognostic factors with different mean loadings across
   cohorts). Without a cohort label, the model absorbs this cohort-level
   survival difference by assigning $\hat\beta$ to the specific factors. The
   indicator breaks this confounding, but residual within-cohort correlation
   (finite sample, $n_c = 150$) accounts for the remaining 0.20.

   Two open sub-questions:

   (a) *Should the cohort indicator be part of the default recommended
   configuration?* Our benchmark on real PDAC data achieves C = 0.636
   **without** the indicator, suggesting confounding is tolerable in practice.
   But this simulation targets the equal-signal hybrid ($a_\text{shared} =
   a_\text{specific}$); whether the PDAC cohorts sit in a harder regime (Q4
   above) where the FP problem worsens is unknown.

   (b) *What additional approaches could reduce the residual false positives?*
   Three candidates:
   - **Signal-ratio sweep** (planned): stress the specific-dominant regime
     ($a_\text{specific} \gg a_\text{shared}$) to quantify how badly the FP
     rate grows and whether the indicator remains sufficient.
   - **Tighter $\beta$ prior**: a tighter normal prior on $\beta$ would shrink
     small spurious coefficients toward zero without the full collapse that the
     `point_normal` prior causes. This has not been tested in this setting.
   - **Cohort-stratified Cox baseline**: replace the shared baseline hazard with
     cohort-stratified strata, natively absorbing cohort-level survival
     differences without indicator columns. This is a more substantial model
     change.

   **Preliminary sweep results** (hybrid, 5 seeds, 5 variance ratios): the FP
   rate is bounded at 0.20–0.50 across all ratios and does not grow
   monotonically. The indicator's advantage is largest at equal signal and
   narrows at high specific-variance ratios. Structural recovery and C-index
   are stable throughout. Provisional recommendation: include the cohort
   indicator as default, and use a $|\hat\beta|$ magnitude threshold rather
   than binary nonzero classification to manage residual false positives.



---

## 10. Recommendation

**Recommended model configuration.** The YFB variant — linear predictor
$\eta = (YF)\beta$, computed from factor *scores* rather than inferred
loadings — is the recommended form. On real PDAC data it achieves external
concordance $C = 0.636$ across five held-out cohorts, exceeding
the DeSurv baseline and the LB alternative ($\eta = L\beta$). Key required settings:
(i) per-platform $z$-standardization before pooling — without
it, ten of twelve preprocessing configurations collapse $\hat\beta \to 0$;
(ii) a `normal` prior on $\beta$ (the `point_normal` prior causes the same
collapse under YFB); and (iii) $K$ selected by cross-validation with the 1-SE
rule ($K = 7$ on the PDAC benchmark). DeSurv-aligned gene selection
(combined-rank, top-3000 per cohort) provides a modest additional gain
($\Delta C \approx +0.012$).

**When joint modeling beats the two-stage baseline.** The simulation makes the
value-add precise: in the *hybrid* scenario — shared prognostic programs
coexist with study-specific non-prognostic ones, the realistic structure for
multi-cohort PDAC data — the joint YFB model achieves $C = 0.81$ versus $0.72$
for EBMF→Cox ($+0.09$). When every factor is prognostic (all-shared) the
two-stage baseline nearly matches, and in the null scenario both correctly sit
at chance. The gain is therefore concentrated exactly where it matters: when
the model must discriminate prognostic shared programs from non-prognostic
platform effects *before* fitting survival. EBMF, being survival-blind,
cannot preferentially sharpen the prognostic latent dimensions; the CAVI loop
does this automatically by coupling the genomics and survival likelihoods.

**Why this matters for multi-cohort genomics.** Three properties distinguish
SSBMF from alternatives:

1. **Joint inference** — genomics and survival are learned simultaneously. The
   survival likelihood shapes which latent programs are represented, not just
   which existing programs are associated after the fact.
2. **Interpretable factors** — each column of $F$ is a sparse gene expression
   program (a GEP), directly analogous to the biological signatures identified
   by unsupervised NMF or EBMF. The coefficient $\beta_k$ then quantifies the
   *prognostic weight* of that program, giving a mechanistic link between
   molecular programs and outcome.
3. **Cross-cohort generalization** — by explicitly separating shared from
   study-specific factors at the factorization stage, the model produces risk
   scores that transfer across heterogeneous platforms and cohorts. The external
   $C = 0.636$ on five held-out PDAC cohorts, trained on two, demonstrates this.
   A model that cannot identify and discard platform-specific variation will
   overfit to cohort identity and generalize poorly.
