---
title: "Multi-Cohort Simulation Study: Results (Working Draft)"
subtitle: "Shared vs. study-specific factor recovery — SSBMF (YFB/LB) vs. EBMF"
author: "Andrew Walther"
date: "2026-06-14"
format:
  pdf:
    latex-engine: xelatex
    keep-tex: false
    fig-pos: "htbp"
    include-in-header:
      text: |
        \usepackage{booktabs}
        \usepackage{geometry}
        \usepackage{titling}
        \geometry{margin=0.55in}
        \renewcommand{\familydefault}{\sfdefault}
        \setlength{\droptitle}{-1.8em}
        \pretitle{\begin{center}\large\bfseries}
        \posttitle{\par\end{center}\vspace{-0.3em}}
  html:
    toc: true
    code-fold: true
    theme: cosmo
fontsize: 9pt
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(knitr)
suppressPackageStartupMessages({ library(dplyr); library(tidyr); library(ggplot2) })

OUT <- "../../results/multi_cohort_sim/outputs"
res <- read.csv(file.path(OUT, "multicohort_sim_results.csv"), stringsAsFactors = FALSE)

scen_levels <- c("all_shared", "hybrid", "nothing_shared")
arm_levels  <- c("EBMF", "YFB_base", "YFB_cohort", "LB_base", "LB_cohort")
res$scenario <- factor(res$scenario, levels = scen_levels)
res$arm      <- factor(res$arm, levels = arm_levels)

n_seeds <- length(unique(res$seed))

mean_sd <- function(x) {
  x <- x[!is.na(x)]
  if (!length(x)) return("--")
  sprintf("%.2f (%.2f)", mean(x), sd(x))
}
```

> **Status: working draft.** This report demonstrates the simulation pipeline on
> the design in the companion *proposal* (`multicohort_sim_proposal_06_14_26`). It
> is intended to carry preliminary results into the June 15 advisor meeting and
> will be **revised after feedback** (DGP, scenarios, metrics, or methodology may
> change). Results are means over `r n_seeds` random seeds.

## 1. What was run

Three scenarios, each generated with $C=2$ cohorts, $n_c=150$, $p=1000$, total
$K=6$ true factors, gene programs drawn from an EBMF fit to the real training
cohorts (TCGA_PAAD + CPTAC), and a Weibull-PH survival outcome (~30% censoring)
driven by the shared factors only.

| Scenario | shared (β≠0) | specific per cohort (β=0) |
|---|---|---|
| **all_shared** | 6 | 0, 0 |
| **hybrid** | 2 | 2, 2 |
| **nothing_shared** | 0 | 3, 3 |

Five arms per scenario: **YFB_base**, **YFB_cohort** (+cohort dummy),
**LB_base**, **LB_cohort**, and **EBMF** (unsupervised, survival-blind). Metrics:
gene-program recovery (mean best $|\text{cor}|$ of estimated vs. true programs,
reported separately for shared and specific factors), the **fraction of true
factors recovered** at $|\text{cor}|>0.7$, specificity-classification accuracy,
held-out C-index (orientation-free), and $\beta$ false-positive prognostic rate.

## 2. Headline results (mean over seeds; SD in parentheses)

```{r main-table}
summ <- res %>%
  group_by(scenario, arm) %>%
  summarise(
    `Rec. shared`    = mean_sd(rec_shared),
    `Rec. specific`  = mean_sd(rec_specific),
    `Frac shared>.7` = mean_sd(frac_shared),
    `Spec. acc.`     = mean_sd(spec_acc),
    `C-index`        = mean_sd(c_index),
    `β FP rate`      = mean_sd(beta_fp_rate),
    .groups = "drop"
  ) %>% arrange(scenario, arm)

kable(summ, align = "llcccccc",
      caption = "Recovery, specificity classification, survival prediction, and false-positive prognostic rate by scenario and model arm. EBMF is survival-blind (no C-index, no β).")
```

## 3. Figures

```{r fig-cindex, fig.width=7, fig.height=2.8, fig.cap="Held-out C-index by scenario and arm (supervised arms only; EBMF has no survival model). Dashed line = chance (0.5). Nothing-shared sits at chance, as designed."}
res %>% filter(arm != "EBMF") %>%
  ggplot(aes(arm, c_index, fill = arm)) +
  geom_hline(yintercept = 0.5, linetype = 2, colour = "grey50") +
  stat_summary(fun = mean, geom = "col", alpha = 0.85) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.3) +
  facet_wrap(~scenario) +
  coord_cartesian(ylim = c(0.4, 0.9)) +
  labs(x = NULL, y = "held-out C-index") +
  theme_bw(base_size = 9) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1), legend.position = "none")
```

```{r fig-recovery, fig.width=7, fig.height=2.8, fig.cap="Gene-program recovery (mean best |cor| of estimated vs. true programs), shared vs. specific factors. EBMF (unsupervised) sets the recovery ceiling; supervised arms trade some recovery for survival prediction."}
res %>%
  select(scenario, arm, rec_shared, rec_specific) %>%
  pivot_longer(c(rec_shared, rec_specific),
               names_to = "type", values_to = "rec") %>%
  mutate(type = recode(type, rec_shared = "shared", rec_specific = "specific")) %>%
  filter(!is.na(rec)) %>%
  ggplot(aes(arm, rec, fill = type)) +
  stat_summary(fun = mean, geom = "col", position = "dodge", alpha = 0.85) +
  facet_wrap(~scenario) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = NULL, y = "factor recovery |cor|", fill = "factor") +
  theme_bw(base_size = 9) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
```

```{r fig-heatmap, fig.width=7, fig.height=3, fig.cap="Example (seed 1, hybrid): per-factor recovery |cor| for each arm against the six true factors (2 shared, 4 specific). Bright = recovered."}
ex_path <- file.path(OUT, "multicohort_sim_example.rds")
if (file.exists(ex_path)) {
  ex <- readRDS(ex_path)
  d  <- ex$hybrid$data
  arms_ex <- intersect(arm_levels, names(ex$hybrid))
  heat <- do.call(rbind, lapply(arms_ex, function(a) {
    bc <- ex$hybrid[[a]]$match$best_cor
    data.frame(arm = a,
               factor = paste0("F", seq_along(bc), "_", substr(d$factor_labels, 1, 4)),
               cor = bc)
  }))
  heat$arm    <- factor(heat$arm, levels = arm_levels)
  heat$factor <- factor(heat$factor, levels = unique(heat$factor))
  ggplot(heat, aes(factor, arm, fill = cor)) +
    geom_tile(colour = "white") +
    geom_text(aes(label = sprintf("%.2f", cor)), size = 2.6) +
    scale_fill_gradient(low = "grey92", high = "steelblue", limits = c(0, 1)) +
    labs(x = "true factor (label)", y = NULL, fill = "|cor|") +
    theme_minimal(base_size = 9) +
    theme(axis.text.x = element_text(angle = 35, hjust = 1))
} else {
  cat("Example fits not found.")
}
```

## 4. What the results say

**1. The model does not fabricate prognostic signal when none exists.** In the
**nothing_shared** scenario (survival is null by construction), every arm sits at
C-index $\approx 0.54$ and the $\beta$ false-positive prognostic rate is very low
(0.03–0.07). The supervised model does not invent a shared prognostic factor out
of purely study-specific structure — the key safety result.

**2. Survival prediction is strong when shared prognostic signal exists.** In
**all_shared** and **hybrid**, the supervised arms reach C-index $\approx
0.74$–$0.87$ (YFB highest). EBMF, being survival-blind, produces no risk score at
all — precisely the gap the supervised model fills, and the central value-add
argument versus an unsupervised factorization.

**3. The supervised model recovers the gene programs as well as — or better than —
the unsupervised EBMF benchmark.** On a fair footing (both using non-negative,
NMF-style priors, Section 5), **YFB** recovers shared programs at $|\text{cor}|
\approx 0.90$–$0.94$ and specific programs at $\approx 0.84$–$0.92$, matching or
exceeding non-negative EBMF (0.68–0.82). So supervision does **not** cost
structural fidelity here; it adds correct survival attribution on top of
equally-good factor recovery.

**4. Specificity is identified natively, without dummy indicators.** YFB's
specificity-classification accuracy is $\approx 0.97$–$1.00$ across scenarios:
shared factors are recovered with cross-cohort loadings, study-specific factors as
cohort-private, with no indicator columns required. This directly addresses the
meeting's central question — the base model **does** separate shared from
study-specific structure on its own.

**5. The cohort dummy indicator helps under genuine heterogeneity, and is
redundant otherwise.** In **hybrid**, **YFB_cohort** improves on YFB_base across
the board — specificity accuracy
`r sprintf("%.2f", mean(res$spec_acc[res$scenario=="hybrid" & res$arm=="YFB_cohort"], na.rm=TRUE))`
vs. `r sprintf("%.2f", mean(res$spec_acc[res$scenario=="hybrid" & res$arm=="YFB_base"], na.rm=TRUE))`,
and a lower $\beta$ false-positive rate
(`r sprintf("%.2f", mean(res$beta_fp_rate[res$scenario=="hybrid" & res$arm=="YFB_cohort"], na.rm=TRUE))`
vs. `r sprintf("%.2f", mean(res$beta_fp_rate[res$scenario=="hybrid" & res$arm=="YFB_base"], na.rm=TRUE))`).
In **all_shared** (no cross-cohort differences) the indicator is unnecessary. So:
helpful when heterogeneity is real, harmless-to-redundant when it is not.

**6. YFB is more robust than LB.** Across scenarios YFB recovers factors more
reliably than LB, which is more sensitive to initialization (lower mean recovery,
higher seed-to-seed variance). This is consistent with YFB being the recommended
model.

## 5. Methodology notes (grounded in prior project work)

- **Non-negativity is the model's established assumption, not a tuning choice.**
  The SSBMF models place a `point_exponential` prior on both $L$ and $F$ — an
  NMF-style non-negative factorization. This is documented prior work: in the
  initialization study (DECISIONS, 2026-05-06, *"Phase 2 — initialization
  constraints"*) the SVD initialization was deliberately set to `abs()` because it
  is *"equally valid for the non-negative point_exponential prior."* The simulation
  therefore generates **non-negative** gene programs (the real-data EBMF templates
  are taken in absolute value), and the unsupervised EBMF benchmark is run with a
  **non-negative** prior too, so both methods share the same representational
  assumption. An earlier draft that used *signed* EBMF templates against the
  non-negative prior depressed supervised recovery to $\approx 0.45$ purely from
  the mismatch — an artifact that the corrected, prior-consistent setup removes
  (recovery $\approx 0.90$).
- **Seed-to-seed spread is replicate Monte-Carlo variability, not initialization
  noise.** SVD initialization is deterministic (re-fitting identical data gives
  identical results), so the variability across seeds reflects different *simulated
  datasets*, which is exactly what averaging over replicates is meant to summarize.
  The choice of a single SVD start is supported by prior work: the 30-restart
  multi-initialization sweep (DECISIONS, 2026-05-06, *"Phase 3 — multi-initialization"*)
  found the CAVI landscape near-unimodal (ELBO range 0.3%) and concluded
  `n_init = 1`; the multi-start wrapper (`code/fit_modular_multistart.R`) is retained
  as a diagnostic but is not needed here. LB shows more replicate variance than YFB.
- **Robust to $K$ misspecification.** Primary runs fix $K = 6$ (true $K$) for a
  clean recovery test, but recovery is stable when $K$ is over-specified — for the
  hybrid scenario, YFB recovery is $0.88/0.91/0.88$ (shared) and $0.86/0.91/0.91$
  (specific) at $K = 6/8/10$ — consistent with the ARD-style pruning the model uses
  in practice (`auto_prune_K`), and with the project's CV-based selection
  (`select_K_cv`, 1-SE) used on the real data.
- **Survival uses the established `normal` β prior.** `point_normal` is avoided
  because it collapses $\beta \to 0$ under YFB (DECISIONS, 2026-05-06), matching the
  benchmark configuration.
- **Equal signal amplitudes (next step).** Shared and study-specific factors
  currently have equal variance; a `shared_signal_ratio` sweep (study-specific
  variance dominating) is the most realistic stress test and is supported by the
  generator.

*With the prior-consistent setup, these results are a positive evaluation of the
model. They will be revised per advisor feedback before any are treated as final.*

```{r reproduce, echo=TRUE, eval=FALSE}
# Reproduce:
#   Rscript results/multi_cohort_sim/run_multicohort_sim.R          # full (5 seeds)
#   Rscript results/multi_cohort_sim/run_multicohort_sim.R --quick  # smoke test
```
