---
title: "SSBMF: LB vs. YFB Benchmark Report"
subtitle: "Supervised Bayesian Matrix Factorization — 2026-05-05 Re-Benchmark"
author: "Andrew Walther"
date: "2026-05-05"
format:
  pdf:
    engine: xelatex
    toc: true
    toc-depth: 2
    number-sections: true
    geometry: "top=0.85in, bottom=0.85in, left=1in, right=1in"
    fontsize: 11pt
    linestretch: 1.1
    colorlinks: true
    linkcolor: "NavyBlue"
    include-in-header:
      text: |
        \usepackage{float}
        \floatplacement{figure}{H}
        \usepackage{placeins}
        \usepackage{booktabs}
        \usepackage{xcolor}
        \usepackage{mdframed}
        \definecolor{keyblue}{RGB}{31,119,180}
        \definecolor{lightgray}{RGB}{245,245,245}
  html:
    theme: cosmo
    toc: true
    toc-depth: 2
    number-sections: true
    embed-resources: true
    self-contained: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(knitr)
library(yaml)

knitr::opts_chunk$set(
  fig.align = "center",
  fig.pos   = "H",
  out.width = "80%",
  dpi       = 150
)

OUT_LB  <- "../../results/benchmark_sim/outputs/LB_benchmark"
OUT_YFB <- "../../results/benchmark_sim/outputs/YFB_benchmark"

read_csv_safe <- function(path) {
  if (file.exists(path)) read.csv(path, stringsAsFactors = FALSE) else NULL
}

lb_tcga  <- read_csv_safe(file.path(OUT_LB, "LB_benchmark_results_tcga_only.csv"))
lb_cptac <- read_csv_safe(file.path(OUT_LB, "LB_benchmark_results_cptac_only.csv"))
lb_merge <- read_csv_safe(file.path(OUT_LB, "LB_benchmark_results_merged.csv"))

yfb_tcga  <- read_csv_safe(file.path(OUT_YFB, "YFB_benchmark_results_tcga_only.csv"))
yfb_cptac <- read_csv_safe(file.path(OUT_YFB, "YFB_benchmark_results_cptac_only.csv"))
yfb_merge <- read_csv_safe(file.path(OUT_YFB, "YFB_benchmark_results_merged.csv"))

# Archived 04_29_26 baseline (v1 preprocessing, per-cohort top-2000 gene filter)
archived_baseline <- data.frame(
  train_mode  = c("tcga_only", "cptac_only", "merged"),
  K_eff       = c(4, 4, 7),
  n_iter      = c(41, 58, 66),
  final_elbo  = c(-935867.6, -797111.0, -1020985.6),
  median_ext  = c(0.6021, 0.6278, 0.50),
  preprocessing = "v1 (per-cohort top-2000 gene filter; 407–644 common genes)",
  stringsAsFactors = FALSE
)
```

## Key Findings {.unnumbered}

This report presents the 2026-05-05 re-benchmark of Supervised Bayesian Matrix Factorization
(SSBMF) following three targeted fixes:

- **Phase A (LB model):** Reverted CAVI inner loop from $\beta \to L \to F$ (Cluster A) back
  to $L \to F \to \beta$ (Gauss-Seidel). Proof: 18-unit ELBO gap on identical data.
  Result: ELBO and convergence now match the archived baseline exactly.

- **Phase B (YFB model):** Normalized EF columns to unit L2 norm before computing
  $Z_F = Y \cdot E_F^{\text{norm}}$. Without normalization, $\|Z_F^{(k)}\| \approx O(\sqrt{pn})$
  drives spike-and-slab to pin $\hat{\beta} \to 0$.

- **Phase C (YFB model):** After CAVI convergence, check training concordance;
  if $C_{\text{train}} < 0.5$, negate all $\hat{\beta}$. The pmax(SVD) initialization
  discards sign information, causing $Z_F$ to be anti-correlated with survival.

**Key result.** The YFB model with `normal` prior substantially outperforms the LB model
on external PDAC validation. YFB (tcga\_only training) achieves median external
C-index $\approx 0.55$–$0.63$, compared to LB median $\approx 0.39$–$0.45$.

---

## Preprocessing Pipeline (v2)

The v2 preprocessing pipeline (introduced during Cluster A development) applies to all
training cohorts and uses the training gene set for all external validation predictions.

1. **Load**: read raw expression matrix per cohort
2. **Log$_2$-transform**: apply $\log_2(x+1)$ for RNA-seq cohorts; skip for proteomics/microarray
3. **Intersect-first**: for merged training, intersect gene lists across cohorts before filtering
4. **Top-2000 variable genes**: select the 2,000 highest-variance genes from the training matrix
5. **Quantile normalisation**: for merged training, QN across pooled cohorts
6. **Rank-transform subjects**: convert each subject's gene expression to ranks (removes
   distributional differences between cohorts at prediction time)
7. **Column-centering**: subtract column means so $Y$ has zero mean per gene

**External validation note.** External cohorts are preprocessed with steps 2+6+7 only
(no variance filter applied), then subset to the training gene set. This gives 1,601–1,919
common genes per cohort — substantially more than the archived v1 pipeline, which applied
a per-cohort top-2000 filter first (yielding only 407–644 common genes). **C-indices
are therefore not directly comparable between the v1 archived baseline and these v2 results.**

---

## Model Descriptions

### LB Model (Cluster A: $\hat{\eta} = E[L]\hat{\beta}$)

The LB model fits Supervised Bayesian MF via CAVI with four modular updates:
$q(L)$, $q(F)$, $q(\beta)$, $q(\tau)$. The survival linear predictor is
$\eta_i = \sum_k \bar{l}_{ik} \hat{\beta}_k$ where $\bar{L} = E_q[L]$ is the
EBNM-shrunk posterior mean of the loading matrix.

**Train/test issue.** Training uses EBNM-shrunk $E[L]$ (modulated by both genomics and
survival via $A_{\text{surv}}$). Test prediction uses OLS projection
$L_{\text{test}} = Y_{\text{test}} \cdot \hat{F} \cdot (\hat{F}^T\hat{F})^{-1}$. These are
structurally different quantities.

**Phase A fix.** The Cluster A commit (55500fd) accidentally changed the factor update
order from $L \to F \to \beta$ (Gauss-Seidel, correct) to $\beta \to L \to F$ (Jacobi for
$\beta$, inferior). Reverting to $L \to F \to \beta$ recovers
ELBO $= {-}935{,}867.6$ at iter 41 on tcga\_only (matches archived baseline exactly).

### YFB Model (Cluster B: $\hat{\eta} = Z_F\hat{\beta}$, where $Z_F = YE_F^{\text{norm}}$)

The YFB model replaces the survival predictor with $\eta_i = Z_{F,i} \hat{\beta}$ where
$Z_F = Y \cdot E_F^{\text{norm}}$ is a matrix of observed projection scores.
Training and test prediction use identical formulas (no EBNM-shrunk $L$ at test time),
closing the train/test mismatch of the LB model.

**Factor updates.** $q(L)$ is pure-genomics (L does not enter Cox under YFB).
$q(F)$ uses $\alpha_F = 0$ (pure-genomics), preventing a positive-feedback instability.
$q(\beta)$ uses $Z_F^{(k)}$ (observed) as the covariate instead of $E[l_k]$.

**Phase B fix.** EF columns normalized to unit L2 norm before $Z_F = Y \cdot E_F^{\text{norm}}$.
**Phase C fix.** Post-convergence sign correction: negate $\hat{\beta}$ if training $C < 0.5$.

**Recommended prior for YFB.** `normal` (pure Gaussian) is required; `point_normal`
(spike-and-slab) pins $\hat{\beta} \to 0$ on PDAC data even after the Phase B scale fix,
because the survival signal in PDAC genomics is too weak relative to the EBNM threshold.

---

## Synthetic Validation

```{r synthetic-table}
# Synthetic results from LB merged and YFB merged full runs
syn_data <- data.frame(
  Model = c("LB (point_normal)", "LB (normal)", "YFB (point_normal)", "YFB (normal)"),
  K_eff = c(2, 2, 4, 4),
  `C-index` = c(0.142, 0.142, 0.881, 0.906),
  Note = c("anti-concordant; no Phase C applied",
           "anti-concordant; no Phase C applied",
           "Phase C flip applied (training C < 0.5)",
           "Phase C flip applied (full 300-iter run)"),
  check.names = FALSE,
  stringsAsFactors = FALSE
)
knitr::kable(syn_data,
  caption = "Synthetic validation (n=250, p=1000, K=5, seed=222). Held-out C-index on 20\\% test set.",
  booktabs = TRUE)
```

**Notes.** The LB model is anti-concordant on the synthetic DGP (seed=222). Inspection of the
EBeta values (max $|\\hat{\\beta}| = 2.25$) confirms that factors are active but aligned
opposite to the true prognostic direction. The LB model does not include a Phase C sign
correction (only YFB does). Adding an analogous training-concordance check to the LB model
would be a natural next step.

---

## Alpha CV Selection (LB Only)

Alpha CV was run for the LB model; the YFB model uses fixed $\alpha = 0.5$.

```{r alpha-cv}
alpha_table <- data.frame(
  `Training mode` = c("tcga_only (n=144)", "cptac_only (n=129)", "merged (n=273)"),
  `Alpha selected` = c("0.50", "0.50", "0.30"),
  Rule = c("1SE", "1SE", "1SE"),
  check.names = FALSE, stringsAsFactors = FALSE
)
knitr::kable(alpha_table,
  caption = "Alpha CV selection for LB model (5-fold, 1SE rule). YFB uses fixed alpha=0.50.",
  booktabs = TRUE)
```

---

## Training Convergence (LB Model)

```{r lb-convergence}
lb_conv <- data.frame(
  `Training mode` = c("tcga\\_only", "cptac\\_only", "merged"),
  n   = c(144, 129, 273),
  p   = c(2000, 2000, 2000),
  K   = c(10, 10, 20),
  K_eff = c(2, 3, 3),
  `N iter` = c(41, 58, 83),
  `Final ELBO` = c("-935,867.6", "-797,192.7 (proxy)", "-"),
  `Alpha` = c("0.50", "0.50", "0.30"),
  check.names = FALSE, stringsAsFactors = FALSE
)
knitr::kable(lb_conv,
  caption = "LB training diagnostics. K\\_eff = number of factors with |beta| > 0.001.
    Final ELBO shown where computable.",
  booktabs = TRUE)
```

---

## External Validation — LB Model

```{r lb-external}
ext_cohorts <- c("Dijk", "Moffitt_GEO_array", "PACA_AU_array", "PACA_AU_seq", "Puleo_array")

get_lb_ext <- function(df, mode, prior) {
  if (is.null(df)) return(rep(NA, length(ext_cohorts)))
  sub <- df[df$section == "3_external" & df$prior_beta == prior, ]
  sapply(ext_cohorts, function(c) {
    r <- sub[sub$cohort == c, "c_index"]
    if (length(r) > 0) round(r[1], 3) else NA
  })
}

lb_ext <- data.frame(
  Cohort = c("Dijk (n=90)", "Moffitt (n=123)", "PACA-AU-array (n=63)",
             "PACA-AU-seq (n=52)", "Puleo (n=288)"),
  `tcga_only` = get_lb_ext(lb_tcga, "tcga_only", "point_normal"),
  `cptac_only` = get_lb_ext(lb_cptac, "cptac_only", "point_normal"),
  `merged` = get_lb_ext(lb_merge, "merged", "point_normal"),
  check.names = FALSE, stringsAsFactors = FALSE
)
lb_ext$`tcga_only`[is.na(lb_ext$`tcga_only`)] <- "—"
lb_ext$`cptac_only`[is.na(lb_ext$`cptac_only`)] <- "—"
lb_ext$`merged`[is.na(lb_ext$`merged`)] <- "—"

knitr::kable(lb_ext,
  caption = "LB model: external validation C-index by training mode. point\\_normal and
    normal priors give identical results for LB (the L update dominates; beta prior
    has negligible effect at the scale of active betas 0.003--0.02).",
  booktabs = TRUE)
```

---

## External Validation — YFB Model (Normal Prior)

```{r yfb-external}
get_yfb_ext <- function(df, prior) {
  if (is.null(df)) return(rep(NA_character_, length(ext_cohorts)))
  sub <- df[df$section == "3_external" & df$prior_beta == prior, ]
  sapply(ext_cohorts, function(c) {
    r <- sub[sub$cohort == c, "c_index"]
    if (length(r) > 0) as.character(round(r[1], 3)) else "—"
  })
}

yfb_ext <- data.frame(
  Cohort = c("Dijk (n=90)", "Moffitt (n=123)", "PACA-AU-array (n=63)",
             "PACA-AU-seq (n=52)", "Puleo (n=288)"),
  `tcga_only` = get_yfb_ext(yfb_tcga, "normal"),
  `cptac_only` = get_yfb_ext(yfb_cptac, "normal"),
  `merged` = get_yfb_ext(yfb_merge, "normal"),
  check.names = FALSE, stringsAsFactors = FALSE
)

knitr::kable(yfb_ext,
  caption = "YFB model (normal prior): external validation C-index by training mode.
    Phase C sign correction applied where training C < 0.5. Merged column shows
    NA if the full merged benchmark is still running.",
  booktabs = TRUE)
```

---

## LB vs. YFB Comparison (Normal Prior)

```{r lb-yfb-comparison}
lb_vals  <- get_lb_ext(lb_tcga,  "tcga_only", "point_normal")
yfb_vals <- get_yfb_ext(yfb_tcga, "normal")

compare_df <- data.frame(
  Cohort = c("Dijk", "Moffitt", "PACA-AU-array", "PACA-AU-seq", "Puleo"),
  `LB tcga_only` = as.numeric(lb_vals),
  `YFB tcga_only (normal)` = as.numeric(yfb_vals),
  Delta = round(as.numeric(yfb_vals) - as.numeric(lb_vals), 3),
  check.names = FALSE, stringsAsFactors = FALSE
)
compare_df$`Favours` <- ifelse(compare_df$Delta > 0, "YFB", ifelse(compare_df$Delta < 0, "LB", "—"))

knitr::kable(compare_df,
  caption = "Direct LB vs. YFB comparison on tcga\\_only training. Delta = YFB C − LB C.
    Positive Delta = YFB wins. YFB (normal prior) outperforms LB on 4/5 cohorts.",
  booktabs = TRUE)
```

---

## Comparison Against 04_29_26 Archived Baseline

```{r archived-comparison}
arch_data <- data.frame(
  `Training mode` = c("tcga_only", "cptac_only", "merged"),
  `Archived median C (v1)` = c(0.602, 0.628, 0.50),
  `LB median C (v2)` = c(
    round(median(as.numeric(get_lb_ext(lb_tcga, "tcga_only", "point_normal")), na.rm=TRUE), 3),
    round(median(as.numeric(get_lb_ext(lb_cptac, "cptac_only", "point_normal")), na.rm=TRUE), 3),
    round(median(as.numeric(get_lb_ext(lb_merge, "merged", "point_normal")), na.rm=TRUE), 3)
  ),
  `YFB median C (v2, normal)` = c(
    round(median(as.numeric(get_yfb_ext(yfb_tcga, "normal")), na.rm=TRUE), 3),
    round(median(as.numeric(get_yfb_ext(yfb_cptac, "normal")), na.rm=TRUE), 3),
    round(median(as.numeric(get_yfb_ext(yfb_merge, "normal")), na.rm=TRUE), 3)
  ),
  `Preprocessing` = c("v1: 407–644 shared genes", "v1: per-cohort", "v1: per-cohort"),
  check.names = FALSE, stringsAsFactors = FALSE
)
knitr::kable(arch_data,
  caption = "Comparison against the 2026-04-29 archived baseline. The archived median C
    was computed with v1 preprocessing (per-cohort top-2000 filter, 407--644 common genes).
    Current v2 uses training-set gene list (1,601--1,919 common genes). Values are
    not directly comparable; see preprocessing note in Section 2.",
  booktabs = TRUE)
```

**Interpretation.** The apparent drop in LB median C (v1: 0.60 → v2: 0.42) is primarily
a preprocessing artifact: the archived v1 result restricted prediction to 407–644 genes
that were highly variable in *both* training *and* each external cohort, creating a more
targeted (but potentially leaky) gene set. The v2 pipeline uses the training-set gene list
exclusively (no information from external cohorts at feature-selection time), giving a
more principled but lower C-index.

The YFB model under v2 preprocessing achieves median C $\approx 0.55$–$0.56$ on
single-cohort training, substantially outperforming LB (median C $\approx 0.39$–$0.42$)
under the same v2 pipeline.

---

## Key Findings and Open Questions

### What changed from the 04_29_26 report

1. **CAVI loop order fixed (LB).** The $\beta \to L \to F$ Jacobi regression introduced
   in Cluster A commit 55500fd is reverted to $L \to F \to \beta$ (Gauss-Seidel).
   Training ELBO and convergence now match the archived baseline exactly.

2. **YFB scale issue fixed (Phase B).** EF column normalization prevents
   $A_{\beta} = O(pn) \Rightarrow \hat{\beta} \to 0$.

3. **YFB sign inversion fixed (Phase C).** Training concordance check after convergence
   corrects the direction of the risk score.

4. **Preprocessing changed (v1 → v2).** The v2 pipeline (intersect-first + QN +
   rank-transform) is more principled but gives lower external C-index due to using
   more noisy genes from the training gene set at prediction time.

### Open questions

1. **LB sign inversion on synthetic.** The LB model is anti-concordant on the seed=222
   synthetic DGP (C=0.142). Adding an analogous Phase C check to LB would be a
   low-effort next step.

2. **YFB merged PDAC.** Both priors give K\_eff=0 or near-zero betas on merged TCGA+CPTAC.
   The pure-genomics $F$ update ($\alpha_F=0$) captures platform variation (RNA-seq vs.
   proteomics) rather than biology; ZF scores are uncorrelated with survival.
   Enabling $\alpha_F > 0$ with proper stabilization may help.

3. **Archived baseline reproducibility.** The archived C=0.60–0.65 used v1 preprocessing.
   A proper apples-to-apples comparison requires rerunning the archived configuration
   (v1 gene sets) under the current codebase.

4. **Phase D (normalize\_AB).** Not yet evaluated; deferred until the above issues
   are resolved. Likely to help only if the A\_surv/A\_gen imbalance is the binding
   constraint (it may not be with v2 preprocessing).

---

## Software and Reproducibility

```{r session}
cat(sprintf("R version: %s\n", R.version$version.string))
cat(sprintf("Date: %s\n", Sys.Date()))
cat(sprintf("LB results: %s\n", OUT_LB))
cat(sprintf("YFB results: %s\n", OUT_YFB))
cat("Key commits:\n")
cat("  5bfc25e — Revert CAVI inner loop to L->F->beta\n")
cat("  4ea414f — YFB Phase B+C: EF normalization + sign correction\n")
```

All results are reproducible from the project root:

```bash
# LB benchmarks
Rscript results/benchmark_sim/run_LB_benchmark.R --train-mode tcga_only
Rscript results/benchmark_sim/run_LB_benchmark.R --train-mode cptac_only
Rscript results/benchmark_sim/run_LB_benchmark.R --train-mode merged

# YFB benchmarks
Rscript results/benchmark_sim/run_YFB_benchmark.R --train-mode merged  # includes synthetic
Rscript results/benchmark_sim/run_YFB_benchmark.R --train-mode tcga_only
Rscript results/benchmark_sim/run_YFB_benchmark.R --train-mode cptac_only

# Unit tests
Rscript tests/run_tests.R  # 171/171 expected
```
