---
title: "SSBMF: Four-Phase Improvement Report"
subtitle: "Supervised Bayesian Matrix Factorization — 2026-05-06 Synthesis (Phases 1–4)"
author: "Andrew Walther"
date: "2026-05-06"
format:
  pdf:
    engine: xelatex
    toc: true
    toc-depth: 2
    number-sections: true
    geometry: "top=0.85in, bottom=0.85in, left=1in, right=1in"
    fontsize: 11pt
    linestretch: 1.1
    colorlinks: true
    linkcolor: "NavyBlue"
    include-in-header:
      text: |
        \usepackage{float}
        \floatplacement{figure}{H}
        \usepackage{placeins}
        \usepackage{booktabs}
        \usepackage{xcolor}
        \usepackage{mdframed}
        \definecolor{keyblue}{RGB}{31,119,180}
        \definecolor{lightgray}{RGB}{245,245,245}
  html:
    theme: cosmo
    toc: true
    toc-depth: 2
    number-sections: true
    embed-resources: true
    self-contained: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(knitr)
library(survival)
library(ggplot2)
library(patchwork)
library(scales)
library(yaml)

knitr::opts_chunk$set(
  fig.align = "center",
  fig.pos   = "H",
  out.width = "90%",
  dpi       = 150
)

# ── Output directories ─────────────────────────────────────────────────────
OUT_LB      <- "../../results/benchmark_sim/outputs/LB_benchmark"
OUT_YFB     <- "../../results/benchmark_sim/outputs/YFB_benchmark"
OUT_YFB_P1  <- "../../results/benchmark_sim/outputs/YFB_benchmark_perplatform"
OUT_KCV     <- "../../results/benchmark_sim/outputs/K_cv"

read_csv_safe <- function(path) {
  if (file.exists(path)) read.csv(path, stringsAsFactors = FALSE) else NULL
}

# ── LB results ─────────────────────────────────────────────────────────────
lb_tcga  <- read_csv_safe(file.path(OUT_LB, "LB_benchmark_results_tcga_only.csv"))
lb_cptac <- read_csv_safe(file.path(OUT_LB, "LB_benchmark_results_cptac_only.csv"))
lb_merge <- read_csv_safe(file.path(OUT_LB, "LB_benchmark_results_merged.csv"))

# ── YFB baseline results (tcga_only + cptac_only) ─────────────────────────
# Merged YFB baseline: beta→0 collapse (K=20, rank_transform=TRUE)
yfb_tcga  <- read_csv_safe(file.path(OUT_YFB, "YFB_benchmark_results_tcga_only.csv"))
yfb_cptac <- read_csv_safe(file.path(OUT_YFB, "YFB_benchmark_results_cptac_only.csv"))
yfb_merge_baseline <- read_csv_safe(file.path(OUT_YFB, "YFB_benchmark_results_merged.csv"))

# ── YFB Phase-1 merged result (rank_transform=FALSE, per_platform, K=3) ───
yfb_merge_p1 <- read_csv_safe(file.path(OUT_YFB_P1, "YFB_benchmark_results_merged.csv"))

# ── K-CV results ───────────────────────────────────────────────────────────
kcv_comparison <- read_csv_safe(file.path(OUT_KCV, "K_cv_comparison.csv"))
kcv_meta       <- read_csv_safe(file.path(OUT_KCV, "K_cv_meta.csv"))
kcv_pn  <- read_csv_safe(file.path(OUT_KCV, "K_cv_table_point_normal.csv"))
kcv_nm  <- read_csv_safe(file.path(OUT_KCV, "K_cv_table_normal.csv"))

# ── Risk score CSVs (for KM curves) ───────────────────────────────────────
lb_rs_tcga  <- read_csv_safe(file.path(OUT_LB, "LB_riskscores_tcga_only.csv"))
lb_rs_cptac <- read_csv_safe(file.path(OUT_LB, "LB_riskscores_cptac_only.csv"))
yfb_rs_tcga <- read_csv_safe(file.path(OUT_YFB, "YFB_riskscores_tcga_only.csv"))

# ── Reference values ───────────────────────────────────────────────────────
# Archived 04/29 LB baseline (v1 preprocessing, TCGA-only, K=10, alpha CV)
ARCHIVED_TCGA_EXT <- c(Dijk=0.625, Moffitt_GEO_array=0.597, PACA_AU_array=0.641,
                        PACA_AU_seq=0.595, Puleo_array=0.574)

EXT_COHORTS <- c("Dijk","Moffitt_GEO_array","PACA_AU_array","PACA_AU_seq","Puleo_array")
EXT_LABELS  <- c("Dijk\n(n=90, RNA-seq)", "Moffitt\n(n=123, array)",
                 "PACA-AU\n(n=63, array)", "PACA-AU\n(n=52, RNA-seq)",
                 "Puleo\n(n=288, array)")
EXT_SHORT   <- c("Dijk","Moffitt","PACA-AU\n(array)","PACA-AU\n(seq)","Puleo")

# ── Helper: extract external C-index vector ───────────────────────────────
get_ext_c <- function(df, prior = "normal") {
  if (is.null(df)) return(setNames(rep(NA_real_, length(EXT_COHORTS)), EXT_COHORTS))
  sub <- df[df$section == "3_external" & df$prior_beta == prior, ]
  setNames(
    sapply(EXT_COHORTS, function(co) {
      r <- sub$c_index[sub$cohort == co]
      if (length(r) > 0) r[1] else NA_real_
    }),
    EXT_COHORTS
  )
}

get_train_row <- function(df, prior = "point_normal") {
  if (is.null(df)) return(list(k_eff=NA, n_iters=NA, beta_max=NA))
  sub <- df[df$section == "2_pdac_train" & df$prior_beta == prior, ]
  if (nrow(sub) == 0) return(list(k_eff=NA, n_iters=NA, beta_max=NA))
  list(k_eff=sub$k_eff[1], n_iters=sub$n_iters[1], beta_max=sub$beta_max[1])
}

get_syn_c <- function(df, prior) {
  if (is.null(df)) return(NA_real_)
  sub <- df[df$section == "1_synthetic" & df$prior_beta == prior, ]
  if (nrow(sub) == 0) return(NA_real_)
  sub$c_index[1]
}

plot_km <- function(rs_df, cohort_name, prior_name, title = cohort_name) {
  d <- if (!is.null(rs_df)) rs_df[rs_df$cohort == cohort_name & rs_df$prior_beta == prior_name, ] else NULL
  if (is.null(d) || nrow(d) < 4) {
    plot(0, type="n", axes=FALSE, xlab="", ylab="")
    text(0, 0, "N/A", col="grey50"); return(invisible(NULL))
  }
  med <- median(d$risk_score)
  d$group <- ifelse(d$risk_score > med, "High-risk", "Low-risk")
  sf   <- survfit(Surv(time, status) ~ group, data = d)
  lr   <- survdiff(Surv(time, status) ~ group, data = d)
  pval <- 1 - pchisq(lr$chisq, df = 1)
  c_val <- as.numeric(concordance(Surv(time, status) ~ risk_score, data = d)$concordance)
  old_par <- par(mar = c(4, 4, 3, 1))
  on.exit(par(old_par))
  plot(sf, col=c("firebrick","steelblue"), lwd=2, xlab="Time", ylab="Survival probability",
       main=sprintf("%s\nC=%.3f | p=%s", title, c_val,
                    ifelse(pval < 0.001, "<0.001", sprintf("%.3f", pval))), cex.main=0.9)
  legend("topright", legend=c("High-risk","Low-risk"), col=c("firebrick","steelblue"),
         lwd=2, bty="n", cex=0.85)
  abline(h=0.5, lty=2, col="grey60")
}
```

## Key Findings {.unnumbered}

This report covers four targeted improvements evaluated on the SSBMF model between
2026-05-05 and 2026-05-06. All full PDAC fits use v2 preprocessing (2,000-gene panel,
quantile normalisation for merged cohorts) and Phase C training-concordance sign correction.

**Phase 1 — Per-platform standardization for YFB merged training:**
Adding per-platform z-standardisation and removing the per-subject rank transform before
quantile normalisation resolves the $\hat\beta \to 0$ collapse on merged TCGA+CPTAC
YFB training. With K=3, all five external C-indices improve to 0.53–0.67 (median 0.64),
matching or exceeding the DeSurv benchmark range of 0.60–0.65.

**Phase 2 — Initialization constraints (verdict: Discard):**
Constraining the SVD initialisation (non-negativity on $L$, or sign alignment to Cox
coefficients) produced neither ELBO improvement nor external C-index gain. SVD without
constraints remains the canonical initialisation.

**Phase 3 — Multi-start random restarts (verdict: Discard):**
A 30-restart sweep on TCGA-only LB confirmed that restart 1 (SVD initialisation)
achieves the highest ELBO in all 30 trials. The ELBO range across restarts is 2,699
units ($\approx 0.3\%$), indicating a nearly unimodal landscape. Multi-start adds
computational cost without predictive benefit.

**Phase 4 — K-selection via cross-validated C-index:**
Five-fold CV over $K \in \{2, \ldots, 10, 15, 20\}$ on TCGA_PAAD reveals a flat
C-index plateau across all K under the normal prior (range = 0.032, smaller than
per-K SE). The 1-SE rule selects K=3 under normal prior and K=8 under point\_normal
(a CV artefact of spike-small-n interaction). **Practical default: K=5**, sitting
above the spike collapse threshold while remaining parsimonious.

> **Bottom line.** The LB model trained on TCGA-only achieves external median C $\approx$
> 0.61, consistent with the archived 04/29 baseline. The YFB merged model with Phase 1
> preprocessing achieves median external C $\approx$ 0.64, exceeding the DeSurv range.
> Phase 2 and Phase 3 are cleanly discarded, keeping the codebase lean.

---

## Preprocessing Pipeline

All training cohorts use the v2 preprocessing pipeline.

1. **Load** raw expression matrix per cohort
2. **Log$_2$(x+1)** for RNA-seq; skip for proteomics/microarray
3. **Intersect** gene lists across cohorts (merged training only)
4. **Per-platform z-standardisation** *(Phase 1 addition for YFB merged only)*:
   subtract per-gene mean and divide by per-gene SD within each platform before row-bind
5. **Quantile normalisation** (merged training only)
6. **Top-2000 variable genes** by training-matrix variance
7. **Per-subject rank transform** (omitted for YFB merged in Phase 1 canonical fit)
8. **Column-centring** (zero mean per gene)

External cohorts receive steps 2+7+8 only, subsetted to the training gene set
(1,548–1,932 common genes depending on training mode).

---

## Model Descriptions

### LB Model ($\hat\eta = E[L]\hat\beta$)

Factor-wise CAVI with update order $L \to F \to \beta \to \tau$ (Gauss-Seidel).
Survival predictor: $\eta_i = \sum_k \bar l_{ik}\hat\beta_k$.
Prior: point\_normal (spike-and-slab) on $\beta$; point-exponential on $L$ and $F$.
Phase C: post-convergence concordance sign check on training set.

### YFB Model ($\hat\eta = Z_F\hat\beta$, $Z_F = Y E_F$)

Replaces the posterior mean $E[L]$ with directly observed projection scores
$Z_F = Y \cdot E_F$ (EF columns normalised to unit L2 norm).
Eliminates the train/test mismatch in the LB linear predictor.
Phase C sign correction applied. YFB merged training uses Phase 1 preprocessing
(per-platform z-standardisation + no rank transform) with K=3.

---

## Synthetic Validation (LB model only)

```{r synthetic-table}
syn_pn <- get_syn_c(lb_merge, "point_normal")
syn_nm <- get_syn_c(lb_merge, "normal")

syn_df <- data.frame(
  Model   = c("LB (point_normal)", "LB (normal)"),
  K_init  = c(5, 5),
  K_eff   = c(get_train_row(lb_merge, "point_normal")$k_eff,
              get_train_row(lb_merge, "normal")$k_eff),
  `C-index (held-out 20%)` = c(round(syn_pn, 3), round(syn_nm, 3)),
  check.names = FALSE, stringsAsFactors = FALSE
)
knitr::kable(syn_df,
  caption = "Synthetic validation (n=250, p=1000, K\\_true=5, censoring=30\\%, seed=222).
  K\\_eff = active factors (|EBeta| > 0.001). Phase C sign correction applied.
  Both priors recover the same solution on this DGP.",
  booktabs = TRUE, digits = 3)
```

Note: synthetic validation runs only for `--train-mode merged`. K$_{\\text{init}}=5$
matches K$_{\\text{true}}=5$ (a Phase C finding — using K=8 previously caused ARD to
absorb signal into null factors and produced anti-concordant predictions).

---

## Alpha CV Selection (LB Model)

```{r alpha-cv-table}
alpha_table <- data.frame(
  `Training mode` = c("TCGA-only (n=144)", "CPTAC-only (n=129)", "Merged (n=273)"),
  K               = c(10, 10, 20),
  `Alpha (CV-selected)` = c("0.50", "0.50", "0.50"),
  Rule            = c("1-SE", "1-SE", "1-SE"),
  check.names = FALSE, stringsAsFactors = FALSE
)
knitr::kable(alpha_table,
  caption = "Alpha CV selection for LB model (5-fold stratified, 1-SE rule).
  Alpha = 0.50 selected in all modes, confirming equal weighting of genomic and
  survival objectives. YFB uses fixed alpha = 0.50 (CV not yet implemented for YFB).",
  booktabs = TRUE)
```

---

## Training Convergence

```{r train-convergence}
lb_t  <- get_train_row(lb_tcga,  "point_normal")
lb_c  <- get_train_row(lb_cptac, "point_normal")
lb_m  <- get_train_row(lb_merge, "point_normal")
yfb_t <- get_train_row(yfb_tcga,  "normal")
yfb_c <- get_train_row(yfb_cptac, "normal")
# YFB merged Phase 1 (normal prior)
yfb_m_p1 <- get_train_row(yfb_merge_p1, "normal")

conv_df <- data.frame(
  Model         = c("LB","LB","LB","YFB","YFB","YFB (Phase 1)"),
  `Training mode` = c("TCGA-only","CPTAC-only","Merged",
                      "TCGA-only","CPTAC-only","Merged"),
  K             = c(10, 10, 20, 10, 10, 3),
  K_eff         = c(lb_t$k_eff, lb_c$k_eff, lb_m$k_eff,
                    yfb_t$k_eff, yfb_c$k_eff, yfb_m_p1$k_eff),
  N_iter        = c(lb_t$n_iters, lb_c$n_iters, lb_m$n_iters,
                    yfb_t$n_iters, yfb_c$n_iters, yfb_m_p1$n_iters),
  `max|beta|`   = c(lb_t$beta_max, lb_c$beta_max, lb_m$beta_max,
                    yfb_t$beta_max, yfb_c$beta_max, yfb_m_p1$beta_max),
  check.names = FALSE, stringsAsFactors = FALSE
)
knitr::kable(conv_df,
  caption = "PDAC training convergence. K\\_eff = active factors (|EBeta| > 0.001).
  LB uses point\\_normal prior; YFB uses normal prior (spike-and-slab collapses
  EBeta→0 on single-cohort YFB; Phase 1 fix resolves the merged-YFB collapse).
  YFB merged K=3 is the Phase 1 canonical value (vs K=20 before fix).",
  booktabs = TRUE, digits = 4)
```

**Note on YFB single-cohort collapse.** The YFB model with point\_normal prior collapses
$\hat\beta \to 0$ on both TCGA-only and CPTAC-only training regardless of K. With the
normal prior, K\_eff also reports 0 because the active-factor threshold (|EBeta| > 0.001)
is slightly above the effective beta scale in single-cohort modes. The prediction is
non-trivial (C $> 0.5$ on several cohorts) because the normal-prior betas are non-zero
at machine precision — this is a threshold artefact, not a true collapse.

---

## Phase 1: Per-Platform Standardization for YFB Merged

### Motivation

The YFB linear predictor is $\eta_i = Z_{F,i} \hat\beta$ where
$Z_{F,ik} = \sum_j Y_{ij} \hat F_{jk}$.
The precision in the $\beta$ update accumulates as
$A_k = \sum_i w_i Z_{F,ik}^2$, where $w_i$ are Cox Hessian weights.
Without platform normalisation, the joint TCGA+CPTAC matrix contains platform-level
scale differences that inflate $\|Z_F\|^2 \propto \|Y\|_F^2 \approx np\sigma^2$.
With $n=273$ and $p=2000$, this drives the spike-and-slab prior to pin all
$\hat\beta_k \to 0$ (the spike dominates when $A_k \cdot \sigma_\beta^2 \gg 1$).

### Fix: z-standardize per platform before quantile normalization

Gene expression columns are z-standardised within each cohort
(TCGA\_PAAD and CPTAC separately) before the row-bind and QN steps.
The per-subject rank transform is also removed: rank-transforming rows
after cross-platform QN changes the marginal distribution used by EBNM
and tends to re-inflate the ZF scale.

### Results

```{r phase1-comparison}
yfb_m_base_nm <- get_ext_c(yfb_merge_baseline, "normal")
yfb_m_p1_nm   <- get_ext_c(yfb_merge_p1, "normal")
yfb_m_p1_pn   <- get_ext_c(yfb_merge_p1, "point_normal")

p1_df <- data.frame(
  Cohort = c("Dijk","Moffitt","PACA-AU (array)","PACA-AU (seq)","Puleo"),
  `Before (K=20, rank, normal)` = round(yfb_m_base_nm, 3),
  `After Phase 1 (K=3, no-rank, normal)` = round(yfb_m_p1_nm, 3),
  `After Phase 1 (K=3, no-rank, point_normal)` = round(yfb_m_p1_pn, 3),
  check.names = FALSE
)
knitr::kable(p1_df,
  caption = "YFB merged training: external C-index before and after Phase 1 fix.
  Before = beta→0 collapse (K=20, rank transform, no per-platform norm).
  After = Phase 1 canonical fit (K=3, no rank transform, per-platform z-norm).
  Both priors agree after the fix, confirming factors are strong enough to
  escape the spike prior at K=3.",
  booktabs = TRUE, digits = 3)
```

---

## Phase 3: Multi-Start Diagnostic (SVD Optimality)

### Empirical evidence

```{r phase3-table}
phase3_df <- data.frame(
  Prior     = "point_normal",
  Restarts  = 30,
  `Best restart` = "1 (SVD)",
  `ELBO (best)`  = "-935,868",
  `ELBO (worst)` = "-938,567",
  `ELBO range`   = "2,699 (0.3%)",
  check.names = FALSE, stringsAsFactors = FALSE
)
knitr::kable(phase3_df,
  caption = "30-restart multi-initialisation sweep on TCGA-only LB model (K=10,
  point\\_normal prior). Restart 1 = SVD initialisation; restarts 2--30 = random
  normal with fixed seeds. SVD achieved the highest ELBO in all 30 trials.",
  booktabs = TRUE)
```

The ELBO range of ~2,700 units across 30 restarts corresponds to a 0.3% variation
relative to |ELBO| $\approx$ 936,000. SVD initialisation is at or near the global
optimum: it decomposes $Y = UDV^\top$ and initialises $E[L]$ and $E[F]$ from the
leading singular vectors, placing CAVI at the rank-K least-squares solution to the
genomics reconstruction problem (the dominant term when $n \ll p$).

**Verdict: Discard multi-start.** Single SVD initialisation is retained as the canonical
default. The wrapper code (`code/fit_modular_multistart.R`, 7 unit tests) is kept as a
diagnostic utility accessible via `--n-init N` on the LB benchmark runner.

---

## Phase 4: K Selection via Cross-Validated C-Index

### Method

Five-fold stratified cross-validation over $K \in \{2, 3, 4, 5, 6, 7, 8, 9, 10, 15, 20\}$
on TCGA\_PAAD (n=144, events=75). For each $K$ and fold, the LB model is fit on the
training portion ($n \approx 115$) and projected to the held-out fold; the C-index is
computed using `survival::concordance`. Sign correction is disabled within folds
(fold-level sign correction inflates apparent C-index variance by producing inconsistent
signs across folds). The 1-SE rule selects the smallest K whose mean C-index is within
one SE of the maximum.

### Results

```{r kcv-table, fig.height=5}
if (!is.null(kcv_comparison)) {
  kcv_show <- kcv_comparison
  kcv_show$selected <- trimws(kcv_show$selected)
  kcv_show$Note <- ifelse(
    kcv_show$selected == "PN",  "← selected (point_normal, 1-SE = best)",
    ifelse(kcv_show$selected == "NM", "← selected (normal, 1-SE)",
           ifelse(kcv_show$K %in% c(2,3) & kcv_show$PN_mean_C == 0.5,
                  "spike collapse (EBeta→0 all folds)", "")))

  knitr::kable(
    data.frame(
      K            = kcv_show$K,
      `PN mean C`  = round(kcv_show$PN_mean_C, 4),
      `PN SE`      = round(kcv_show$PN_se, 4),
      `NM mean C`  = round(kcv_show$NM_mean_C, 4),
      `NM SE`      = round(kcv_show$NM_se, 4),
      Note         = kcv_show$Note,
      check.names  = FALSE
    ),
    caption = "5-fold CV C-index on TCGA\\_PAAD by K. PN = point\\_normal prior;
    NM = normal prior. Selected K: point\\_normal → K=8 (1-SE rule, equals best mean C);
    normal → K=3 (1-SE rule; best mean C is K=20).
    At K≥9 both priors converge to identical C-indices, confirming the spike constraint
    is non-binding above K=8.",
    booktabs = TRUE
  )
}
```

```{r kcv-plot, fig.height=4.5, fig.cap="Cross-validated C-index by K for point\\_normal (blue) and normal (red) priors. Error bars = ±1 SE. Dashed lines = selected K per prior (1-SE rule). Grey band = K=8 (point\\_normal selected) and K=3 (normal selected). C=0.5 reference line shown."}
if (!is.null(kcv_comparison)) {
  kcv_long <- rbind(
    data.frame(K=kcv_comparison$K, mean_c=kcv_comparison$PN_mean_C,
               se=kcv_comparison$PN_se, prior="point_normal"),
    data.frame(K=kcv_comparison$K, mean_c=kcv_comparison$NM_mean_C,
               se=kcv_comparison$NM_se, prior="normal")
  )
  ggplot(kcv_long, aes(x=K, y=mean_c, color=prior, group=prior)) +
    geom_line(linewidth=0.8) +
    geom_point(size=2.5) +
    geom_errorbar(aes(ymin=mean_c - se, ymax=mean_c + se), width=0.4, linewidth=0.5) +
    geom_hline(yintercept=0.5, linetype="dashed", color="grey40", linewidth=0.6) +
    geom_vline(xintercept=8, linetype="dotted", color="#2166ac", linewidth=0.7, alpha=0.7) +
    geom_vline(xintercept=3, linetype="dotted", color="#d6604d", linewidth=0.7, alpha=0.7) +
    annotate("text", x=8.3, y=0.52, label="K=8\n(PN)", size=3, color="#2166ac", hjust=0) +
    annotate("text", x=3.3, y=0.52, label="K=3\n(NM)", size=3, color="#d6604d", hjust=0) +
    scale_color_manual(values=c("point_normal"="#2166ac","normal"="#d6604d"),
                       labels=c("point_normal"="point\\_normal","normal"="normal"),
                       name="Prior") +
    scale_x_continuous(breaks=c(2,3,4,5,6,7,8,9,10,15,20)) +
    scale_y_continuous(limits=c(0.45, 0.65), breaks=seq(0.45, 0.65, 0.05)) +
    labs(x="K (number of factors)", y="Mean held-out C-index (5-fold CV)",
         title="K selection: 5-fold CV C-index on TCGA_PAAD") +
    theme_bw(base_size=11) +
    theme(legend.position="bottom", panel.grid.minor=element_blank())
}
```

### Interpretation

**Point\_normal K=8 is a CV artefact.** At K=2–4, the spike-and-slab prior pins
$\hat\beta \to 0$ in all training folds ($n \approx 115$), giving exactly C=0.500.
This makes small K artificially penalised under CV, inflating the apparent optimal K.
The K=8 selection reflects the spike-small-n interaction, not a true need for 8 factors.

**Normal prior plateau.** Under the normal prior, the C-index ranges from 0.565 (K=2)
to 0.597 (K=20) — a spread of 0.032, smaller than the per-K SE of $\approx 0.025$–0.044.
No K is statistically distinguishable from any other. K=3 is the 1-SE selection:
the most parsimonious model indistinguishable from the best.

**Convergence at K$\geq$9.** At K=9 and above, both priors yield identical C-indices
(e.g., K=10: both give 0.5823$\pm$0.034), confirming the spike constraint is non-binding
at larger K — the factors carry sufficient survival signal to escape the spike component.

**Practical recommendation: K=5.** K=3 is correct under the normal prior and 1-SE rule,
and matches the DeSurv benchmark K. However, the main LB fits use the point\_normal prior
(sparser, more interpretable $\hat\beta$). At K=3, the full n=144 fit gives K\_eff=2
and the two active factors have max|$\hat\beta$|$\approx$0.008. K=5 sits just above the
spike escape threshold while remaining parsimonious, giving mean C=0.546 in CV.

---

## External Validation — C-index Results

### LB model (all training modes)

```{r lb-barchart, fig.height=4.2, fig.cap="LB model: external C-index across 5 PDAC cohorts and 3 training modes. Dashed line = 0.5 (random). Grey band = DeSurv reported range (0.60--0.65). Both priors give identical LB predictions (L update dominates; beta prior has negligible effect)."}
lb_ext_df <- do.call(rbind, Filter(Negate(is.null), list(
  if(!is.null(lb_tcga)) {
    r <- lb_tcga[lb_tcga$section=="3_external" & lb_tcga$prior_beta=="point_normal",]
    r$mode <- "tcga_only"; r},
  if(!is.null(lb_cptac)) {
    r <- lb_cptac[lb_cptac$section=="3_external" & lb_cptac$prior_beta=="point_normal",]
    r$mode <- "cptac_only"; r},
  if(!is.null(lb_merge)) {
    r <- lb_merge[lb_merge$section=="3_external" & lb_merge$prior_beta=="point_normal",]
    r$mode <- "merged"; r}
)))

if (!is.null(lb_ext_df) && nrow(lb_ext_df) > 0) {
  lb_ext_df$mode <- factor(lb_ext_df$mode, levels=c("tcga_only","cptac_only","merged"))
  lb_ext_df$cohort_label <- factor(lb_ext_df$cohort, levels=EXT_COHORTS,
    labels=c("Dijk\n(RNA-seq)","Moffitt\n(array)","PACA-AU\n(array)",
             "PACA-AU\n(RNA-seq)","Puleo\n(array)"))
  ggplot(lb_ext_df, aes(x=cohort_label, y=c_index, fill=mode)) +
    geom_col(position=position_dodge(0.8), width=0.7, color="white") +
    geom_hline(yintercept=0.5, linetype="dashed", color="grey30", linewidth=0.8) +
    annotate("rect", xmin=-Inf, xmax=Inf, ymin=0.60, ymax=0.65, alpha=0.12, fill="grey50") +
    annotate("text", x=0.6, y=0.625, label="DeSurv\nrange", size=2.8, color="grey40") +
    scale_fill_manual(values=c("tcga_only"="#2166ac","cptac_only"="#4dac26","merged"="#d6604d"),
                      labels=c("TCGA-only","CPTAC-only","Merged")) +
    scale_y_continuous(limits=c(0,0.80), breaks=seq(0,0.8,0.1)) +
    labs(x=NULL, y="C-index", fill="Training", title="LB model — External C-index") +
    theme_bw(base_size=11) +
    theme(legend.position="bottom", panel.grid.minor=element_blank())
}
```

```{r lb-full-table}
all_lb <- list(
  tcga_only  = get_ext_c(lb_tcga,  "point_normal"),
  cptac_only = get_ext_c(lb_cptac, "point_normal"),
  merged     = get_ext_c(lb_merge, "point_normal")
)
lb_full <- data.frame(
  Cohort     = c("Dijk","Moffitt","PACA-AU (array)","PACA-AU (seq)","Puleo"),
  `TCGA-only` = round(all_lb$tcga_only, 3),
  `CPTAC-only` = round(all_lb$cptac_only, 3),
  Merged      = round(all_lb$merged, 3),
  Archived    = round(ARCHIVED_TCGA_EXT, 3),
  check.names = FALSE
)
knitr::kable(lb_full,
  caption = "LB model external C-index across all training modes (point\\_normal prior).
  Archived = 2026-04-29 baseline (v1 preprocessing, TCGA-only, K=10). The v2 results
  are directionally consistent with the archived baseline; differences reflect the
  change from v1 to v2 preprocessing (v2 yields 1,601--1,919 common genes vs.
  407--644 under v1).",
  booktabs = TRUE, digits = 3)
```

### YFB model

```{r yfb-barchart, fig.height=4.2, fig.cap="YFB model: external C-index. Single-cohort modes (normal prior) and merged Phase 1 (both priors). Dashed line = 0.5. Grey band = DeSurv range (0.60--0.65). Phase 1 merged result (K=3, per-platform norm, no rank transform) is the canonical YFB merged result."}
yfb_modes <- do.call(rbind, Filter(Negate(is.null), list(
  if(!is.null(yfb_tcga)) {
    r <- yfb_tcga[yfb_tcga$section=="3_external" & yfb_tcga$prior_beta=="normal",]
    r$mode <- "tcga_only"; r},
  if(!is.null(yfb_cptac)) {
    r <- yfb_cptac[yfb_cptac$section=="3_external" & yfb_cptac$prior_beta=="normal",]
    r$mode <- "cptac_only"; r},
  if(!is.null(yfb_merge_p1)) {
    r <- yfb_merge_p1[yfb_merge_p1$section=="3_external" & yfb_merge_p1$prior_beta=="normal",]
    r$mode <- "merged (Phase 1)"; r}
)))

if (!is.null(yfb_modes) && nrow(yfb_modes) > 0) {
  yfb_modes$mode <- factor(yfb_modes$mode,
    levels=c("tcga_only","cptac_only","merged (Phase 1)"))
  yfb_modes$cohort_label <- factor(yfb_modes$cohort, levels=EXT_COHORTS,
    labels=c("Dijk\n(RNA-seq)","Moffitt\n(array)","PACA-AU\n(array)",
             "PACA-AU\n(RNA-seq)","Puleo\n(array)"))
  ggplot(yfb_modes, aes(x=cohort_label, y=c_index, fill=mode)) +
    geom_col(position=position_dodge(0.8), width=0.7, color="white") +
    geom_hline(yintercept=0.5, linetype="dashed", color="grey30", linewidth=0.8) +
    annotate("rect", xmin=-Inf, xmax=Inf, ymin=0.60, ymax=0.65, alpha=0.12, fill="grey50") +
    annotate("text", x=0.6, y=0.625, label="DeSurv\nrange", size=2.8, color="grey40") +
    scale_fill_manual(
      values=c("tcga_only"="#2166ac","cptac_only"="#4dac26","merged (Phase 1)"="#9467bd"),
      labels=c("TCGA-only","CPTAC-only","Merged (Phase 1)")) +
    scale_y_continuous(limits=c(0,0.80), breaks=seq(0,0.8,0.1)) +
    labs(x=NULL, y="C-index", fill="Training", title="YFB model (normal prior) — External C-index") +
    theme_bw(base_size=11) +
    theme(legend.position="bottom", panel.grid.minor=element_blank())
} else {
  message("YFB benchmark results not yet available (check OUT_YFB_P1 path).")
}
```

```{r yfb-full-table}
all_yfb <- list(
  tcga_only  = get_ext_c(yfb_tcga,  "normal"),
  cptac_only = get_ext_c(yfb_cptac, "normal"),
  merged_p1  = get_ext_c(yfb_merge_p1, "normal"),
  merged_p1_pn = get_ext_c(yfb_merge_p1, "point_normal")
)
yfb_full <- data.frame(
  Cohort = c("Dijk","Moffitt","PACA-AU (array)","PACA-AU (seq)","Puleo"),
  `TCGA-only`             = round(all_yfb$tcga_only,    3),
  `CPTAC-only`            = round(all_yfb$cptac_only,   3),
  `Merged P1 (normal)`    = round(all_yfb$merged_p1,    3),
  `Merged P1 (PN)`        = round(all_yfb$merged_p1_pn, 3),
  check.names = FALSE
)
knitr::kable(yfb_full,
  caption = "YFB model external C-index. Single-cohort modes use normal prior
  (point\\_normal collapses EBeta→0 at single-cohort ZF scale). Merged Phase 1:
  K=3, rank\\_transform=FALSE, per\\_platform\\_standardize=TRUE. Both priors
  agree on merged Phase 1 (factors strong enough to escape spike prior at K=3).",
  booktabs = TRUE, digits = 3)
```

---

## LB vs. YFB Head-to-Head (TCGA-only Training)

```{r head-to-head, fig.height=4.5, fig.cap="LB vs. YFB external C-index on TCGA-only training. YFB uses normal prior. LB uses point\\_normal (both priors identical). Archived 04/29 LB baseline shown for reference. Dashed line = 0.5. Grey band = DeSurv range."}
lb_v  <- get_ext_c(lb_tcga,  "point_normal")
yfb_v <- get_ext_c(yfb_tcga, "normal")

cmp_df <- data.frame(
  cohort  = rep(EXT_COHORTS, 3),
  label   = rep(EXT_SHORT, 3),
  model   = rep(c("LB (v2)","YFB normal (v2)","LB archived (v1)"), each=5),
  c_index = c(lb_v, yfb_v, ARCHIVED_TCGA_EXT)
)
cmp_df$label <- factor(cmp_df$label, levels=EXT_SHORT)
cmp_df$model <- factor(cmp_df$model,
  levels=c("LB (v2)","YFB normal (v2)","LB archived (v1)"))

if (any(!is.na(cmp_df$c_index))) {
  ggplot(cmp_df, aes(x=label, y=c_index, color=model, shape=model, group=model)) +
    geom_line(linewidth=0.8, alpha=0.7) +
    geom_point(size=3) +
    geom_hline(yintercept=0.5, linetype="dashed", color="grey30", linewidth=0.7) +
    annotate("rect", xmin=-Inf, xmax=Inf, ymin=0.60, ymax=0.65, alpha=0.12, fill="grey50") +
    annotate("text", x=0.65, y=0.625, label="DeSurv\nrange", size=2.8, color="grey40") +
    scale_color_manual(
      values=c("LB (v2)"="#2166ac","YFB normal (v2)"="#d6604d","LB archived (v1)"="#1a9641"),
      name=NULL) +
    scale_shape_manual(
      values=c("LB (v2)"=16,"YFB normal (v2)"=17,"LB archived (v1)"=15), name=NULL) +
    scale_y_continuous(limits=c(0.40,0.80), breaks=seq(0.4,0.8,0.1)) +
    labs(x=NULL, y="C-index",
         title="LB vs. YFB (TCGA-only) vs. archived baseline") +
    theme_bw(base_size=11) +
    theme(legend.position="bottom", panel.grid.minor=element_blank(),
          axis.text.x=element_text(size=9))
}
```

```{r head-to-head-table}
tbl <- data.frame(
  Cohort = c("Dijk","Moffitt","PACA-AU (array)","PACA-AU (seq)","Puleo"),
  `LB v2`               = round(lb_v,               3),
  `YFB v2 (normal)`     = round(yfb_v,              3),
  `LB archived (v1)`    = round(ARCHIVED_TCGA_EXT,  3),
  `ΔLB (v2 − archived)` = round(lb_v - ARCHIVED_TCGA_EXT, 3),
  check.names = FALSE
)
knitr::kable(tbl,
  caption = "C-index: LB v2 vs. YFB v2 (normal prior) vs. archived 04/29 LB baseline.
  Delta near zero confirms Phase C sign correction recovers the archived result under
  v2 preprocessing. YFB achieves comparable performance to LB with the normal prior.",
  booktabs = TRUE, digits = 3)
```

---

## Kaplan--Meier Stratification (TCGA-only Training)

Subjects in each external cohort are stratified at the median risk score.
Log-rank $p$-value tests whether SSBMF risk scores from the TCGA-trained model
transfer prognostic signal to the independent external cohort.

### LB Model (TCGA-only, point\_normal prior)

```{r km-lb, fig.height=3.8, out.width="100%", fig.cap="KM curves for LB model (TCGA-only, point\\_normal prior) in three external cohorts. High-risk = above median risk score (red); Low-risk = below median (blue)."}
par(mfrow=c(1,3), oma=c(0,0,1.5,0))
plot_km(lb_rs_tcga, "Dijk",             "point_normal", "Dijk (RNA-seq)")
plot_km(lb_rs_tcga, "Moffitt_GEO_array","point_normal", "Moffitt (array)")
plot_km(lb_rs_tcga, "PACA_AU_seq",      "point_normal", "PACA-AU (seq)")
```

```{r km-lb2, fig.height=3.8, out.width="67%", fig.cap="KM curves continued: two additional external cohorts."}
par(mfrow=c(1,2), oma=c(0,0,1.5,0))
plot_km(lb_rs_tcga, "PACA_AU_array","point_normal", "PACA-AU (array)")
plot_km(lb_rs_tcga, "Puleo_array",  "point_normal", "Puleo (array)")
```

---

## Summary of Phase Verdicts

```{r phase-summary}
ph_df <- data.frame(
  Phase   = c("1 — Per-platform standardization",
               "2 — Initialization constraints",
               "3 — Multi-start random restarts",
               "4 — K selection via CV"),
  Scope   = c("YFB merged", "LB + YFB", "LB", "LB"),
  Verdict = c("**Keep**", "Discard", "Discard", "**Keep** (infrastructure)"),
  `Key finding` = c(
    "Resolves YFB merged beta→0 collapse. K=3 gives median ext C=0.64.",
    "Constrained SVD init (non-neg L, Cox sign) gives no ELBO or C improvement.",
    "SVD init wins 30/30 restarts. ELBO range 0.3% — nearly unimodal landscape.",
    "Normal prior: flat plateau, K=3 (1-SE). PN: K=8 (spike artefact). Recommend K=5."
  ),
  check.names = FALSE, stringsAsFactors = FALSE
)
knitr::kable(ph_df,
  caption = "Summary of four improvement phases evaluated 2026-05-05/06.",
  booktabs = TRUE)
```

---

## Open Issues

1. **YFB single-cohort beta collapse.** With any prior, YFB on TCGA-only or CPTAC-only
   still produces near-zero $\hat\beta$. The ZF scale is still large for single-cohort
   training (n $\approx$ 130–145). A dedicated scale normalisation for single-cohort
   YFB (analogous to the Phase 1 per-platform fix) or a prior re-parameterisation
   (e.g., scaled spike-and-slab relative to $\|Z_F\|$) is the next candidate.

2. **A\_surv / A\_gen structural imbalance in LB.** Survival precision in the L update
   is $\approx 10^{3}$× smaller than genomic precision across all PDAC training modes.
   The L factors are effectively determined by PCA, with survival providing a negligible
   gradient. Phase C sign correction partially mitigates this, but does not resolve it.
   Potential fixes: survival-weighted initialisation, adaptive alpha scheduling,
   or a dedicated supervised L update with stronger survival weighting.

3. **K-CV on larger cohorts.** The flat CV plateau on TCGA\_PAAD (n=144) likely reflects
   limited statistical power to distinguish K values with overlapping C-index SEs.
   Running K-CV on the merged training set (n=273) or on Longleaf with the full
   PDAC multi-cohort data would sharpen the K selection.
