---
title: "Cohort Indicator Column: Model Comparison"
subtitle: "Current SSBMF vs. Augmented L-matrix Design"
author: "Andrew Walther"
date: "2026-05-22"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
    code-fold: true
    code-summary: "Show R code"
    fig-width: 9
    fig-height: 5
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| include: false
# Set working directory to repo root so all source() paths are relative to root.
# This is needed because Quarto renders from the document's directory by default.
proj_root <- normalizePath(file.path(getwd(), "../../.."))
knitr::opts_knit$set(root.dir = proj_root)
```

```{r load-libs}
#| include: false
library(ggplot2)
library(patchwork)
library(DT)
source("code/update_beta.R")
source("code/update_L.R")
source("code/update_F.R")
source("code/update_tau.R")
source("code/compute_elbo.R")
source("code/train_test_split.R")
source("code/predict.R")
suppressMessages(tryCatch(source("code/fit_modular.R"),
                          error = function(e) invisible(NULL)))
theme_set(theme_bw(base_size = 13))
```

## The β→0 Collapse Problem {#sec-collapse}

When TCGA (RNA-seq) and CPTAC (proteomics) data are trained together, the
dominant axis of variation in Y is **platform technology**, not biology.
The CAVI algorithm locks onto that axis and produces β→0 — a useless prognostic
score.

```{r collapse-sim}
set.seed(7)
n <- 100; p <- 200; K_bio <- 2
cohort <- rep(c("TCGA", "CPTAC"), each = n / 2)

# Biological factors (shared across cohorts)
L_bio  <- matrix(rnorm(n * K_bio), n, K_bio)
F_bio  <- matrix(rnorm(p * K_bio) * rbinom(p * K_bio, 1, 0.3), p, K_bio)

# Platform offset: strong signal on all genes for CPTAC
platform_offset <- rnorm(p, mean = 4, sd = 0.5)  # large shift
c_vec   <- as.integer(cohort == "CPTAC")
Y_clean <- L_bio %*% t(F_bio) + matrix(rnorm(n * p, sd = 0.5), n, p)
Y_dirty <- Y_clean + outer(c_vec, platform_offset)

# Column-centre both (as done in the real pipeline)
Y_clean <- sweep(Y_clean, 2, colMeans(Y_clean), "-")
Y_dirty <- sweep(Y_dirty, 2, colMeans(Y_dirty), "-")

# PCA on Y_dirty
pca_dirty <- prcomp(Y_dirty, center = FALSE)
pca_df <- data.frame(
  PC1    = pca_dirty$x[, 1],
  PC2    = pca_dirty$x[, 2],
  Cohort = cohort
)
var_exp <- round(summary(pca_dirty)$importance[2, 1:2] * 100, 1)
```

```{r collapse-plots, fig.width=11, fig.height=4.5}
# --- Heatmap: Y without vs with platform offset ---
# Sample 50 genes and 60 patients for display
gene_idx    <- sample(p, 50)
patient_idx <- c(sample(which(cohort == "TCGA"), 30),
                 sample(which(cohort == "CPTAC"), 30))

make_heatmap_df <- function(Y, label) {
  mat <- Y[patient_idx, gene_idx]
  df  <- as.data.frame(as.table(mat))
  colnames(df) <- c("Patient", "Gene", "Expression")
  df$Patient  <- as.integer(df$Patient)
  df$Cohort   <- ifelse(df$Patient <= 30, "TCGA", "CPTAC")
  df$Dataset  <- label
  df
}
hm_df <- rbind(make_heatmap_df(Y_clean, "Biological signal only"),
               make_heatmap_df(Y_dirty, "With platform offset (σ=4)"))

p_heat <- ggplot(hm_df, aes(Gene, Patient, fill = Expression)) +
  geom_tile() +
  scale_fill_gradient2(low = "#2166ac", mid = "white", high = "#d73027",
                       midpoint = 0, limits = c(-8, 8), oob = scales::squish) +
  facet_wrap(~ Dataset) +
  geom_hline(yintercept = 30.5, colour = "black", linewidth = 0.8) +
  annotate("text", x = 50, y = 15, label = "TCGA", size = 3.5, colour = "white") +
  annotate("text", x = 50, y = 45, label = "CPTAC", size = 3.5, colour = "white") +
  labs(x = "Gene (sample)", y = "Patient", fill = "Expr.",
       title = "Gene expression heatmap") +
  theme(axis.text = element_blank(), axis.ticks = element_blank(),
        strip.text = element_text(size = 11))

# --- PCA scatter ---
p_pca <- ggplot(pca_df, aes(PC1, PC2, colour = Cohort)) +
  geom_point(alpha = 0.8, size = 2) +
  scale_colour_manual(values = c(TCGA = "#1f78b4", CPTAC = "#e31a1c")) +
  labs(title = "PCA of Y with platform offset",
       x = sprintf("PC1 (%.1f%% var)", var_exp[1]),
       y = sprintf("PC2 (%.1f%% var)", var_exp[2])) +
  theme(legend.position = "bottom")

p_heat + p_pca + plot_layout(widths = c(2, 1))
```

**Interpretation:** PC1 captures `r var_exp[1]`% of variance and is entirely
driven by cohort membership. Any supervised factorisation that fits a single
shared L matrix will burn its first (and dominant) factor on this axis,
crowding out the biological signal and forcing β→0.

## Current Model {#sec-current}

```{=html}
<div style="font-family:monospace; background:#f8f9fa; border:1px solid #dee2e6;
            border-radius:6px; padding:20px; line-height:1.8; font-size:0.92em;">
<strong style="font-size:1.05em;">Current SSBMF</strong>
<hr style="border-color:#dee2e6; margin:8px 0;">

<table style="border-collapse:collapse; width:100%;">
<tr>
  <td style="color:#1a6ea8; font-weight:bold; padding:4px 12px 4px 0;">Genomics:</td>
  <td><span style="color:#1a6ea8;">Y</span> &nbsp;=&nbsp;
      <span style="color:#2ca02c;">L</span> &times;
      <span style="color:#9467bd;">F'</span> &nbsp;+&nbsp;
      <span style="color:#7f7f7f;">E</span>
  </td>
  <td style="color:#888; padding-left:30px; font-size:0.9em;">
    Y (n&times;p) &nbsp;&nbsp;
    <span style="color:#2ca02c;">L (n&times;K)</span> &nbsp;&nbsp;
    <span style="color:#9467bd;">F (p&times;K)</span> &nbsp;&nbsp;
    <span style="color:#7f7f7f;">E (n&times;p)</span>
  </td>
</tr>
<tr>
  <td style="color:#1a6ea8; font-weight:bold; padding:4px 12px 4px 0;">Survival:</td>
  <td>&eta; &nbsp;=&nbsp;
      <span style="color:#2ca02c;">L</span> &times;
      <span style="color:#d62728;">&beta;</span>
  </td>
  <td style="color:#888; padding-left:30px; font-size:0.9em;">
    &eta; (n&times;1) &nbsp;&nbsp;
    <span style="color:#2ca02c;">L (n&times;K)</span> &nbsp;&nbsp;
    <span style="color:#d62728;">&beta; (K&times;1)</span>
  </td>
</tr>
</table>

<div style="margin-top:14px; background:#e9ecef; border-radius:4px; padding:10px 14px;">
<strong>CAVI update order (each iteration):</strong><br>
&nbsp;&nbsp;<em>k = 1..K:</em>&nbsp;&nbsp;
  <span style="color:#2ca02c;">update q(L<sub>k</sub>)</span> &rarr;
  <span style="color:#9467bd;">update q(F<sub>k</sub>)</span> &rarr;
  <span style="color:#d62728;">update q(&beta;<sub>k</sub>)</span>
  &nbsp;&nbsp;<span style="color:#888;">[EBNM for L, F and &beta;]</span><br>
&nbsp;&nbsp;<em>then:</em>&nbsp;&nbsp;
  update q(&tau;)
  &nbsp;&nbsp;<span style="color:#888;">[closed-form MLE]</span>
</div>

<div style="margin-top:10px; color:#666; font-size:0.88em;">
  Priors: L, F, &beta; all use EBNM (point-normal default). &tau; ~ Gamma.
</div>
</div>
```

## Proposed Augmented Model {#sec-augmented}

The fix appends **C−1 fixed cohort indicator columns** to L (corner-point
encoding). For two cohorts (TCGA + CPTAC) this is exactly one binary column.
The corresponding F row f_c is estimated with a Normal(0, σ²) prior via a
**closed-form conjugate update** — no EBNM call needed. β_cohort ≡ 0 by
construction: the cohort column never enters the Cox partial likelihood.

```{=html}
<div style="display:grid; grid-template-columns:1fr 1fr; gap:20px; margin-top:10px;">

<!-- LEFT: Current model (abbreviated) -->
<div style="font-family:monospace; background:#fff3cd; border:1px solid #ffc107;
            border-radius:6px; padding:18px; line-height:1.8; font-size:0.9em;">
<strong>Current SSBMF</strong>
<hr style="border-color:#ffc107; margin:6px 0;">
Y &nbsp;=&nbsp; <span style="color:#2ca02c;">L</span> &times; <span style="color:#9467bd;">F'</span> + E<br>
&eta; =&nbsp; <span style="color:#2ca02c;">L</span> &times; <span style="color:#d62728;">&beta;</span>
<div style="margin-top:12px; font-size:0.88em; color:#666;">
  L (n&times;K) — latent factors<br>
  F (p&times;K) — gene loadings (EBNM)
</div>
</div>

<!-- RIGHT: Augmented model -->
<div style="font-family:monospace; background:#d4edda; border:1px solid #28a745;
            border-radius:6px; padding:18px; line-height:1.8; font-size:0.9em;">
<strong>Augmented SSBMF (cohort-cols-L)</strong>
<hr style="border-color:#28a745; margin:6px 0;">
Y &nbsp;=&nbsp; [<span style="color:#2ca02c;">L</span> | <span style="color:#e67e22; font-weight:bold;">c</span>] &times; [<span style="color:#9467bd;">F'</span>] + E<br>
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;[<span style="color:#e67e22; font-weight:bold;">f_c'</span>]<br>
&eta; =&nbsp; <span style="color:#2ca02c;">L</span> &times; <span style="color:#d62728;">&beta;</span> &nbsp;&nbsp;<em>(c never enters &eta;)</em>
<div style="margin-top:12px; font-size:0.88em; color:#666;">
  L (n&times;K) — latent factors (unchanged)<br>
  <span style="color:#e67e22; font-weight:bold;">c (n&times;1) — FIXED binary indicator (not updated)</span><br>
  F (p&times;K) — gene loadings (EBNM, unchanged)<br>
  <span style="color:#e67e22; font-weight:bold;">f_c (p&times;1) — platform loadings (Normal conjugate)</span>
</div>
</div>

</div>

<div style="font-family:monospace; background:#f8f9fa; border:1px solid #dee2e6;
            border-radius:6px; padding:14px 18px; margin-top:14px; line-height:1.8; font-size:0.9em;">
<strong>Augmented CAVI update order (each iteration):</strong><br>
&nbsp;&nbsp;<em>k = 1..K:</em>&nbsp;&nbsp;
  <span style="color:#2ca02c;">update q(L<sub>k</sub>)</span> &rarr;
  <span style="color:#9467bd;">update q(F<sub>k</sub>)</span> &rarr;
  <span style="color:#d62728;">update q(&beta;<sub>k</sub>)</span>
  &nbsp;&nbsp;<span style="color:#888;">[EBNM — unchanged]</span><br>
&nbsp;&nbsp;<em>then:</em>&nbsp;&nbsp;
  <span style="color:#e67e22; font-weight:bold;">update q(f_c)</span>
  &nbsp;&nbsp;<span style="color:#888;">[NEW: Normal conjugate, closed-form]</span><br>
&nbsp;&nbsp;<em>then:</em>&nbsp;&nbsp;
  update q(&tau;)
  &nbsp;&nbsp;<span style="color:#888;">[passes augmented matrices — unchanged function]</span>
</div>
```

## CAVI Loop Change Summary {#sec-changes}

```{r change-table}
#| echo: false
changes <- data.frame(
  Component = c(
    "compute_R_k", "update_L_k (EBNM)", "update_F_k (EBNM)",
    "update_beta_k", "compute_var_term", "predict_*",
    "update_F_cohort_col", "compute_normal_kl",
    "fit_modular.R", "fit_cox_on_yf.R"
  ),
  Status = c(
    "Unchanged", "Unchanged", "Unchanged",
    "Unchanged", "Unchanged", "Unchanged",
    "New", "New",
    "Modified", "Modified"
  ),
  Notes = c(
    "Augmented EL_aug/EF_aug passed in; cohort contribution subtracted automatically",
    "Loop k=1..K only; c is a fixed design column — never updated",
    "Loop k=1..K only; f_c uses new closed-form function",
    "beta is a K-vector; cohort never enters the Cox partial likelihood",
    "Binary EL: EL²=EL → zero cohort L-variance; passes through correctly",
    "Returns global EF (p×K) only; cohort fields optional in return list",
    "Closed-form Normal conjugate update for f_c; ~20 lines",
    "ELBO KL term for Normal posterior vs N(0,σ²); returns negative scalar",
    "~50 new lines, all guarded by if(!is.null(cohort_id))",
    "Same as fit_modular.R + ZF dimension guard (two locations)"
  ),
  stringsAsFactors = FALSE
)

status_colours <- c(
  Unchanged = "#d4edda",
  New       = "#cce5ff",
  Modified  = "#fff3cd"
)

DT::datatable(
  changes,
  options  = list(dom = "t", pageLength = 15, ordering = FALSE),
  rownames = FALSE,
  class    = "compact stripe"
) |>
  DT::formatStyle(
    "Status",
    backgroundColor = DT::styleEqual(
      names(status_colours), unname(status_colours)
    ),
    fontWeight = "bold"
  )
```

## Mathematical Derivation of q(f_c) {#sec-math}

The cohort indicator column c is **fixed** (observed, not estimated). Only
the corresponding row of F — the platform loading vector f_c — is inferred.

**Likelihood** (rows in cohort B, i.e., c_i = 1):

$$
p(\mathbf{R}_c \mid \mathbf{f}_c, \tau) = \prod_{i:\, c_i=1} \prod_{j=1}^{p}
\mathcal{N}\!\left(R_{ij};\; f_{cj},\; \tau_j^{-1}\right)
$$

where $R_{ij} = Y_{ij} - \sum_{k=1}^{K} E[l_{ik}] E[f_{kj}]$ is the global-factor
residual (Y minus the K biological factors).

**Prior:** $f_{cj} \overset{\text{ind}}{\sim} \mathcal{N}(0, \sigma^2)$

**Posterior precision and mean** — conjugate Gaussian update:

$$
A_{cj} = n_c \,\tau_j + \sigma^{-2}, \qquad
B_{cj} = \tau_j \sum_{i:\,c_i=1} R_{ij}
$$

$$
\bar{f}_{cj} = \frac{B_{cj}}{A_{cj}}, \qquad
\text{Var}_q(f_{cj}) = A_{cj}^{-1}, \qquad
E_q[f_{cj}^2] = A_{cj}^{-1} + \bar{f}_{cj}^2
$$

This is the **p-vector** update (one scalar per gene):

```
n_c   <- sum(c_vec)                          # scalar: number of cohort-B patients
B_c   <- colSums(R_cohort * c_vec) * Tau     # p-vector
A_c   <- n_c * Tau + 1 / sigma_F_cohort^2   # p-vector
EF_c  <- B_c / A_c                           # p-vector (posterior means)
EF2_c <- 1 / A_c + EF_c^2                   # p-vector (posterior second moments)
```

**KL divergence** (contribution to ELBO):

$$
\text{KL}\!\left[\mathcal{N}(\bar{\mathbf{f}}_c, A_c^{-1}) \;\|\;
\mathcal{N}(\mathbf{0}, \sigma^2 I)\right]
= \frac{1}{2} \sum_{j=1}^{p}
\left(\frac{E[f_{cj}^2]}{\sigma^2} + \log(A_{cj}\,\sigma^2) - 1\right)
$$

**Sign convention:** `compute_normal_kl()` returns **−KL ≤ 0** (matching the
existing `compute_ebnm_kl()` convention). The caller adds it:

```r
elbo_full <- elbo_full + compute_normal_kl(EF_cohort, EF2_cohort, sigma_F_cohort)
```

> **Contrast with EBNM factors:** EBNM requires numerical optimisation of an
> adaptive shrinkage prior. The cohort f_c update is exact in closed form —
> no `ebnm()` call needed. This is why it can be implemented in ~20 lines and
> runs in negligible time even for p = 10,000.

## Before/After ELBO Equations {#sec-elbo}

```{=html}
<div style="display:grid; grid-template-columns:1fr 1fr; gap:20px; margin-top:10px;">

<div style="font-family:monospace; background:#f8f9fa; border:1px solid #dee2e6;
            border-radius:6px; padding:16px; font-size:0.88em; line-height:2;">
<strong>Current ELBO</strong>
<hr style="border-color:#dee2e6; margin:6px 0;">
E[log p(Y | L, F, &tau;)]<br>
+ E[log p(L)]<br>
+ E[log p(F)]<br>
+ E[log p(&beta;)]<br>
+ E[log p(&tau;)]<br>
+ E[log PL(&beta;, L)]<br>
&minus; KL[q(L) &Vert; p(L)]<br>
&minus; KL[q(F) &Vert; p(F)]<br>
&minus; KL[q(&beta;) &Vert; p(&beta;)]<br>
&nbsp;
</div>

<div style="font-family:monospace; background:#d4edda; border:1px solid #28a745;
            border-radius:6px; padding:16px; font-size:0.88em; line-height:2;">
<strong>Augmented ELBO</strong>
<hr style="border-color:#28a745; margin:6px 0;">
E[log p(Y | L<sub>aug</sub>, F<sub>aug</sub>, &tau;)] &nbsp;<span style="color:#888;">(aug. L,F)</span><br>
+ E[log p(L)] &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;<span style="color:#888;">(unchanged)</span><br>
+ E[log p(F)] &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;<span style="color:#888;">(unchanged)</span><br>
+ E[log p(&beta;)] &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;<span style="color:#888;">(unchanged)</span><br>
+ E[log p(&tau;)] &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;<span style="color:#888;">(unchanged)</span><br>
+ E[log PL(&beta;, L)] &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;<span style="color:#888;">(unchanged)</span><br>
&minus; KL[q(L) &Vert; p(L)] &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;<span style="color:#888;">(unchanged)</span><br>
&minus; KL[q(F) &Vert; p(F)] &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;<span style="color:#888;">(unchanged)</span><br>
&minus; KL[q(&beta;) &Vert; p(&beta;)] &nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;<span style="color:#888;">(unchanged)</span><br>
<span style="color:#28a745; font-weight:bold;">&minus; KL[q(f_c) &Vert; N(0,&sigma;²)] &larr; NEW (one scalar)</span>
</div>

</div>
```

The genomics likelihood term automatically incorporates the cohort contribution
because $L_\text{aug}$ and $F_\text{aug}$ now include the cohort column. All other
terms are structurally unchanged — the only genuinely new term is the KL for f_c,
computed by `compute_normal_kl()` in a single `sum()` call.

## Simulation: Platform PC Absorption {#sec-sim}

A small controlled experiment (n=80, p=100, K=2 biological factors,
σ_platform = 3) to verify that the augmented model absorbs the platform PC into
f_c, leaving the biological factors uncontaminated.

```{r absorption-sim}
set.seed(42)
n_sim   <- 80
p_sim   <- 100
K_sim   <- 2

cohort_sim <- rep(c("A", "B"), each = n_sim / 2)
c_sim      <- as.integer(cohort_sim == "B")

# Biological factors
L_bio_sim <- matrix(rnorm(n_sim * K_sim), n_sim, K_sim)
F_bio_sim <- matrix(rnorm(p_sim * K_sim) * rbinom(p_sim * K_sim, 1, 0.4),
                    p_sim, K_sim)

# Strong platform offset
F_plat_true <- rnorm(p_sim, mean = 3, sd = 0.3)
Y_sim <- L_bio_sim %*% t(F_bio_sim) +
         outer(c_sim, F_plat_true) +
         matrix(rnorm(n_sim * p_sim, sd = 0.4), n_sim, p_sim)
Y_sim <- sweep(Y_sim, 2, colMeans(Y_sim), "-")

tt_sim <- rexp(n_sim, 0.1)
ss_sim <- rbinom(n_sim, 1, 0.7)

# --- Fit base model (no cohort correction) ---
fit_base_sim <- fit_supervised_mf_modular(
  Y_sim, tt_sim, ss_sim, K = K_sim,
  max_iter = 60, alpha = 0.5, verbose = FALSE
)

# --- Proxy for augmented model: subtract per-cohort column means from Y ---
# This is mathematically equivalent to the full cohort-cols-L model in the
# weak-prior limit (sigma_F_cohort -> Inf), and demonstrates the platform
# absorption property. The full augmented fit_supervised_mf_modular(cohort_id=...)
# is implemented in Steps 3-8 of this branch.
Y_corrected <- Y_sim
for (g in unique(cohort_sim)) {
  idx <- cohort_sim == g
  Y_corrected[idx, ] <- sweep(Y_sim[idx, , drop = FALSE], 2,
                               colMeans(Y_sim[idx, , drop = FALSE]), "-")
}

fit_aug_sim <- fit_supervised_mf_modular(
  Y_corrected, tt_sim, ss_sim, K = K_sim,
  max_iter = 60, alpha = 0.5, verbose = FALSE
)
```

```{r absorption-plots, fig.width=11, fig.height=4.5}
make_loading_df <- function(fit, cohort, label) {
  data.frame(
    PC1    = fit$EL[, 1],
    PC2    = fit$EL[, 2],
    Cohort = cohort,
    Model  = label
  )
}

df_base <- make_loading_df(fit_base_sim, cohort_sim, "Base (no cohort column)")
df_aug  <- make_loading_df(fit_aug_sim,  cohort_sim, "Augmented (cohort-cols-L)")
plot_df <- rbind(df_base, df_aug)

violin_df <- rbind(
  data.frame(Loading = fit_base_sim$EL[, 1], Cohort = cohort_sim,
             Model = "Base"),
  data.frame(Loading = fit_aug_sim$EL[, 1],  Cohort = cohort_sim,
             Model = "Augmented")
)

p_scatter <- ggplot(plot_df, aes(PC1, PC2, colour = Cohort)) +
  geom_point(alpha = 0.8, size = 2) +
  facet_wrap(~ Model) +
  scale_colour_manual(values = c(A = "#1f78b4", B = "#e31a1c")) +
  labs(title = "Factor 1 vs Factor 2 scores", x = "Factor 1", y = "Factor 2") +
  theme(legend.position = "bottom")

p_violin <- ggplot(violin_df, aes(Cohort, Loading, fill = Cohort)) +
  geom_violin(alpha = 0.7, trim = FALSE) +
  geom_jitter(width = 0.1, size = 0.8, alpha = 0.5) +
  facet_wrap(~ Model) +
  scale_fill_manual(values = c(A = "#1f78b4", B = "#e31a1c")) +
  labs(title = "Factor 1 scores by cohort",
       subtitle = "Split = residual platform effect",
       y = "Factor 1 loading") +
  theme(legend.position = "none")

p_scatter / p_violin
```

```{r absorption-summary}
#| echo: false
sep_base <- abs(mean(fit_base_sim$EL[cohort_sim == "B", 1]) -
                mean(fit_base_sim$EL[cohort_sim == "A", 1]))
sep_aug  <- abs(mean(fit_aug_sim$EL[cohort_sim == "B", 1]) -
                mean(fit_aug_sim$EL[cohort_sim == "A", 1]))
cat(sprintf("Mean Factor-1 cohort separation (base model):      %.3f\n", sep_base))
cat(sprintf("Mean Factor-1 cohort separation (augmented model): %.3f\n", sep_aug))
```

**Interpretation:** In the base model, Factor 1 is dominated by the platform
offset: cohort-B patients have systematically higher scores. The augmented
proxy (cohort-mean-subtracted Y, mathematically equivalent to cohort-cols-L in
the weak-prior limit) shows the same factor no longer separated by cohort —
the platform signal is absorbed, leaving the biological factors free. The cohort
separation drops from `r round(sep_base, 2)` to `r round(sep_aug, 2)`.
Once `fit_supervised_mf_modular(cohort_id=...)` is implemented in Steps 3–8,
this section will be updated to use the full augmented model directly.
