---
title: "SSBMF Status Update - May 8, 2026"
author: "Andrew Walther"
format:
  pdf:
    latex-engine: xelatex
    keep-tex: false
    fig-pos: "htbp"
    include-in-header:
      text: |
        \usepackage{enumitem}
        \usepackage{booktabs}
        \usepackage{geometry}
        \usepackage{titling}
        \usepackage{array}
        \geometry{margin=0.42in}
        \renewcommand{\familydefault}{\sfdefault}
        \setlist[itemize]{leftmargin=1.1em, topsep=0.1em, itemsep=0.06em, parsep=0em, partopsep=0em}
        \setlength{\parskip}{0.05em}
        \RedeclareSectionCommand[beforeskip=0.28em,afterskip=0.14em]{section}
        \renewcommand{\arraystretch}{0.9}
        \setlength{\droptitle}{-1.8em}
        \pretitle{\begin{center}\large\bfseries}
        \posttitle{\par\end{center}\vspace{-0.7em}}
fontsize: 9pt
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(knitr)
library(ggplot2)
library(patchwork)

read_csv_safe <- function(path) {
  if (file.exists(path)) read.csv(path, stringsAsFactors = FALSE) else NULL
}

median_ext <- function(df, prior) {
  sub <- df[df$section == "3_external" & df$prior_beta == prior, "c_index"]
  round(median(sub, na.rm = TRUE), 3)
}

train_row <- function(df, prior) {
  df[df$section == "2_pdac_train" & df$prior_beta == prior, ][1, ]
}

OUT_LB <- "../../results/benchmark_sim/outputs/LB_benchmark"
OUT_YFB <- "../../results/benchmark_sim/outputs/YFB_benchmark"
OUT_YFB_P1 <- "../../results/benchmark_sim/outputs/YFB_benchmark_perplatform"
OUT_KCV <- "../../results/benchmark_sim/outputs/K_cv"

lb_tcga  <- read_csv_safe(file.path(OUT_LB, "LB_benchmark_results_tcga_only.csv"))
lb_cptac <- read_csv_safe(file.path(OUT_LB, "LB_benchmark_results_cptac_only.csv"))
lb_merge <- read_csv_safe(file.path(OUT_LB, "LB_benchmark_results_merged.csv"))

yfb_tcga  <- read_csv_safe(file.path(OUT_YFB, "YFB_benchmark_results_tcga_only.csv"))
yfb_cptac <- read_csv_safe(file.path(OUT_YFB, "YFB_benchmark_results_cptac_only.csv"))
yfb_merge_p1 <- read_csv_safe(file.path(OUT_YFB_P1, "YFB_benchmark_results_merged.csv"))

kcv_lb  <- read_csv_safe(file.path(OUT_KCV, "K_cv_comparison_LB.csv"))
kcv_yfb <- read_csv_safe(file.path(OUT_KCV, "K_cv_comparison_YFB.csv"))

lb_tcga_row  <- train_row(lb_tcga, "point_normal")
lb_cptac_row <- train_row(lb_cptac, "point_normal")
lb_merge_row <- train_row(lb_merge, "point_normal")
yfb_tcga_row  <- train_row(yfb_tcga, "normal")
yfb_cptac_row <- train_row(yfb_cptac, "normal")
yfb_merge_row <- train_row(yfb_merge_p1, "normal")
```

## 1. Recent Improvements: Problems Addressed and Resolved

- **Risk direction is now corrected.** The 2026-05-05 Phase C sign correction
  fixed a major anti-concordance problem in both LB and YFB. This is now a
  resolved issue and the basis for interpreting all current external results.
- **Merged YFB collapse was resolved on 2026-05-06.** The 2026-05-05 report
  still showed merged YFB as effectively collapsed; the later 2026-05-06
  update supersedes that result. Per-platform z-standardization plus removal of
  the rank transform recovered non-zero prognostic coefficients for merged
  TCGA+CPTAC.
- **Initialization constraints were tested and discarded.** Constrained
  singular value decomposition initialization and Cox-sign alignment did not
  improve ELBO or external prognosis.
- **Multi-start was tested and discarded.** A 30-restart sweep showed the
  standard singular value decomposition initialization already wins on ELBO, so
  routine restarts are not the current bottleneck.
- **K-selection is now a clearer methodological decision.** The 2026-05-06 work
  added an implemented K-fold CV pipeline and clarified what is currently
  reliable for LB versus still unresolved for YFB.

**Brief note on YFB model development.**

- **LB formulation:** prognosis enters through the latent loading matrix,
  $\eta = L\beta$.
- **YFB reformulation:** prognosis instead uses observed factor scores,
  $\eta = (YF)\beta$, which removes the LB train/test mismatch.
- **Derivation impact:** the survival term no longer acts through the original
  $L$-based predictor, so YFB required a separate Cox-on-YF derivation and
  YFB-specific update rules.

## 2. Current Results and Value Add: LB and YFB

```{r key-results-table}
key_results <- data.frame(
  Model = c("LB", "LB", "LB", "YFB", "YFB", "YFB"),
  `Training set` = c(
    "TCGA-only",
    "CPTAC-only",
    "Merged TCGA+CPTAC",
    "TCGA-only",
    "CPTAC-only",
    "Merged TCGA+CPTAC"
  ),
  `Prior(s)` = c(
    "normal / point_normal",
    "normal / point_normal",
    "normal / point_normal",
    "normal / point_normal",
    "normal / point_normal",
    "normal / point_normal"
  ),
  `Median external C by prior` = c(
    sprintf("normal %.3f / point_normal %.3f",
            median_ext(lb_tcga, "normal"), median_ext(lb_tcga, "point_normal")),
    sprintf("normal %.3f / point_normal %.3f",
            median_ext(lb_cptac, "normal"), median_ext(lb_cptac, "point_normal")),
    sprintf("normal %.3f / point_normal %.3f",
            median_ext(lb_merge, "normal"), median_ext(lb_merge, "point_normal")),
    sprintf("normal %.3f / point_normal %.3f",
            median_ext(yfb_tcga, "normal"), median_ext(yfb_tcga, "point_normal")),
    sprintf("normal %.3f / point_normal %.3f",
            median_ext(yfb_cptac, "normal"), median_ext(yfb_cptac, "point_normal")),
    sprintf("normal %.3f / point_normal %.3f",
            median_ext(yfb_merge_p1, "normal"), median_ext(yfb_merge_p1, "point_normal"))
  ),
  `Factor / beta summary` = c(
    "LB priors agree; 2 active; max|beta|=0.0085",
    "LB priors agree; 3 active; max|beta|=0.0152",
    "LB priors agree; 3 active; max|beta|=0.0213",
    "point_normal collapses; normal has near-zero betas",
    "point_normal collapses; normal has near-zero betas",
    "both priors agree; 3 active; max|beta|=0.0677"
  ),
  check.names = FALSE, stringsAsFactors = FALSE
)
knitr::kable(key_results, booktabs = TRUE, digits = 3)
```

Notes: LB gives the same current external predictions under `normal` and
`point_normal`. For YFB single-cohort fits, `point_normal` collapses to
beta approximately zero, so the `normal` prior gives the only non-degenerate
single-cohort result. Merged YFB uses the same TCGA+CPTAC training set, with
per-platform standardization and no rank transform; under those settings, both
priors agree.

- **LB is now the stable benchmark model.**
  - Median external C-index is **0.610** for TCGA-only, **0.625** for
    CPTAC-only, and **0.644** for merged training.
  - Best cohort result is **PACA-AU seq: 0.656-0.673**.
  - LB consistently identifies **2-3 active factors**.
- **Why only one prior is emphasized for LB.**
  - In the current LB benchmarks, **point-normal and normal give effectively
    identical external predictions**, so showing both is redundant; the document
    reports point-normal because it is the current default prior.
- **YFB remains mixed in single-cohort mode.**
  - With the normal prior, median external C-index is **0.553** for TCGA-only
    and **0.561** for CPTAC-only.
  - In the current single-cohort fits, the reported betas are effectively
    near-zero, so these models are not yet giving a convincing activated-factor
    solution even though some external C-indices are above 0.5.
- **Why only the normal prior is emphasized for single-cohort YFB.**
  - In the current YFB single-cohort benchmarks, the **point-normal prior
    collapses to beta≈0**, so the normal prior is the only interpretable
    single-cohort YFB result. On the merged TCGA+CPTAC fit with the working
    preprocessing settings, **both priors agree**.
- **Merged YFB is currently the strongest “value add” result.**
  - On the same merged **TCGA+CPTAC** training set, using
    **per-platform standardization and no rank transform**, YFB achieves
    **$K = 3$**, **$K_{\mathrm{eff}} = 3$**, **max$|\hat\beta| = 0.0677$**, and
    external C-index **0.532-0.666** across all five cohorts.
  - Median external C-index is **0.644**, matching merged LB.

\newpage

## 3. Open Questions and Decisions

```{r k-options}
k_options <- data.frame(
  `K-selection path` = c("ARD from K_max", "Bayesian cross-validation", "K-fold CV grid search"),
  `Current status in repo` = c(
    "Partially present via shrinkage and K_eff behavior",
    "Not yet implemented in a decision-ready form",
    "Implemented; currently the only operational selection workflow"
  ),
  `What recent results suggest` = c(
    "Useful regularization, but not yet a sufficient standalone decision rule in PDAC",
    "Methodologically appealing, but currently a future development path",
    "Reliable for LB; still unreliable for YFB because of fold-level sign instability"
  ),
  check.names = FALSE, stringsAsFactors = FALSE
)
knitr::kable(k_options, booktabs = TRUE)
```

- **Current LB recommendation.**
  - Under the 2026-05-06 K-CV sweep, the **normal prior selects $K=3$** under
    the 1-SE rule.
  - The **point-normal prior selects $K=8$**, but this is interpreted as a
    shrinkage/small-$n$ artifact rather than strong evidence that 8 factors are
    biologically necessary.
  - The current practical compromise is **$K=5$ for LB**.
- **Current YFB limitation.**
  - YFB K-CV is **not yet decision-ready**.
  - Point-normal collapses to **C = 0.5** in all folds; normal prior gives
    **anti-concordant C-index** across K because held-out folds do not yet have
    stable sign correction.
- **Main decisions for discussion.**
  - Should **K-fold CV remain the operational rule for LB** while YFB
    K-selection remains open?
  - Should the next methodological push be **Bayesian CV** or **ARD-style
    large-$K_{\max}$ shrinkage**?

- **Other open issues.**
  - **Single-cohort YFB still needs a scale-aware fix** analogous to the merged
    preprocessing fix that worked on the merged TCGA+CPTAC training set.
  - **LB still appears genomics-dominated in the loading update**, so survival
    influence on $L$ may remain too weak even after sign correction.

```{r summary-figure, fig.height=4.8, fig.cap="Current evidence across synthetic validation, PDAC external performance, and K-selection. Dashed lines mark selected K values from the 2026-05-06 sweep."}
# Panel A: synthetic
syn_df <- data.frame(
  Model = c("LB", "LB", "YFB", "YFB"),
  Prior = c("point_normal", "normal", "point_normal", "normal"),
  C_index = c(
    lb_merge[lb_merge$section == "1_synthetic" & lb_merge$prior_beta == "point_normal", "c_index"][1],
    lb_merge[lb_merge$section == "1_synthetic" & lb_merge$prior_beta == "normal", "c_index"][1],
    yfb_merge_p1[yfb_merge_p1$section == "1_synthetic" & yfb_merge_p1$prior_beta == "point_normal", "c_index"][1],
    yfb_merge_p1[yfb_merge_p1$section == "1_synthetic" & yfb_merge_p1$prior_beta == "normal", "c_index"][1]
  )
)

p_syn <- ggplot(syn_df, aes(x = Model, y = C_index, fill = Prior)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6, color = "white") +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey40") +
  scale_fill_manual(values = c("point_normal" = "#4c78a8", "normal" = "#f58518")) +
  coord_cartesian(ylim = c(0.5, 1.0)) +
  labs(title = "Synthetic", x = NULL, y = "C-index") +
  theme_bw(base_size = 8) +
  theme(legend.position = "top", panel.grid.minor = element_blank())

# Panel B: median external results
ext_df <- data.frame(
  Model = c(rep("LB", 6), rep("YFB", 6)),
  Mode = c("TCGA-only", "TCGA-only", "CPTAC-only", "CPTAC-only", "Merged", "Merged",
           "TCGA-only", "TCGA-only", "CPTAC-only", "CPTAC-only", "Merged", "Merged"),
  Prior = c("point_normal", "normal", "point_normal", "normal", "point_normal", "normal",
            "point_normal", "normal", "point_normal", "normal", "point_normal", "normal"),
  Median = c(
    median_ext(lb_tcga, "point_normal"), median_ext(lb_tcga, "normal"),
    median_ext(lb_cptac, "point_normal"), median_ext(lb_cptac, "normal"),
    median_ext(lb_merge, "point_normal"), median_ext(lb_merge, "normal"),
    median_ext(yfb_tcga, "point_normal"), median_ext(yfb_tcga, "normal"),
    median_ext(yfb_cptac, "point_normal"), median_ext(yfb_cptac, "normal"),
    median_ext(yfb_merge_p1, "point_normal"), median_ext(yfb_merge_p1, "normal")
  )
)

p_ext <- ggplot(ext_df, aes(x = Mode, y = Median, fill = Prior)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.62, color = "white") +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey40") +
  facet_wrap(~Model, nrow = 1) +
  scale_fill_manual(values = c("point_normal" = "#4c78a8", "normal" = "#f58518")) +
  coord_cartesian(ylim = c(0.45, 0.7)) +
  labs(title = "Median External C-index", x = NULL, y = "Median C") +
  theme_bw(base_size = 8) +
  theme(legend.position = "none", panel.grid.minor = element_blank(),
        axis.text.x = element_text(size = 7))

# Panel C: K-CV
kcv_plot_df <- rbind(
  data.frame(Model = "LB", Prior = "point_normal", K = kcv_lb$K, MeanC = kcv_lb$PN_mean_C),
  data.frame(Model = "LB", Prior = "normal", K = kcv_lb$K, MeanC = kcv_lb$NM_mean_C),
  data.frame(Model = "YFB", Prior = "point_normal", K = kcv_yfb$K, MeanC = kcv_yfb$PN_mean_C),
  data.frame(Model = "YFB", Prior = "normal", K = kcv_yfb$K, MeanC = kcv_yfb$NM_mean_C)
)

sel_df <- data.frame(
  Model = c("LB", "LB", "YFB", "YFB"),
  Prior = c("point_normal", "normal", "point_normal", "normal"),
  K = c(8, 3, 2, 9)
)

p_kcv <- ggplot(kcv_plot_df, aes(x = K, y = MeanC, color = Prior)) +
  geom_line(linewidth = 0.55) +
  geom_point(size = 1.2) +
  geom_vline(data = sel_df, aes(xintercept = K, color = Prior),
             linetype = "dashed", linewidth = 0.45, show.legend = FALSE) +
  geom_hline(yintercept = 0.5, linetype = "dotted", color = "grey40") +
  facet_wrap(~Model, nrow = 1) +
  scale_color_manual(values = c("point_normal" = "#4c78a8", "normal" = "#f58518")) +
  coord_cartesian(ylim = c(0.38, 0.62)) +
  labs(title = "K-fold CV", x = "K", y = "Mean C") +
  theme_bw(base_size = 8) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())

p_syn / p_ext / p_kcv + plot_layout(heights = c(0.9, 1.05, 1.05))
```
