---
title: "SSBMF Status Update: DeSurv Gene Selection, Factor Diagnostics, and Recommended Configuration (May 28, 2026)"
author: "Andrew Walther"
format:
  pdf:
    latex-engine: xelatex
    keep-tex: false
    fig-pos: "htbp"
    include-in-header:
      text: |
        \usepackage{enumitem}
        \usepackage{booktabs}
        \usepackage{geometry}
        \usepackage{titling}
        \usepackage{array}
        \geometry{margin=0.42in}
        \renewcommand{\familydefault}{\sfdefault}
        \setlist[itemize]{leftmargin=1.1em, topsep=0.1em, itemsep=0.06em, parsep=0em, partopsep=0em}
        \setlength{\parskip}{0.05em}
        \RedeclareSectionCommand[beforeskip=0.28em,afterskip=0.14em]{section}
        \renewcommand{\arraystretch}{0.9}
        \setlength{\droptitle}{-1.8em}
        \pretitle{\begin{center}\large\bfseries}
        \posttitle{\par\end{center}\vspace{-0.7em}}
fontsize: 9pt
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(knitr)
library(ggplot2)
library(dplyr)

OUT <- "../../results/benchmark_sim/outputs/desurv_comparison"
csv_path <- file.path(OUT, "desurv_comparison_results.csv")
res <- if (file.exists(csv_path)) read.csv(csv_path, stringsAsFactors = FALSE) else NULL

mean_c <- function(id) if (!is.null(res)) round(mean(res$c_index[res$model == id]), 3) else NA
```

## 1. Objective

This update covers the evaluation of **DeSurv-aligned gene selection** as a replacement for the variance-only selection used in the prior benchmark (M4/M5). The question: does matching the gene selection strategy from Young et al. (2025) — selecting genes that are both highly expressed *and* highly variable within each platform — improve external concordance and yield more interpretable factor structure? A five-configuration comparison (D1–D5) was run on the merged TCGA\_PAAD + CPTAC training cohort ($n = 273$) and evaluated on 5 held-out PDAC cohorts.

## 2. Gene Selection: Original vs. DeSurv-Aligned

The two approaches differ in three respects: *what* is ranked, *when* it is computed (before or after cross-cohort normalization), and *where* it is computed (per-platform or on the merged matrix).

| | Original (D1/D2) | DeSurv-aligned (D3/D4/D5) |
|---|---|---|
| Metric | Within-cohort variance | Combined mean + variance rank |
| Top $n$ | 2,000 | 3,000 per cohort |
| Timing | Post-normalization on merged matrix | Pre-normalization on log-transformed data |
| Selection scope | Cross-cohort (merged matrix) | Per-platform, then intersect |
| Genes after intersection | 2,000 | 2,064 |

**Combined mean+variance rank:** for each cohort independently, genes are ranked by their mean expression and by their variance after log$_2$(count+1) transformation. The combined rank is their sum; the top 3,000 per cohort are retained. The intersection across cohorts is taken before per-platform z-standardization and merging. This retains genes that are both reliably detected above the expression floor and genuinely variable across samples — variance-only selection can include noisy low-expressed genes near the detection limit.

## 3. Configurations and Results

```{r results-table}
tab <- data.frame(
  ID    = c("D1","D2","D3","D4","D5"),
  Label = c("LB orig (M4)","YFB orig (M5)","LB DeSurv","YFB DeSurv","YFB DeSurv+cohort"),
  Model = c("LB","YFB","LB","YFB","YFB"),
  Genes = c(2000, 2000, 2064, 2064, 2064),
  K     = c(3, 3, 7, 7, 8),
  K_eff = c(1, 2, 2, 2, 2),
  Mean_C = c(mean_c("D1"), mean_c("D2"), mean_c("D3"), mean_c("D4"), mean_c("D5")),
  stringsAsFactors = FALSE
)
knitr::kable(tab, booktabs = TRUE, row.names = FALSE,
  col.names = c("ID","Label","Model","Genes","K","$K_{\\text{eff}}$","Mean C"),
  escape = FALSE,
  caption = "Five configurations on merged TCGA + CPTAC ($n=273$). K = factors requested (CV-selected with biological floor $K\\geq3$); $K_{\\text{eff}}$ = active factors ($|\\hat{\\beta}_k| > 0.001$) after convergence. Mean C = mean concordance across 5 held-out PDAC cohorts.")
```

**DeSurv gene selection improves external concordance in both model families.** For YFB: $\Delta C = +0.012$ (D4 vs. D2). For LB: $\Delta C = +0.006$ (D3 vs. D1). The YFB formulation additionally outperforms LB on the same DeSurv gene set: $\Delta C = +0.014$ (D4 vs. D3). These effects are additive, and both exceed the pre-specified $\pm 0.005$ practical threshold.

**Cohort indicator (D5) reduces performance.** Adding a corner-point cohort indicator to D4 ($K = 8$) yields mean C = `r mean_c("D5")` — 0.022 below D4. The per-platform z-standardization applied before merging already removes the bulk platform offset; the additional cohort column absorbs degrees of freedom that would otherwise be allocated to survival-relevant factor structure. The indicator is harmful at $K = 7$–$8$ and should be excluded from the primary configuration.

## 4. Factor Diagnostics: Convergence, Survival Associations, and Gene Programs

Full diagnostics for all five configurations are in `docs/reports/desurv_factor_diagnostics_05_27_26.pdf`. Key findings:

**ARD shrinkage consistently identifies two active programs.** All five configurations converge to $K_{\text{eff}} = 2$ active programs regardless of the total factors requested ($K = 3$–$8$). The empirical Bayes normal mixture prior shrinks inactive factors to $\hat{\beta} \approx 0$; the two surviving programs carry opposite survival directions — one adverse, one protective — in every configuration.

**D4 active factors (recommended configuration):**

- *Factor 3 (adverse):* $\hat{\beta}_3 = +0.0113$; high-loading patients have worse survival; log-rank $p < 0.05$ on the training cohort. Median-split K-M curves show early separation with the high-loading group accumulating events faster across the follow-up period.
- *Factor 7 (protective):* $\hat{\beta}_7 = -0.0408$; high-loading patients have better survival. Separation is visible throughout follow-up and directionally consistent in all five external cohorts (C $\geq 0.55$ across Dijk, Moffitt GEO, PACA-AU array, PACA-AU seq, Puleo).

The survival directions are internally consistent: $\hat{\beta}$ sign predicts which Kaplan-Meier arm fares worse in every active factor across all five configurations. This directional coherence is a necessary condition for the model's survival signal being real rather than artifactual.

**YFB coefficient scale:** $\hat{\beta}$ values for YFB configurations (D2, D4, D5) are smaller in magnitude than their LB counterparts (D1, D3) because the YFB formulation normalizes $F$ columns to unit Euclidean norm, compressing the inner product scale by $\sim\sqrt{p}$. The survival signal encoded is equivalent; only the parameterization differs.

**Convergence:** YFB models (D2, D4, D5) converge in $< 50$ iterations on average; LB models (D1, D3) require up to 150 iterations. All configurations reach stable RMSE values with no oscillation or divergence.

**Proportional hazards:** Schoenfeld residual tests on the D4 training-cohort fit find no statistically significant evidence of PH violation at the 0.05 level (global test), supporting standard Cox coefficient interpretation for the two active programs.

## 5. Recommended Configuration

**D4 (primary):** YFB, DeSurv-aligned gene selection (2,064 genes), per-platform z-standardization, $K = 7$ (CV-selected), no cohort indicator. Mean external C = `r mean_c("D4")`. Scoring new patients requires only $\hat{\eta}_{\text{new}} = Y_{\text{new}} F \hat{\beta}$ — no latent-variable approximation. Factor columns of $F$ are directly interpretable gene weight vectors and serve as the input to downstream pathway enrichment analysis.

**D3 (sensitivity):** LB, same DeSurv gene set, $K = 7$. Mean external C = `r mean_c("D3")`. Reproduces the same two-program, adverse+protective structure as D4, confirming the finding is not specific to the YFB parameterization. LB $\hat{\beta}$ magnitudes are on a more natural per-factor scale ($\hat{\beta}_2 = -0.44$, $\hat{\beta}_3 = +0.21$) and may aid interpretability, at the cost of an OLS approximation for new-patient scoring.

## 6. Results Figures

```{r fig-cohort, fig.height=3.2, fig.cap="Per-cohort external concordance index for the three YFB configurations: D2 (original M5 baseline), D4 (DeSurv-aligned, recommended), and D5 (DeSurv-aligned + cohort indicator). Dashed line = chance (C = 0.5). DeSurv selection (D4) improves over the baseline in 4 of 5 cohorts; the cohort indicator (D5) consistently reduces performance, with the largest regressions in PACA-AU-seq and PACA-AU-array."}
if (!is.null(res)) {
  plot_df <- res[res$model %in% c("D2", "D4", "D5"), ] |>
    mutate(
      label = factor(model,
        levels = c("D2", "D4", "D5"),
        labels = c("D2: YFB orig (M5, C=0.624)",
                   "D4: YFB DeSurv (recommended, C=0.636)",
                   "D5: YFB DeSurv + cohort (C=0.614)")),
      cohort = factor(cohort,
        levels = c("Dijk","Moffitt_GEO_array","PACA_AU_array","PACA_AU_seq","Puleo_array"),
        labels = c("Dijk","Moffitt\nGEO","PACA-AU\narray","PACA-AU\nseq","Puleo"))
    )
  ggplot(plot_df, aes(x = cohort, y = c_index, fill = label)) +
    geom_col(position = position_dodge(0.75), width = 0.65, color = "white") +
    geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey40") +
    scale_fill_manual(
      values = c("D2: YFB orig (M5, C=0.624)"               = "#f58518",
                 "D4: YFB DeSurv (recommended, C=0.636)"    = "#e45756",
                 "D5: YFB DeSurv + cohort (C=0.614)"        = "#b279a2"),
      name = NULL) +
    coord_cartesian(ylim = c(0.45, 0.72)) +
    labs(x = NULL, y = "C-index") +
    theme_bw(base_size = 8) +
    theme(legend.position = "top",
          legend.text     = element_text(size = 7),
          panel.grid.minor = element_blank(),
          axis.text.x     = element_text(size = 8))
} else {
  plot.new(); text(0.5, 0.5, "Results CSV not found")
}
```


## 7. Open Items

- **Pathway enrichment on D4 Factors 3 and 7.** Submit top-weighted genes to MSigDB / GSEA hallmark gene sets; cross-reference with Moffitt basal/classical and Bailey et al. (2016) four-subtype annotations in TCGA to assess whether the two programs correspond to established PDAC molecular axes.
- **Compare patient loadings $\hat{L}_{i,3}$ and $\hat{L}_{i,7}$ to Moffitt subtype scores** available in TCGA\_PAAD to quantify concordance with classical vs. basal subtypes.
- **YFB point-normal prior K-CV** still returns C = 0.5 for all $K$ — open issue; normal prior (current default) is preferred.
