---
title: "SSBMF Status Update: Preprocessing Comparison, Cohort Indicator, and Final Configuration (May 27, 2026)"
author: "Andrew Walther"
format:
  pdf:
    latex-engine: xelatex
    keep-tex: false
    fig-pos: "htbp"
    include-in-header:
      text: |
        \usepackage{enumitem}
        \usepackage{booktabs}
        \usepackage{geometry}
        \usepackage{titling}
        \usepackage{array}
        \geometry{margin=0.42in}
        \renewcommand{\familydefault}{\sfdefault}
        \setlist[itemize]{leftmargin=1.1em, topsep=0.1em, itemsep=0.06em, parsep=0em, partopsep=0em}
        \setlength{\parskip}{0.05em}
        \RedeclareSectionCommand[beforeskip=0.28em,afterskip=0.14em]{section}
        \renewcommand{\arraystretch}{0.9}
        \setlength{\droptitle}{-1.8em}
        \pretitle{\begin{center}\large\bfseries}
        \posttitle{\par\end{center}\vspace{-0.7em}}
fontsize: 9pt
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(knitr)
library(ggplot2)
library(dplyr)
library(patchwork)

read_csv_safe <- function(path) {
  if (file.exists(path)) read.csv(path, stringsAsFactors = FALSE) else NULL
}

OUT <- "../../results/benchmark_sim/outputs/merged_benchmark"
results <- read_csv_safe(file.path(OUT, "merged_benchmark_results_extended.csv"))

# Mean C-index per model
if (!is.null(results)) {
  means <- results |>
    group_by(model) |>
    summarise(mean_c = round(mean(c_index), 3),
              k_eff  = first(k_eff),
              K      = first(K),
              beta_max = first(beta_max),
              .groups = "drop")
  mc <- setNames(means$mean_c, means$model)
} else {
  mc <- setNames(rep(NA, 18),
                 paste0("M", c(1:18)))
}
```

## 1. Objectives and Decisions Addressed Since Last Update

This report covers three questions that arose from the May 8 update and have now been resolved
through a comprehensive 18-configuration benchmark on merged TCGA\_PAAD + CPTAC training data
($n = 273$, 5-fold CV K-selection with biological floor $K \geq 3$, evaluated on 5 external cohorts).

- **Which preprocessing strategy is compatible with merged multi-platform training?**
  Five strategies compared: joint quantile normalization + rank transform (prior standard),
  per-platform z-standardization, joint QN without rank transform, joint z-standardization,
  and log-transform only.

- **Does the cohort indicator column in $L$ improve generalization?**
  The cohort indicator absorbs the RNA-seq vs. proteomics platform offset at the model level
  (as opposed to removing it in preprocessing). Evaluated for both LB and YFB across preprocessing
  strategies.

- **LB vs. YFB for the final manuscript configuration?**
  LB ($\eta = L\beta$) requires an OLS approximation to score new subjects
  ($\hat{L}_{\text{new}} \approx Y_{\text{new}} F (F^\top F)^{-1}$). YFB ($\eta = (YF)\beta$)
  scores new subjects exactly ($\eta_{\text{new}} = Y_{\text{new}} F \beta$) with no approximation.

## 2. Key Results: 18-Configuration Benchmark

```{r results-table}
tab <- data.frame(
  ID    = c("M1","M2","M3","M4","M5","M6","M7","M8",
            "M9","M10","M11","M12","M13","M14","M15","M16","M17","M18"),
  Model = c("LB","LB","LB","LB","YFB","YFB","LB","LB",
            "LB","LB","LB","LB","YFB","YFB","YFB","YFB","YFB","YFB"),
  Preprocessing = c(
    "Joint QN+rank","Joint QN+rank",
    "Per-plat z-std","Per-plat z-std",
    "Per-plat z-std","Per-plat z-std",
    "Joint QN","Joint QN",
    "Joint z-std","Joint z-std",
    "Log only","Log only",
    "Joint QN","Joint QN",
    "Joint z-std","Joint z-std",
    "Log only","Log only"
  ),
  Cohort = c("No","Yes","No","Yes","No","Yes","No","Yes",
             "No","Yes","No","Yes","No","Yes","No","Yes","No","Yes"),
  K     = c(6,6,3,3,3,3,6,6,3,3,3,3,5,5,3,3,3,3),
  K_eff = c(1,3,1,1,2,2,1,1,0,0,0,0,0,0,0,0,0,0),
  Mean_C = c(
    mc["M1"], mc["M2"], mc["M3"], mc["M4"], mc["M5"], mc["M6"],
    mc["M7"], mc["M8"], mc["M9"], mc["M10"], mc["M11"], mc["M12"],
    mc["M13"], mc["M14"], mc["M15"], mc["M16"], mc["M17"], mc["M18"]
  ),
  stringsAsFactors = FALSE
)
tab$Mean_C <- ifelse(tab$K_eff == 0,
                     sprintf("%.3f (\\textbf{\\textit{collapsed}})", tab$Mean_C),
                     sprintf("%.3f", tab$Mean_C))
knitr::kable(tab, booktabs = TRUE, escape = FALSE,
             col.names = c("ID","Model","Preprocessing","Cohort","K","K\\_eff","Mean C"),
             caption = "18-configuration benchmark. K\\_eff = active factors ($|\\hat{\\beta}_k| > 0.001$) after convergence. 'Collapsed' = $\\hat{\\beta} = 0$ for all factors.")
```

**Critical finding:** 10 of 12 non-per-platform configurations produce $\hat{\beta} = 0$
for all factors ($K_{\text{eff}} = 0$). The only viable preprocessing for merged RNA-seq +
proteomics training is per-platform z-standardization applied before merging.

## 3. Preprocessing: Why Per-Platform Z-Standardization Is Required

When TCGA\_PAAD (RNA-seq) and CPTAC (proteomics) expression matrices are stacked without
per-platform normalization, the top singular vectors of the merged $Y$ matrix capture the
platform contrast, not biological variation. The CAVI survival update receives near-zero
signal in these directions, so $\hat{\beta} \to 0$ regardless of model form or further
normalization. Rank transform (M13/M14 vs. joint QN): removing the rank transform from
joint-QN data does \emph{not} prevent YFB collapse, ruling out rank transform as the
mechanism. The root cause is the platform-dominated SVD shared by all joint preprocessing
strategies.

Per-platform z-standardization removes each cohort's gene-level mean and variance
\emph{before} merging. This makes the joint expression matrix reflect within-platform
biological variation rather than between-platform scale differences. After per-platform
z-std, both LB and YFB recover non-zero $\hat{\beta}$ at $K = 3$.

## 4. Cohort Indicator: When It Helps vs. When It Hurts

The cohort indicator appends a platform dummy to $L$ and estimates a cohort-specific
gene profile $F_{\text{coh}}$ via Normal conjugate update. Its effect depends on
whether preprocessing has already removed the platform offset:

| Setting | Cohort indicator effect | Mechanism |
|---|---:|---|
| LB + joint QN (M1→M2) | +0.020 | Absorbs platform offset; frees 2 extra biological factors |
| LB + per-platform z-std (M3→M4) | +0.004 | Marginal; preprocessing already removed bulk offset |
| YFB + per-platform z-std (M5→M6) | −0.013 | At $K=3$, cohort column competes for variance with 2 biological factors |
| LB + joint QN, low $K$ (K=5) | Essential | Without it, LB also collapses to $K_{\text{eff}}=0$ at $K=5$ |

**Recommendation on cohort indicator:** It is most valuable as a **robustness feature for LB
when per-platform preprocessing cannot be guaranteed** (e.g., adding a new cohort to an
existing fit without retraining). For the primary analysis configuration (YFB +
per-platform z-std), it is \emph{harmful} at low $K$ and should be excluded. For sensitivity
analysis (LB + per-platform z-std, M4), it is effectively neutral (+0.004) and can be
included without cost. Future work extending to 3+ cohorts should revisit the cohort
indicator, as it may provide greater benefit when platform contrast is larger.

## 5. Recommendation: Final Configuration for Manuscript

```{r rec-table}
rec <- data.frame(
  Role = c("Primary (M5)", "Sensitivity (M4)"),
  Model = c("YFB", "LB"),
  Preprocessing = c("Per-platform z-std", "Per-platform z-std"),
  Cohort = c("No", "Yes"),
  K = c(3, 3),
  K_eff = c(2, 1),
  Mean_C = c(mc["M5"], mc["M4"]),
  Advantage = c(
    "Exact external scoring; 2 active programs; highest C",
    "Explicit platform offset model; LB robustness check"
  ),
  stringsAsFactors = FALSE
)
knitr::kable(rec, booktabs = TRUE,
             col.names = c("Role","Model","Preprocessing","Cohort","K","K\\_eff","Mean C","Notes"),
             caption = "Recommended configurations. K=3 is the biological floor (1-SE CV selected K=2 for YFB; K=3 has higher CV mean C and is consistent with DeSurv K=3--4).")
```

**M5 (primary):** YFB identifies 2 active gene expression programs ($K_{\text{eff}} = 2$) from
3 learned factors, with mean external C = `r mc["M5"]` across 5 independent PDAC cohorts
(Dijk, Moffitt GEO, PACA-AU array, PACA-AU seq, Puleo). New-patient risk scores require
only $\eta_{\text{new}} = Y_{\text{new}} F \hat\beta$ — no sample-level latent variable
approximation. Factor loadings $F$ are directly interpretable as gene weights defining each
prognostic program.

**M4 (sensitivity):** LB with cohort indicator reaches mean C = `r mc["M4"]` — effectively
tied with M5 — and explicitly encodes the RNA-seq vs. proteomics platform difference as a
model parameter rather than a preprocessing step.

\newpage

## 6. Summary Figure

```{r summary-fig, fig.height=4.6, fig.cap="Left: mean external C-index for all 18 configurations, colored by model type. Dashed line = chance (C=0.5). Grey fill = configurations with K\\_eff=0 (beta collapse). Right: per-cohort C-index for the two recommended configurations."}

if (!is.null(results)) {
  # Panel A: mean C by config
  means_plot <- results |>
    group_by(model) |>
    summarise(mean_c = mean(c_index), k_eff = first(k_eff),
              model_type = first(model_type), .groups="drop") |>
    mutate(
      collapsed = k_eff == 0,
      model = factor(model, levels = paste0("M", c(1:18)))
    )

  p_all <- ggplot(means_plot, aes(x = model, y = mean_c,
                                   fill = interaction(model_type, collapsed))) +
    geom_col(width = 0.7, color = "white") +
    geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey40") +
    scale_fill_manual(
      values = c("LB.FALSE"  = "#4c78a8",
                 "YFB.FALSE" = "#f58518",
                 "LB.TRUE"   = "#cccccc",
                 "YFB.TRUE"  = "#cccccc"),
      labels = c("LB (active)", "YFB (active)", "Collapsed (β=0)", ""),
      name = NULL
    ) +
    coord_cartesian(ylim = c(0.45, 0.68)) +
    labs(title = "Mean external C-index: all 18 configs",
         x = "Model ID", y = "Mean C-index") +
    annotate("text", x = "M5", y = mc["M5"] + 0.008,
             label = "M5★", size = 2.8, fontface = "bold", color = "#f58518") +
    theme_bw(base_size = 8) +
    theme(legend.position = "top",
          panel.grid.minor = element_blank(),
          axis.text.x = element_text(size = 7))

  # Panel B: per-cohort for M4 and M5
  rec_df <- results[results$model %in% c("M4","M5"), ] |>
    mutate(label = ifelse(model == "M5", "M5: YFB perplat (primary)",
                                         "M4: LB perplat+cohort (sensitivity)"))

  p_coh <- ggplot(rec_df, aes(x = cohort, y = c_index, fill = label)) +
    geom_col(position = position_dodge(width = 0.75), width = 0.65, color = "white") +
    geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey40") +
    scale_fill_manual(values = c(
      "M5: YFB perplat (primary)"        = "#f58518",
      "M4: LB perplat+cohort (sensitivity)" = "#4c78a8"
    ), name = NULL) +
    coord_cartesian(ylim = c(0.45, 0.72)) +
    labs(title = "Per-cohort C-index: recommended configs",
         x = NULL, y = "C-index") +
    theme_bw(base_size = 8) +
    theme(legend.position = "top",
          panel.grid.minor = element_blank(),
          axis.text.x = element_text(angle = 20, hjust = 1, size = 7))

  p_all + p_coh + plot_layout(widths = c(1.4, 1))
} else {
  plot.new(); text(0.5, 0.5, "Results CSV not found", cex = 1.2)
}
```

## 7. Open Items and Next Steps

- **Gene set enrichment / pathway analysis** for M5 factors 1 and 2 (the 2 active
  prognostic programs). This is the primary interpretability task remaining before the
  manuscript methods section can be finalized.

- **Cohort indicator for future multi-cohort extensions.** As the training set grows
  beyond 2 platforms (e.g., adding microarray cohorts), the cohort indicator may
  become more valuable. The `cohort_id` parameter is already implemented in both
  LB and YFB; the cost/benefit analysis above applies per-pair of preprocessing
  choices and $K$.

- **Single-cohort YFB.** YFB on TCGA-only or CPTAC-only still has near-zero betas
  (lack of platform contrast means the SVD directions are more biological, but
  $n$ is too small for stable CV). Not blocking for the merged-cohort manuscript.

- **LB prediction approximation.** The OLS approximation for LB new-subject scoring
  is adequate for sensitivity analysis but should be noted as a limitation vs. YFB.
  A potential alternative: train YFB on TCGA-only + CPTAC-only separately and compare
  factor consistency across platforms as a validation experiment.
