---
title: "SSBMF Status Update: Model Refinement, Validation, and Biological Interpretation (July 15, 2026)"
author: "Andrew Walther"
format:
  pdf:
    latex-engine: xelatex
    keep-tex: false
    fig-pos: "htbp"
    include-in-header:
      text: |
        \usepackage{enumitem}
        \usepackage{booktabs}
        \usepackage{geometry}
        \usepackage{titling}
        \usepackage{array}
        \geometry{margin=0.42in}
        \renewcommand{\familydefault}{\sfdefault}
        \setlist[itemize]{leftmargin=1.1em, topsep=0.1em, itemsep=0.06em, parsep=0em, partopsep=0em}
        \setlength{\parskip}{0.05em}
        \RedeclareSectionCommand[beforeskip=0.28em,afterskip=0.14em]{section}
        \renewcommand{\arraystretch}{0.9}
        \setlength{\droptitle}{-1.8em}
        \pretitle{\begin{center}\large\bfseries}
        \posttitle{\par\end{center}\vspace{-0.7em}}
  html:
    toc: false
    theme: cosmo
fontsize: 9pt
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(knitr); library(dplyr)
DS   <- "../../results/benchmark_sim/outputs/desurv_comparison/desurv_comparison_results.csv"
MC   <- "../../results/multi_cohort_sim/outputs/multicohort_sim_results.csv"
CI   <- "../../results/benchmark_sim/outputs/desurv_comparison/external_cindex_ci.csv"
DIFF <- "../../results/benchmark_sim/outputs/desurv_comparison/external_paired_diff_ci.csv"
rd  <- function(p) if (file.exists(p)) read.csv(p, stringsAsFactors = FALSE) else NULL
desurv <- rd(DS); mc <- rd(MC); ext_ci <- rd(CI); ext_diff <- rd(DIFF)
rec_mean <- if (!is.null(desurv)) round(mean(desurv$c_index[desurv$model == "D4"]), 3) else 0.627
```

## Headline

The recommended model has **not changed** since the 6/18 meeting — it remains **YFB**
($\eta = (YF)\hat\beta$), DeSurv-aligned gene selection, $K=7$ latent programs, no cohort indicator.
What changed is that its numbers were **corrected** and its factor count was **stress-tested**. The
scientific conclusion is unchanged and now better supported: the model uses **2 of its 7 programs**
for survival prediction, and this is robust, not a tuning artifact. **Mean external C-index = `r rec_mean`
across 5 held-out PDAC cohorts.**

## What was done since 6/18

- **Objective put on a principled footing (§2 of the full report).** The genomics/survival scale
  imbalance the lab flagged was addressed; the redundant $\lambda$ mixing parameter was retired
  ($\alpha$ already plays its role). Honest finding: for YFB this gave **no performance change** —
  its genomics half and survival half never share a coordinate needing rebalancing. A separate
  train/test preprocessing bug (external cohorts were rank-transformed while training was
  per-platform z-standardized — the reverse) was fixed; **this fix alone** explains the headline
  moving from 0.636 to **`r rec_mean`** (a correction, not a regression).
- **Why 7 factors when DeSurv used 3? (§4)** Cross-validation genuinely selects $K=7$, but only
  **2 factors are survival-active** ($K_{\text{eff}}=2$) — matching DeSurv's own ~1 prognostic
  factor. Two independent optimization strategies (warm-start from a higher-$K$ fit, and a joint
  $(K,\alpha)$ Bayesian-optimization search) both land on $K_{\text{eff}}=2$, and a
  deflation-initialization test analytically ruled out initialization as the cause of an earlier
  ambiguity; $K\in\{4,5,7,8\}$ are statistically indistinguishable (spread 0.0015). $K=7$ is kept
  only for one-step reproducibility; $K=4/K=5$ are validated equivalents.
- **Study-specific baseline hazard (§5).** Implemented the requested `+ strata(study)` as an
  optional stratified Cox partial likelihood. Performance-neutral (0.6267 → 0.6263), so available
  but off by default.
- **Biological interpretation of the two active programs (§6).** Complete, and confirmed by five
  independent methods.

## Validation across simulated and real data

**Simulation — joint model vs. a two-step (factorize-then-Cox) baseline** (re-confirmed under the
corrected model this cycle):

- *Increasing survival signal from zero:* at zero signal, joint $\approx$ two-step (both near chance);
  as signal grows, the joint model exceeds the two-step baseline by a growing margin, in 3 of 4 data-generating
  designs. One artificial design (exactly-sparse, symmetric factor loadings) reverses this — traced
  to a specific, understood factor-collapse failure mode, not a survival-coupling defect.
- *Shared vs. study-specific factors:*

```{r mc-table}
if (!is.null(mc)) {
  lab <- c(all_shared="All factors shared", hybrid="Hybrid (some shared, some specific)",
           nothing_shared="No shared prognostic factors")
  rows <- lapply(names(lab), function(s) data.frame(
    Scenario = lab[[s]],
    `Joint (YFB)` = round(mean(mc$c_index[mc$scenario==s & mc$arm=="YFB_base"]),3),
    `Two-step (EBMF+Cox)` = round(mean(mc$c_index[mc$scenario==s & mc$arm=="EBMF"]),3),
    check.names = FALSE))
  knitr::kable(do.call(rbind, rows), row.names = FALSE, booktabs = TRUE,
               caption = "Held-out C-index by cross-cohort factor structure (5 seeds).")
}
```

  When no shared prognostic signal exists, joint and two-step are equivalent; in the realistic
  hybrid case (only some factors prognostic) the joint model leads — as much through **stability**
  as accuracy: the unsupervised two-step baseline collapses on 2 of 5 seeds (which lowers its mean to
  0.717), while the joint model is stable across all seeds (mean 0.815) and classifies each factor
  as shared vs. study-specific with perfect accuracy.

**Real data — 5 held-out PDAC cohorts** (RNA-seq, microarray, proteomics):

```{r ext-table}
if (!is.null(ext_ci)) {
  d4 <- ext_ci %>% filter(model == "YFB (recommended)") %>% arrange(desc(c_index)) %>%
    mutate(`95% CI` = sprintf("(%.3f, %.3f)", ci_lo, ci_hi)) %>%
    select(cohort, c_index, `95% CI`)
  names(d4)[1:2] <- c("Held-out cohort", "External C-index")
  knitr::kable(d4, row.names = FALSE, booktabs = TRUE, digits = 3,
               caption = sprintf("External validation with 95%% bootstrap CI (mean = %.3f).", rec_mean))
}
```

Four of five cohorts clear chance with their CI lower bound above 0.5; **Moffitt microarray does
not** (CI 0.467–0.625), confirming quantitatively what was previously only a qualitative flag.
**Against a two-step (unsupervised EBMF then Cox) baseline on the same patients, a paired bootstrap
finds the recommended model significantly more concordant when pooled across all 5 cohorts: +0.042
(95% CI 0.013–0.071).** Individually only the largest cohort (Puleo, $n=288$) reaches significance
alone; the other four all favor the recommended model but are underpowered on their own. Per-platform
z-standardization is essential — most non-per-platform preprocessing collapses the survival
coefficients to zero.

## Biology of the two active programs

An **adverse** program (basal-like / squamous / MET-EGFR-driven) and a **protective** program
(classical / differentiated-epithelial) — the established PDAC subtype-survival axis. Confirmed by
five independent methods: gene-set enrichment, PurIST subtype concordance
($\rho=\pm0.58$), external-cohort hazard ratios (adverse HR>1 in 5/5 cohorts, protective HR<1 in
5/5), and direct gene-list overlap with DeSurv's published programs.

## Positioning and open items (for discussion)

- **Positioning:** SBMF is best framed as the **Bayesian counterpart to DeSurv** — same design,
  data, and scoring, differentiated by empirical-Bayes shrinkage, posterior uncertainty,
  error-in-variables loadings, and a per-gene noise model. A **matched, same-protocol head-to-head
  with DeSurv is a pending next step**, not a current superiority claim.
- **Open items worth raising:** (1) factor-stability across seeds not yet characterized; (2) the LB
  parameterization retains a factor-collapse sensitivity (the recommended YFB model does not); (3) the
  YFB spike-and-slab prior collapses in cross-validation (normal prior used throughout, and is the
  right call); (4) requested workflow items (automated commit review, RAG literature database, HPC
  workflow) and a review of DeSurv's own simulation construction remain on the backlog.

*Full detail, figures, and per-item disposition of the 6/18 feedback: `docs/reports/ssbmf_progress_report_07_15_26`.*
