---
title: "PDAC Data Audit"
subtitle: "Dataset Inventory and Integration Assessment for Supervised Bayesian Matrix Factorization"
author: "Andrew Walther"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    theme: cosmo
    smooth-scroll: true
    self-contained: true
  pdf:
    toc: true
    toc-depth: 3
    number-sections: true
    geometry: "margin=1in"
    mainfont: "Helvetica Neue"
    monofont: "Menlo"
    colorlinks: true
    fig-cap-location: bottom
    tbl-cap-location: top
    keep-tex: false
    fig-pos: "H"
    include-in-header:
      text: |
        \usepackage{booktabs}
        \usepackage{longtable}
        \usepackage{float}
        \usepackage{amsmath}
        \usepackage{amssymb}
        \floatplacement{figure}{H}
        \floatplacement{table}{H}
execute:
  echo: false
  message: false
  warning: false
  cache: false
---

# Executive Summary

This document provides a structured inventory of the PDAC (Pancreatic Ductal Adenocarcinoma)
datasets available for use with the supervised Bayesian matrix factorization model implemented
in `multiomicsGEP`. Seven datasets are available, spanning RNA-seq, microarray, and proteomics
platforms. Each dataset includes a gene expression matrix, survival data (time-to-event with
censoring), and CAF (cancer-associated fibroblast) subtype annotations.

All seven datasets are compatible with `fit_supervised_mf_modular()` after standardised
preprocessing via the `load_data_internal()` function. With top-5000 variance-filtered genes,
the estimated total compute time for all datasets is approximately one minute.

# Data Location and File Structure

## Source Directory

```
~/OneDrive/UNC Dissertation (Liu)/PDAC_data/
├── load_data_internal.R          ← Standardised loader function
└── original/                     ← Raw data files (7 datasets)
```

## Per-Dataset File Convention

Each dataset `{name}` has four associated files:

| File | Contents | Format |
|------|----------|--------|
| `{name}.rds` | Expression matrix + sample metadata + feature annotations | R list |
| `{name}.survival_data.rds` | Survival times, event indicators, whitelist flags | R data.frame |
| `{name}_subtype.csv` | CAF subtype classifications (SCISSORS, PurIST, etc.) | CSV |
| `{name}.caf_subtype.rds` | Same subtype data in RDS format | R object |

Additionally, `cmbSubtypes.RData` is present as an ancillary reference file for combined
subtype analysis. It is not required for the matrix factorization pipeline.

# Dataset Inventory

```{r}
#| label: tbl-inventory
#| tbl-cap: "Summary of available PDAC datasets"

inventory <- data.frame(
  Dataset = c("TCGA_PAAD", "CPTAC", "Dijk", "Moffitt_GEO_array",
              "PACA_AU_array", "PACA_AU_seq", "Puleo_array"),
  Platform = c("RNA-seq", "Proteomics", "RNA-seq", "Microarray",
               "Microarray", "RNA-seq", "Microarray"),
  Genes = c(20531, 28057, 49677, 19749, 47265, 56206, 20087),
  Samples_Total = c(181, 140, 90, 357, 131, 92, 309),
  Samples_Valid = c(144, 134, 90, 123, 88, 69, 288),
  Events = c(75, 68, 81, 83, 54, 41, 181),
  Censored = c(69, 66, 9, 40, 34, 28, 107),
  Censoring_Pct = c(47.9, 49.3, 10.0, 32.5, 38.6, 40.6, 37.2),
  Time_Range = c("0.1 -- 76.2", "0.0 -- 45.5", "2.0 -- 119.8",
                 "1.0 -- 59.0", "0.3 -- 47.0", "0.3 -- 50.4", "1.1 -- 146.4"),
  Log2_Applied = c("Yes", "No", "Yes", "No", "No", "Yes", "No"),
  stringsAsFactors = FALSE
)

knitr::kable(inventory, align = c("l", "l", "r", "r", "r", "r", "r", "r", "c", "c"))
```

**Notes:**

- *Samples_Valid* reflects samples retained after survival validation (time > 0, non-missing
  time and event) and dataset-specific filtering applied by `load_data_internal()`.
- *Log2_Applied* indicates whether `load_data_internal()` applies a `log2(x + 1)` transformation
  to the expression matrix. Datasets already on a log or normalised scale (microarray,
  proteomics) are not re-transformed.
- *Time_Range* is in months for all datasets.

# Per-Dataset Profiles

## TCGA_PAAD {#sec-tcga}

**Source:** The Cancer Genome Atlas — Pancreatic Adenocarcinoma project.

**Platform:** RNA-seq (raw counts → log2-transformed by loader).

**Dimensions:** 20,531 genes × 181 samples (144 after filtering).

**Expression range (raw):** 0 -- 6,699,497 (counts); after log2(x + 1): 0 -- 22.7.

**Feature annotation:** `featInfo` is a data.frame with columns `SYMBOL` and `ENTREZID`.
Gene symbols are set as row names by the loader.

**Sample filtering:** Retains samples with non-missing tumour grade (`!is.na(Grade)`).

**Survival:** 75 events / 69 censored (47.9% censoring). Follow-up range: 0.1 -- 76.2 months.

**Sample metadata:** Rich clinical annotation (75 columns) including tumour grade, purity
estimates, KRAS mutation status, molecular subtype classifications (Moffitt, Collisson, Bailey),
staging, and treatment history.

**Additional objects in `.rds`:** `decoderWeights`, `subTME` (tumour microenvironment scores).

**Suitability:** Ideal starting dataset — widely studied, moderate size, comprehensive
annotation. Recommended as the smoke-test target.

## CPTAC

**Source:** Clinical Proteomic Tumor Analysis Consortium — PDAC proteomics.

**Platform:** Mass spectrometry proteomics (already normalised; not log-transformed).

**Dimensions:** 28,057 proteins × 140 samples (134 after filtering).

**Expression range:** 0 -- 24.8 (normalised abundance, likely log2-transformed upstream).

**Feature annotation:** `featInfo` is a list of length 1 (minimal annotation).

**Sample filtering:** Retains samples with `histology_diagnosis == "PDAC"`.

**Survival:** 68 events / 66 censored (49.3%). Follow-up: 0 -- 45.5 months.

**Note:** This is a *proteomics* dataset, not transcriptomics. The supervised MF model
was derived for gene expression, but the mathematical framework (Y = LF' + E with Cox PH)
applies to any continuous omics matrix. Results should be interpreted in the proteomics context
(protein abundance programs rather than gene expression programs). Features are proteins, not
genes — top-feature tables will report protein identifiers rather than gene symbols.

## Dijk

**Source:** Dijk et al. — PDAC RNA-seq cohort.

**Platform:** RNA-seq (raw counts → log2-transformed by loader).

**Dimensions:** 49,677 genes × 90 samples (90 after filtering — all retained).

**Expression range (raw):** 0 -- 89,627 (counts); after log2(x + 1): 0 -- 16.5.

**Feature annotation:** `featInfo` is a data.frame with `SYMBOL`, `GeneID`, `Biotype` —
the richest feature annotation among the datasets.

**Sample filtering:** No filtering applied (all 90 samples have valid survival).

**Survival:** 81 events / 9 censored (10.0% censoring — very low). Follow-up: 2.0 -- 119.8
months.

**Note:** The very low censoring rate (10%) means nearly all patients experienced the event.
This is unusual and may reflect the study design or patient population. High event rates
provide strong signal for Cox model estimation but reduce generalisability.

**Sample metadata:** Includes age, sex, lymph node metastasis ratio, organism part.
The `.rds` also contains a `cnt` object (raw count matrix) in addition to `ex`.

## Moffitt_GEO_array {#sec-moffitt}

**Source:** Moffitt et al. — GEO microarray cohort.

**Platform:** Microarray (already normalised; not log-transformed by loader).

**Dimensions:** 19,749 genes × 357 samples (123 after filtering).

**Expression range:** 0 -- 13.1 (log2-scale microarray intensities).

**Feature annotation:** `featInfo` is a data.frame with column `SYMBOL`.

**Sample filtering:** Retains primary tumour specimens only (`specimen_type == "Primary"`).
The large reduction from 357 to 123 samples reflects exclusion of non-primary specimens
(metastases, adjacent normal, stroma).

**Survival:** 83 events / 40 censored (32.5%). Follow-up: 1.0 -- 59.0 months.

**Sample metadata:** Includes Moffitt tumour/stroma classifications, GEO accession IDs.
The `.rds` also contains `decoderWeights`.

## PACA_AU_array {#sec-paca-array}

**Source:** Pancreatic Cancer — Australia (ICGC) — microarray platform.

**Platform:** Microarray (already normalised; not log-transformed by loader).

**Dimensions:** 47,265 probes × 131 samples (88 after filtering).

**Expression range:** 1.0 -- 14.1 (log2-scale microarray intensities).

**Feature annotation:** `featInfo` is a data.frame with `PROBEID`, `ENTREZID`, `ENSEMBL`,
`SYMBOL`. Multiple probes may map to the same gene — the loader resolves this by taking
the median across duplicated gene symbols.

**Sample filtering:** Retains primary PDAC tumours only (`Sample.type == "Primary tumour"`
AND `HistoSubtype == "Pancreatic Ductal Adenocarcinoma"`). Also aggregates duplicate genes.

**Survival:** 54 events / 34 censored (38.6%). Follow-up: 0.3 -- 47.0 months.

**Sample metadata:** ICGC identifiers, qpure tumour purity scores, histological subtypes.

## PACA_AU_seq {#sec-paca-seq}

**Source:** Pancreatic Cancer — Australia (ICGC) — RNA-seq platform.

**Platform:** RNA-seq (raw counts → log2-transformed by loader).

**Dimensions:** 56,206 genes × 92 samples (69 after filtering).

**Expression range (raw):** 0 -- 1,224,510 (counts); after log2(x + 1): 0 -- 20.2.

**Feature annotation:** `featInfo` is a data.frame with `ENSEMBL`, `ENTREZID`, `SYMBOL`.
Duplicate genes aggregated by median.

**Sample filtering:** Same criteria as PACA_AU_array (primary PDAC tumours). Column names
are suffixed with `_seq` to distinguish from the array cohort.

**Survival:** 41 events / 28 censored (40.6%). Follow-up: 0.3 -- 50.4 months.

**Note:** This is the same patient cohort as PACA_AU_array but profiled on a different
platform. Paired analysis could compare factor recovery across platforms.

## Puleo_array {#sec-puleo}

**Source:** Puleo et al. — large European PDAC microarray cohort.

**Platform:** Microarray (already normalised; not log-transformed by loader).

**Dimensions:** 20,087 genes × 309 samples (288 after filtering).

**Expression range:** 0.3 -- 14.3 (log2-scale microarray intensities).

**Feature annotation:** `featInfo` is a data.frame with column `SYMBOL`.

**Sample filtering:** All samples retained (no additional filtering beyond survival validation).

**Survival:** 181 events / 107 censored (37.2%). Follow-up: 1.1 -- 146.4 months.

**Note:** Largest dataset by valid sample count (n=288). Longest follow-up (>12 years).
Provides the most statistical power for survival modelling. The rich sample metadata
includes TNM grading, resection margin status, sex, and sampling site.

# The `load_data_internal()` Function

## Location

```
~/OneDrive/UNC Dissertation (Liu)/PDAC_data/load_data_internal.R
```

## Interface

```r
load_data_internal(dataname)
# Returns: list(ex, sampInfo, featInfo, dataname)
```

**Input:** A dataset name string matching one of the supported datasets.

**Output:**

| Field | Type | Description |
|-------|------|-------------|
| `ex` | numeric matrix | Gene expression (genes × samples) |
| `sampInfo` | data.frame | Sample metadata with `ID`, `time`, `event`, `keep` columns |
| `featInfo` | character vector | Gene/feature symbols (row names of `ex`) |
| `dataname` | character | Dataset name tag |

## Processing Pipeline

The function executes the following steps for each dataset:

1. **Load raw data:** Read `.rds` (expression + metadata), `.survival_data.rds` (survival),
   and `_subtype.csv` (CAF annotations) from `data/original/`.

2. **Dataset-specific preprocessing:**

```{r}
#| label: tbl-preprocessing
#| tbl-cap: "Dataset-specific preprocessing applied by load_data_internal()"

preproc <- data.frame(
  Dataset = c("Dijk", "CPTAC", "TCGA_PAAD", "Moffitt_GEO_array",
              "PACA_AU_array", "PACA_AU_seq", "Puleo_array"),
  Sample_Filter = c(
    "None (all retained)",
    "histology == PDAC",
    "!is.na(Grade)",
    "specimen_type == Primary",
    "Primary tumour + PDAC histology",
    "Primary tumour + PDAC histology",
    "None (all retained)"
  ),
  Expression_Transform = c(
    "log2(x + 1)",
    "None",
    "log2(x + 1); rownames set to SYMBOL",
    "None",
    "Median-aggregate duplicate genes; remove NA symbols",
    "Median-aggregate duplicates; log2(x + 1); suffix _seq to IDs",
    "None"
  ),
  stringsAsFactors = FALSE
)

knitr::kable(preproc, align = c("l", "l", "l"))
```

3. **Survival validation:** Retain samples where `time > 0`, `!is.na(time)`, and
   `!is.na(event)`. The intersection of dataset-specific and survival-valid sample indices
   defines the final `keep` flag.

4. **Sample metadata assembly:** Left-join survival and CAF subtype data. Add `keep`
   indicator column. Validate event indicators are 0/1, times are positive.

5. **Data integrity checks:** Verify expression matrix has column names, `sampInfo` has
   non-missing `ID`, no duplicate sample IDs, expression columns exactly match `sampInfo$ID`
   (reorder if needed).

## Datasets Referenced But Not Present

The loader supports four additional datasets — `Grunwald`, `Hayashi`, `Linehan`, and `Olive` —
but their `.rds` files are **not present** in the `original/` directory. These are skipped
in the current analysis.

# Integration with Supervised Bayesian MF

## Input Contract

`fit_supervised_mf_modular(Y, time, status, K, ...)` expects:

| Parameter | Type | Shape | Description |
|-----------|------|-------|-------------|
| `Y` | numeric matrix | n × p | Patients (rows) × genes (columns) |
| `time` | numeric vector | n | Survival/censoring time |
| `status` | integer vector | n | Event indicator (1 = event, 0 = censored) |

## Required Transformation

`load_data_internal()` returns `ex` as **genes × samples** (rows = genes, columns = samples).
The model expects **patients × genes**. Therefore:

```r
data    <- load_data_internal("TCGA_PAAD")
keeps   <- which(data$sampInfo$keep == 1)
Y       <- t(data$ex[, keeps])     # transpose: genes×samples → patients×genes
time    <- data$sampInfo$time[keeps]
status  <- as.integer(data$sampInfo$event[keeps])
```

## Gene Filtering Recommendation

With 20K--56K genes per dataset, variance-based filtering to the top 5,000 most variable
genes is recommended as the default. This is standard practice in genomics dimensionality
reduction: it removes low-variance housekeeping genes that contribute noise without
informative signal, and substantially reduces runtime. The filter should be configurable
(including the option to disable it) to allow sensitivity analysis.

## Estimated Compute Time

```{r}
#| label: tbl-runtime
#| tbl-cap: "Estimated runtime per dataset (K=5, max_iter=100, top 5,000 genes)"

runtime <- data.frame(
  Dataset = c("Dijk", "CPTAC", "TCGA_PAAD", "Moffitt_GEO_array",
              "PACA_AU_array", "PACA_AU_seq", "Puleo_array", "**Total**"),
  n = c(90, 134, 144, 123, 88, 69, 288, NA),
  p_filtered = c(5000, 5000, 5000, 5000, 5000, 5000, 5000, NA),
  Est_Seconds = c(4, 6, 6, 5, 4, 3, 12, 40),
  stringsAsFactors = FALSE
)

knitr::kable(runtime, align = c("l", "r", "r", "r"))
```

Benchmarked against the synthetic simulation (n=250, p=1000, K=5) which completes in ~2.6
seconds on a 2024 MacBook Pro. Runtime scales linearly with n × p × K.

# Recommendations

1. **Primary smoke-test dataset:** TCGA_PAAD — well-studied, moderate dimensions, rich
   clinical annotation, balanced censoring.

2. **Run all seven datasets** sequentially in a single script invocation. Total compute
   is approximately one minute with top-5000 gene filtering.

3. **Cross-platform comparisons:** PACA_AU_array and PACA_AU_seq share the same patient
   cohort on different platforms — factor recovery comparison is a natural future analysis.

4. **CPTAC caveat:** This is a proteomics dataset. Results should be interpreted as protein
   abundance programs rather than gene expression programs.

5. **Dijk caveat:** Very low censoring (10%) — nearly all patients experienced the event.
   Cox model estimation will be strong but generalisability may be limited.

6. **Gene filtering:** Default to top 5,000 by variance, with a configurable `top_n_genes`
   parameter that accepts `NULL` to disable filtering.

7. **Longleaf future flag:** Data may also be accessible at `/proj/rashidlab/data/PDAC` on
   the UNC Longleaf HPC cluster. A `PDAC_DATA_ROOT` environment variable should be added
   in a future update to support both local and cluster paths.
